### 2011-06-11 seqbias-paper.tex
starting over, after the great hard drive crash of 2011

\documentclass[letterpaper]{article}
\usepackage{qcjstd}
\usepackage{graphicx}
\usepackage{lscape}
\usepackage{multicol}
\usepackage{geometry}
\usepackage{amsmath}
\usepackage{multirow}
%\usepackage{bookman}
%\usepackage{cmbright}

\makeatletter
\newenvironment{tablehere}
  {\def\@captype{table}}
  {\vspace{1em}}

\newenvironment{figurehere}
  {\def\@captype{figure}}
  {\vspace{1em}}
\makeatother

\geometry{left=0.8in,right=0.8in, top=1in, bottom=1in}

\title{A Model for Sequence Bias in RNA-Seq}

\begin{document}

%\maketitle

\qcjtitle{\sc{(Draft)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}

\begin{abstract}
Quantification of RNA abundance in RNA-Seq experiments is often conflated with
protocol specific bias as a result of PCR amplification, differing primer
affinities, and random priming with a non-uniform distribution of primer
sequences.  We propose a method to measure and compensate for these influences
using a simple graphical model, trained without the need of existing gene
annotations.
\end{abstract}

\begin{multicols}{2}

\section{Introduction}

In the last few years, RNA-Seq has emerged as a promising alternative to
microarrays in quantifying RNA abundance. But, as microarray technology has
brought with it technical challenges ranging from robust normalization to
addressing cross-hybridization, RNA-Seq presents a new set of problems.

A significant example is the effect of transcript length and quantification.
As noted by Oshlack, et. al., transcript length biases discovery of differentially
expression, since more reads are aligned to longer transcripts, granting
higher statistical significance in tests differential expression
\cite{Oshlack2009}. Recently, some methods have been developed attempting to account for
this effect in gene set enrichment tests \cite{Young2010} \cite{Gao2011}.
Additionally, differing transcript lengths introduce ``edge
effects'', as very short fragments from the end of a transcript will not be
sequenced. NEUMA \cite{Lee2010} and FPKM \cite{Trapnell2010} are two methods
that attempt to take this into account.

The underlying nucleotide sequence also influences quantification in a manner
that is complex and protocol specific. In an ideal experiment, the number of
reads mapping to a particular position in the genome is a function of RNA
abundance and should not be additionally dependent on the sequence. As pointed
out recently, this is not the case \cite{Dohm2008} \cite{Li2010}
\cite{Hansen2010}. As illustration, Figure \ref{fig:freqs} plots this
nonuniformity in nucleotide frequencies on five data sets (see, Table
\ref{tab:datasets}), each using a different protocol.

These biases may adversely effect transcript discovery, as low level noise may
be over reported in some regions, and in others, active transcription may be
under reported. They render untrustworthy comparisons of relative abundance
between genes, and any test of differential expression hangs on the assumption
that these biases are identical between replicates.  Additionally, in many tests
of differential expression---similarly the effects of transcript length---a higher
read count will result in higher statistical confidence. It follows that the
sensitivity of such a test will also be biased by sequence, affecting downstream
systems biology analysis such as gene ontology enrichment tests.

This bias, though observed primarily in the 5' end of a read, is not resolved by
trimming the reads prior to mapping (see, supplementary Figure 1), suggesting
it is not a result of erroneous base calling, and that a more sophisticated
means of correction is needed.

\begin{figure*}
\begin{center}
\includegraphics[width=0.75\textwidth]{fig/freqs.pdf}
\end{center}
\caption{Nucleotide frequencies relative to the start of each read (labeled
position 0), respecting strand. The divergence of the nucleotide distribution
is summarized using the symmetrized Kullback-Leibler divergence (see Section
3).}
\label{fig:freqs}
\end{figure*}

Some work has been done to model and correct for these biases.  Li, et. al.,
\cite{Li2010} propose two models. The firsts is a Poisson linear model, in which
read counts across a transcript follow inhomogeneous Poisson process. The read
count at position $i$ within the transcript is Poisson distributed with
parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of weights determined by the
nucleotide at each position surrounding the read start, in addition to a term
capturing the abundance of the transcript.

The second model is based on multiple additive regression trees, or MART
\cite{Friedman2003}.  In their tests, MART shows a moderate improvement over the
Poisson linear model. Both models are fit to a number of abundant test genes,
requiring existing gene annotations for the reference genome.

Another model, proposed by Hansen, et. al., \cite{Hansen2010} directly estimates
the distribution of initial heptamers within reads, then estimates a presumed
background heptamer distribution by sampling from the end of reads. Read counts
are then adjusted by the ratio of the two. Specifying two distributions over
heptamers (i.e. foreground and background distributions) requires $2(4^7-1) =
32,766$ parameters, so while no gene annotations are needed to train such a
model, a significant number of reads are required.

Version 0.9.0 of Cufflinks \cite{Trapnell2010} also introduces a model to
correct for sequence bias. No description of their method is yet available, nor
is there any implementation of the model that can be run independently of
Cufflinks to facilitate testing, so we have withheld any comparison in our
tests.

Besides the direct influence of nucleotide composition on sequencing efficiency,
the ``uniqueness'' of a sequence within the genome affects how likely it is to
be mapped correctly, thus influencing quantification in a non-trivial way. Some
methods exist attempting to quantify uniqueness \cite{Koehler2010} and to take
it into account in the measure of transcript abundance \cite{Li2010b}
\cite{Lee2010}. Though we make no attempt to address this problem, any
probabilistic characterization of uniqueness or ``mapability'' could easily be
combined with the probabilistic characterization of sequence bias that we
described here.

\section{Method}

\subsection{Principle}

We assume a simple model of an RNA-Seq experiment: the number of reads $x_i$
aligned to genomic position $i$ is an unbiased estimate of RNA abundance.
Furthermore, we assume reads may be treated as i.i.d. samples. That
is, if $N$ reads are generated, and $m_i$ is the event that a generated read
maps to position $i$,
$$ E[ x_i ] = N \Pr[ m_i ] $$

The experiment may be considered unbiased  with regards to sequence if having
observed the nucleotide sequence $s_i$ surrounding position $i$,
$$ E[ x_i | s_i ] = N \Pr[ m_i | s_i ] = N \Pr[ m_i ] = E[ x_i ] $$

From Bayes' rule,
$$ \Pr[ m_i | s_i ] = \frac{ \Pr[ s_i | m_i ] \Pr[ m_i ] }{ \Pr[ s_i ] } $$

This suggests a natural reweighing scheme. First, define the \emph{sequence
bias} $b_i$ at position $i$ as,
$$ b_i = \frac{ \Pr[ s_i ] } { \Pr[ s_i | m_i ] } $$

Now, if we reweight the read count at position $i$ by $b_i$, we
have,
\begin{align*}
E[ b_i x_i  | s_i ] &= b_i E[ x_i | s_i ] \\
&= N b_i \Pr[ m_i | s_i ] \\
&= N \frac{ \Pr[ m_i | s_i ] \Pr[ s_i ] }{ \Pr[ s_i | m_i ] } \\
&= N \Pr[ m_i ] \\
&= E[ x_i ]
\end{align*}
Thus the reweighted read counts are made to be unbiased.

To estimate the bias $b_i$, we must make estimates of the background sequence
probability $\Pr[s_i]$ and the foreground sequence probability $\Pr[ s_i | m_i
]$, the later being the probability of the sequence given a read being sampled
from its position. Any estimation of sequence probabilities presents a challenge
of finding a model that sufficiently complex to capture the common features of
the training data but avoids overfitting.

Towards that end, we propose training a Bayesian network classifier trained on
examples of foreground and background sequences. Though our goal is not
classification, by using the machinery of classification we can avoid a model
that is over-parametrized. Parameters that are not informative in
discriminating between foreground and background are not included in the model.
The Bayesian network can then used to evaluate sequence probability, and thus
bias at any genomic position.

This discussion so far ignores one complication. The RNA abundance that we wish to
estimate is not itself independent of of the nucleotide sequence. Notably,
exonic DNA is tends to be more GC-rich than intergenic DNA. If background
sequences are sampled uniformly from the genome we run the risk of incorrectly
adjusting for biological sequence bias, rather than technical sequence bias. To
avoid this, we propose using paired training data. Each foreground training
sequence is paired with a background sequence taken from nearby position that is
likely to have similar abundance and general nucleotide composition.
Alternatively, we could pair foreground samples with background samples from
within the same transcript, but we prefer to avoid dependence on existing gene
annotations as such a model would be less suitable for de-novo discovery or
annotation applications.

Hansen, et. al., \cite{Hansen2010} implicitly characterizes and corrects for
bias in the same manner proposed here. Their approach differs primarily in how
sequence probabilities are estimated. Li, et. al., \cite{Li2010} offers a
different perspective, taking advantage of gene annotations to train a model
that optimally explains variance in base level counts across transcripts.

\subsection{Estimation}

To estimate sequencing bias, we train a Bayesian network classifier in which
each node represents a position in the sequence, relative to the read start, and
edges encode dependency between positions. Bayesian networks have been
applied to recognize motifs in nucleotide sequences in the past, in particular
in modeling splice cites \cite{Cai2000} \cite{Chen2005} and transcription factor
binding sites \cite{BenGal2005} \cite{Grau2006} \cite{Pudimat2005}.

In our model, we do not rely on constraining the set of networks (e.g. to
trees), and instead approximate the NP-Hard problem of determining the optimal
network structure using a fast hill climbing algorithm. Furthermore, we
train our model \emph{discriminatively}; only parameters that are deemed
informative in discriminating between foreground and background sequences are
included in the model. We thus seek to train a model that reduces bias, without
including uninformative parameters that would only increase variance.

\subsubsection{Sampling}

To begin, we generate $n$ training examples, where $n = 200,000$ is typical, half of
which labeled as foreground, the other background. To do so, we sample the
positions of $n/2$ reads. Duplicate reads are ignored, to avoid overfitting to
any exceedingly highly expressed locus.  For each foreground position $i$
sampled, we then sample a background position by drawing a random offset $\delta
\sim \text{Normal}( 0, \sigma )$, where by default $\sigma = 50$ is chosen, and
taking $i' = \lceil i + \delta \rceil$. By using such a scheme we attempt to
mitigate the effects of biological sequence bias, sampling positions that are
more likely to be biologically similar.

This sampling procedure produces a set of $n$ genomic positions with
accompanying labels $T = \{ (i_1,x_1), (i_2,x_2), \dots, (i_n,x_n) \}$. The
label $x_k$ is binary, indicating classification as background ($x_{i_k} =
0$) or foreground ($x_{i_k} = 1$). We will denote the nucleotide sequence at
position $i_k$ as $s_{i_k}$.

\subsubsection{Training}

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to BNC algorithm \cite{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood.
\begin{align*}
\ell &= \sum_{k=1}^{n} \log \Pr( x_{i_k} | s_{i_k} ) \\
&=
\sum_{k=1}^{n} \log \frac{ \Pr(s_{i_k} | x_{i_k}) \Pr( x_{i_k} ) }{
\sum_{x \in \{0,1\}} \Pr( s_{i_k} | x ) \Pr(x) } \\
%&= \sum_{k=1}^{n} \log \frac{ \Pr( s_{i_k} | x_{i_k} ) }
%{ \sum_{x \in \{0,1\}} \Pr( s_{i_k} | x ) }
\end{align*}

Where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
0.5$) if we sample foreground and background positions equally.

As we will be estimating parameters and evaluating the likelihood on the same
set of samples, simply maximizing the likelihood would severely overfit the
training set. We thus penalize model complexity using the Akaike information
criterion with the second order sample size correction \cite{Hurvich1989}.
Where $m$ is the number of parameters needed to specify the model, we maximize,
$$ \ell' = \ell - m - \frac{(m + 1)(m + 2)}{n - m - 2} $$

At each step of the optimization procedure, every possible edge or position
addition, removal, or edge reversal that produces a valid network (i.e. does not
introduce a cycle) is evaluated, and the alteration that increases the score
$\ell'$ the most is kept.  This process is repeated until a local maxima is
found, in which no single alteration to the network will increase the score.

Given the network structure, the parameters are estimated directly from the
observed nucleotide frequencies. This choice of parameters maximizes the
likelihood of the model, but not necessarily the conditional likelihood. They
are equivalent asymptotically, thus, as per the BNC algorithm, we adopt this
approximation rather than  resorting to numerical methods which might render the
algorithm too slow to be practical.

The run time of the training procedure is further reduced in practice by imposing the
following two restrictions on the structure of the network,
\begin{enumerate}
\item The in-degree (i.e. number of parents) of any node must be less than some
number $p_{\text{max}}$.
\item $|j - i| \le d_{\text{max}}$ for all edges $(i,j)$ and some number $d_{\text{max}}$.
\end{enumerate}
The second of these encodes the assumption that distant nucleotides are
effectively independent.

\begin{figure*}
\begin{center}
\includegraphics[width=0.75\textwidth]{fig/structure2.pdf}
\end{center}
\caption{The network structure learned on the Trapnell data set, where 0 is the
read start. Dotted boxes represent positions not included in the model (i.e.
positions deemed uninformative).}
\label{fig:structure}
\end{figure*}

Figure \ref{fig:structure} shows an example of the structure learned when this
procedure is applied to 200,000 reads from the Trapnell data set
\cite{Trapnell2010}. The structure learned is non-trivial, and
positions that uninformative are excluded automatically.

\section{Results}

Because we cannot observe directly the underlying RNA-abundance, our evaluation
strategy relies on testing two assumptions we make of an ideal, unbiased experiment.
\begin{enumerate}
\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}) measured
from reads within exons should not differ greatly from frequencies measured by
sampling uniformly within the same exons.
\item Read counts across exons should follow approximately a Poisson process.
\end{enumerate}

We applied our procedure (labeled ``BN'') as well as those of Li, et. al, and
Hansen, et. al., which are implemented in the R packages \texttt{mseq} and
\texttt{Genominator}, respectively. The procedures were applied to four publicly
available data sets \cite{Bullard2010} \cite{Mortazavi2008} \cite{Trapnell2010}
\cite{Wetterbom2010}, as well as an unpublished data set of our own (see, Table
\ref{tab:datasets}).

\begin{tablehere}
\begin{center}
\footnotesize{
\begin{tabular}{lllr}
\textbf{Experiment} & \textbf{Species} & \textbf{Platform} & \textbf{Read Length} \\ \hline
Wetterbom \cite{Wetterbom2010} & Chimpanzee & ABI & 33 \\
Katze (unpublished) & Macaque & ABI & 50 \\
Bullard \cite{Bullard2010} & Human & Illumina & 35 \\
Mortazavi \cite{Mortazavi2008} & Mouse & Illumina & 33 \\
Trapnell \cite{Trapnell2010} & Mouse & Illumina & 75
\end{tabular}
}
\end{center}
\caption{Data sets on which the methods are evaluated.}
\label{tab:datasets}
\end{tablehere}

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b in the Chimpanzee genome). For evaluation, we drew a set
long, highly expressed exons from the remaining chromosomes. In particular, for
each reference sequence, beginning with the set of exons annotated by Ensembl
release 60 \cite{Hubbard2009}, we removed any exons with known alternate splice
sites, then chose the top 1000 exons by read count, restricting ourselves those
at least 100nt long.

The differences in the methods being tested necessitated training procedures
unique to each.

Li recommends that their MART and GLM models be trained using the 100 most abundant
genes. We used 1000 exons from chromosomes 1--8, otherwise chosen in a manner
identical to that which was used to select the test exons.  Both the GLM and
MART model were trained using the default parameters.

Hansen recommends using all of the reads to estimate heptamer frequencies used by
their model. The training procedure works by simple tallying of frequencies and
is thus relatively fast. We choose to use all of the reads from chromosomes 1--8.
We evaluated three variations of this scheme. The first uses the default
parameters, evaluating 7-mer frequencies averaged over the first two positions
of the read (labeled ``Avg-7mer''). We also evaluated estimating 7-mer
frequencies over only the first position, as well as 4-mer frequencies over only
the first position. Though many variations of this scheme may be used, we found these
additional two to be particularly informative. The background frequencies in
each variation were were estimated from positions 18--23 within each read.

All data sets were mapped using Bowtie \cite{Langmead2009} against the latest
genome assemblies available from the UCSC Genome Browser \cite{Rhead2010} during
preparation.

\subsection{Kullback-Leibler Divergence}

Plotting the nucleotide frequencies (Figure  \ref{fig:freqs}) we observe an
obvious bias. To quantify the non-uniformity observed in these plots, we use the
symmetrized Kullback-Leibler (KL) divergence \cite{Kullback1951}. Comparing only
nucleotide frequencies may miss dinucleotide or higher order bias. Thus, for a fair
comparison, we compute the KL divergence for $k$-mers, with $k \in \{ 1 \dots 6
\}$.

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as,
$$D( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
Where the sum is over all $k$-mers. We compute the divergence at each
position relative to the read start, by tallying (overlapping) $k$-mer
frequencies at each position, weighted by the bias coefficient at that position,
and comparing to a background distribution. The background distribution is
estimated by averaging the $k$-mer frequencies at positions $-60 \dots -20$ and
$20 \dots 60$, where the observed bias was very low in each data set.  We further
take the precaution of ignoring any positions with more than 10,000 reads, to
avoid the measure being dominated by any small group of outliers. The only
qualitative effect this was seen to have was in the Katze data set, in which the
frequency plots are otherwise quite skewed.

The result of this this analysis is plotted in Figure
\ref{fig:kl} for the Trapnell data set. Similar plots for the other data sets are
provided in the supplement.

\begin{figurehere}
\begin{center}
\includegraphics[width=0.5\textwidth]{fig/kl.pdf}
\end{center}
\caption{Plots of the KL divergence over overlapping $k$-mers in the Trapnell
experiment.}
\label{fig:kl}
\end{figurehere}

To summarize these plots, we summed the divergence from position -20
through 20, for each of the methods, then normalized this number by dividing by
the summed divergence before adjustment. The results of this analysis are plotted in
Figure \ref{fig:kl-delta}.

\begin{figurehere}
\begin{center}
\includegraphics[width=0.35\textwidth]{fig/kl-delta.pdf}
\end{center}
\caption{Summed KL divergence, normalized by unadjusted divergence, so that a
number less than 1 indicates a decrease in divergence, or an increase in
uniformity.}
\label{fig:kl-delta}
\end{figurehere}

\subsection{Poisson Regression}

In this comparison, we measure how well the counts conform to a Poisson process.
The assumption of positional read counts following a Poisson distribution has
been shown to be a poor fit \cite{Srivastava2010}, yet measuring the improvement in
fit derived from correcting for technical bias remains a principled and easily
interpreted criteria.

We perform maximum-likelihood fitting of two models. In the null model, the
Poisson rate is fixed for each exon. That is, for position $j$ within
gene $i$, the rate is,
$$ \lambda_{ij} = a_i $$
Where $a_i$ is the parameter being fit. For comparison, we then fit a model in
which the rate is proportional to the predicted bias coefficients.
$$ \lambda'_{ij} = a_i b_{ij} $$

If the null model has log-likelihood $L$ and the bias corrected model
$L'$, we measure the change in fit as,
$$ \log \left( \frac{ L_{i} } { L'_{i} } \right) $$
When the fit improves, this number is positive, and negative when it is
worsened. The values computed from our test are listed in Table \ref{tab:pois}.

\begin{tablehere}
\begin{center}
\tiny{
\begin{tabular}{lrrrrrr}
         &\textbf{Avg-7mer}&\textbf{7mer}&\textbf{4mer}&\textbf{GLM}&\textbf{MART}&\textbf{BN} \\
\textbf{Katze}            &  0.004  & 0.017  & 0.055 &0.166  &0.248&0.305 \\
\textbf{Wetterbom}        & -0.076  &-0.058  &-0.002 &0.240  &0.070&0.557 \\
\textbf{Bullard}          &  0.124  & 0.192  & 0.139 &0.285  &0.198&0.380 \\
\textbf{Mortazavi}        &  0.059  & 0.123  & 0.099 &0.284  &0.310&0.401 \\
\textbf{Trapnell}         &  0.094  & 0.163  & 0.104 &0.265  &0.332&0.349 \\
\end{tabular}
}
\end{center}
\caption{Change in fit, compared to the null model of uniform Poisson rate across exons.}
\label{tab:pois}
\end{tablehere}

Computing one number to summarize the fit of the model ignores variance in the
likelihood from exon to exon. This number could be dominated by a few test
regions. Figure \ref{fig:pois} plots the distribution of the change in fit
across the 1000 test exons, revealing a similar, but more nuanced result.

\begin{figurehere}
\begin{center}
\includegraphics[width=0.4\textwidth]{fig/pois.pdf}
\end{center}
\caption{Plotted is the change in fit of each of the test exons. A
positive value indicates a better fit in the bias corrected model.}
\label{fig:pois}
\end{figurehere}

\section{Discussion}

We have demonstrated that sequence bias can confound, sometimes severely,
quantification in RNA-Seq experiments, and we have introduced an effective
method to account for this bias without the need of existing gene annotations.

In our results, estimating initial 4-mer or 7-mer frequencies was not seen to be
as effective as the other models, even when data generated using random hexamer
priming was used. Given the large number of parameters needed to estimate 7-mer
frequencies, a likely explanation is that the model is overfit to the training
set. Though the training set consisted of at least 600,000 reads in
each experiment, it seems more reads, or a greater diversity of reads are needed
to produce an accurate estimate.

Our method generalizes this approach, attempting to overcome this problem by
using an estimation of sequence probability that is more robust to overfitting
and can account for bias beyond the initial heptamer. This approach is
competitive with that of Li, et. al., despite not requiring gene annotations.

Because we do not require annotations, CHiP-Seq, and other high-throughput
sequencing experiments, may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one CHiP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, supplementary). Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in CHiP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into CHiP-Seq peak calling algorithms, and potentially improve
accuracy.

Our method can also very naturally capture dependency between nucleotides in the
sequence. This of particular importance when working with reads generated from
the AB SOLiD platform, in which the di-base, ``colorspace'' coding scheme is
used, encoding two adjacent nucleotides as a single color. A color bias at one
position will manifest itself as a nucleotide bias at two positions. Capturing
this effect requires a model that considers the dependency between nearby
positions. In our results, the MART model, as well as our own, perform
significantly better than the GLM model, illustrating this proposition.

RNA-Seq is most often used to compare levels of expression, and so a natural
concern is the consistency of the bias between samples. In the data we examined,
the bias appears to be largely, but not entirely consistent (see,
supplementary). Similarly, in Figure \ref{fig:freqs}, the three data sets
sequenced on the Illumina platform display similar patterns of nonuniformity,
yet differ in magnitude. Batch effects in RNA-Seq remain a legitimate concern,
that should be evaluated and corrected for.

\section{Availability}

The method is implemented in the R/Bioconductor \cite{Gentleman2004} package \texttt{seqbias}.

\bibliographystyle{plain}
\bibliography{seqbias-paper}

\end{multicols}

\end{document}

### 2011-06-11 seqbias-supplement.tex
starting over, after the great hard drive crash of 2011

\documentclass[letterpaper]{article}
\usepackage{qcjstd}
\usepackage{lscape}
\usepackage{comment}
\usepackage{geometry}
\usepackage{graphicx}
%\usepackage{palatcm}

\geometry{left=1in,right=1in, top=1in, bottom=1in}

\begin{document}

\qcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}

%\begin{center}
%\large{
%\sc{SUPPLEMENTARY}

%\sc{A model for Sequence Bias in RNA-Seq}
%}
%\end{center}

\section{Trimming Reads}

Observing the nonuniform distribution of nucleotide frequencies surrounding the
5' end reads, a natural step to take would be to trim the 5' end before mapping,
in the hope that simply removing the portion of the read in which the bias
occurs will also remove the bias. The figure below demonstrates that this is not
the case. Trimming the initial heptamer in the Mortazavi data set does nothing to
reduce the bias, and simply shifts the plot by seven positions. This indicates
that the issue is \emph{sampling bias}, rather than a bias in base calling.

%\begin{figure}[b]
\begin{center}
\includegraphics[width=0.35\textwidth]{fig/trimmed-freqs.pdf}
\end{center}
%\caption{Nucleotide frequencies for trimmed and untrimmed reads from the
%Mortazavi dataset.}
%\label{fig:trimmed}
%\end{figure}

\section{CHiP-Seq}

Though we found the MART model \cite{Li2010} to be competitive with our own,
yet ours offers the advantage that no gene annotations are needed for training. In
RNA-Seq this is useful in applications of de-novo gene discovery, but it also
allows our method to be applied to CHiP-Seq and another short read data. Here we
examine one publicly available CHiP-Seq data set from Cao, et. al.
\cite{Cao2010}.

Measuring nucleotide frequencies, we find the bias to be significantly less
than that observed the RNA-Seq data sets we tested, but the data was by no means
unbiased.

%\begin{figure}[b]
\begin{center}
\includegraphics[width=0.5\textwidth]{fig/freqs_cao.pdf}
\end{center}
%\caption{Nucleotide frequencies from CHiP-Seq reads.}
%\label{fig:freqs_cao}
%\end{figure}

In the analysis of the RNA-Seq data, we used the assumption of continuous
transcription across annotated exons to evaluate the efficacy of the models, as
well as to train the GLM and MART models. Evaluating bias correction on CHiP-Seq
data necessitates dropping the Poisson regression test, as well as the GLM and
MART models from our comparison, which can not be applied to such data.

We did however repeat our analysis using the Kullback-Leibler divergence.  We
trained each method using reads from reads from chromosomes 1--8. The remaining
chromosomes were segmented into 500nt bins, and the KL divergence was sampled
from the 50,000 segments with the highest read counts.

First we plot directly the positional KL divergence, computed identically

\begin{center}
\includegraphics[width=0.5\textwidth]{fig/cao-kl.pdf}
\end{center}

Secondly, we summarize these plots, computing one normalized divergence
estimate by summing the KL divergence of positions -20 through 20, and
divided by the unadjusted summed divergence to normalize.

\begin{center}
\includegraphics[width=0.5\textwidth]{fig/cao-kl-delta.pdf}
\end{center}

We see that our method is effective at reducing the bias in this ChiP-Seq data.

\section{Variability Between Replicates}

Measuring differential expression or isoform switching is a primary application of
RNA-Seq. If the bias were inconsistent between replicates, it would call into
question the accuracy of such tests. In our experiments, we have observed the
bias to be mostly, but not entirely consistent between replicates. Below, we
plot the frequencies from the four runs in the Trapnell data set.

\begin{center}
\includegraphics[width=0.5\textwidth]{fig/replicates.pdf}
\end{center}

Though we do not know if the variability is always minor, assuming this is so,
there remains the risk that the sequence bias would greatly effect the depth to
which a locus is sequenced, and thus the statistical significance of
differential expression tests. This would bias the discovery of differentially
expressed genes, but not the test itself.

\section{Additional Plots}

Below we plot the adjusted and unadjusted KL divergence of each sample.

%\begin{figure}[b]
%\begin{center}
\includegraphics[width=\textwidth]{fig/kl-all.pdf}
%\end{center}
%\caption{Positional KL divergence for each dataset and method.}
%\label{fig:kl-all}
%\end{figure}

\bibliographystyle{plain}
\bibliography{seqbias-supplement}

\end{document}

### 2011-06-12 seqbias-paper.tex
compiling in bioinformatics template

@@
-\documentclass[letterpaper]{article}
-\usepackage{qcjstd}
-\usepackage{graphicx}
-\usepackage{lscape}
-\usepackage{multicol}
-\usepackage{geometry}
-\usepackage{amsmath}
-\usepackage{multirow}
-%\usepackage{bookman}
-%\usepackage{cmbright}
+\documentclass{bioinfo}
+\copyrightyear{2011}
+\pubyear{2011}
 
-\makeatletter
-\newenvironment{tablehere}
-  {\def\@captype{table}}
-  {\vspace{1em}}
-
-\newenvironment{figurehere}
-  {\def\@captype{figure}}
-  {\vspace{1em}}
-\makeatother
-
-\geometry{left=0.8in,right=0.8in, top=1in, bottom=1in}
-
-\title{A Model for Sequence Bias in RNA-Seq}
+\usepackage{comment}
 
 \begin{document}
+\firstpage{1}
 
-%\maketitle
+\title{Correcting for bias in high throughput sequencing data}
+\author[Jones \textit{et~al}]
+{Daniel C. Jones\,$^{1,}$
+\footnote{to whom correspondence should be addressed}\hspace{0.5em}
+Walter L. Ruzzo\,$^{1,2}$
+Xinxia Peng\,$^{3}$
+Michael G. Katze$^{3}$
+}
 
-\qcjtitle{\sc{(Draft)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\address{
+$^{1}$Deportment of Computer Science and Engineering, University of
+Washington, Seattle, WA 98195-2350, USA\\
+$^{2}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109\\
+$^{3}$Department of Microbiology, University of Washington, Seattle, WA
+98195-7242, USA}
+
+\editor{Associate Editor: XXXXXXX}
+
+\maketitle
 
@@
 \end{abstract}
-
-\begin{multicols}{2}
 
@@
 
+\begin{comment}
 \begin{figure*}
@@
 \end{figure*}
+\end{comment}
 
@@
 
+%% TODO
+\begin{comment}
 \begin{tablehere}
@@
 \end{tablehere}
+\end{comment}
 
@@
 
+%% TODO
+\begin{comment}
 \begin{figurehere}
@@
 \end{figurehere}
+\end{comment}
 
@@
 
+\begin{comment}
 \begin{tablehere}
@@
 
-\end{multicols}
-
 \end{document}

### 2011-06-12 seqbias-paper.tex
redo freqs figure, etc

@@
 Washington, Seattle, WA 98195-2350, USA\\
-$^{2}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109\\
+$^{2}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA\\
 $^{3}$Department of Microbiology, University of Washington, Seattle, WA
@@
 \begin{abstract}
-Quantification of RNA abundance in RNA-Seq experiments is often conflated with
-protocol specific bias as a result of PCR amplification, differing primer
-affinities, and random priming with a non-uniform distribution of primer
-sequences.  We propose a method to measure and compensate for these influences
-using a simple graphical model, trained without the need of existing gene
-annotations.
+
+\section{Motivation:}
+
+Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
+often conflated with protocal-specific bias, influenced by PCR amplification,
+and differing primer affinities and mixtures, for example. The result is
+potentially inaccurate quantification, affecting de novo gene annotation and
+isoform quantification in RNA-Seq, and peak calling in ChiP-Seq.
+
+\section{Results:}
+
+We present a method to measure and correct for these influences using a simple
+graphical model. Our model does not rely on existing gene annotations, making it
+applicable to any high throughput sequencing data. We evaluate our method on
+several datasets, and by multiple criteria, demonstrating that it effectively
+decreases bias and increases uniformity.
+
+\section{Availability:}
+The method is implemented in the \texttt{seqbias} R/Bioconductor package,
+available freely under the LGPL licence from
+\href{http://bioconductor.org}{http://bioconductor.org}.
+
+\section{Contact:}
+\href{dcjones@cs.washington.edu}{dcjones@cs.washington.edu}
+
 \end{abstract}
@@
 microarrays in quantifying RNA abundance. But, as microarray technology has
-brought with it technical challenges ranging from robust normalization to
-addressing cross-hybridization, RNA-Seq presents a new set of problems.
+brought with it technical challenges ranging from developing robust
+normalization to accounting for cross-hybridization, RNA-Seq presents a new set
+of problems.
 
-A significant example is the effect of transcript length and quantification.
+A significant example is the effect of transcript length on quantification.
 As noted by Oshlack, et. al., transcript length biases discovery of differentially
@@
 in modeling splice cites \cite{Cai2000} \cite{Chen2005} and transcription factor
-binding sites \cite{BenGal2005} \cite{Grau2006} \cite{Pudimat2005}.
+binding sites \cite{Ben-Gal2005} \cite{Grau2006} \cite{Pudimat2005}.
 
@@
 All data sets were mapped using Bowtie \cite{Langmead2009} against the latest
-genome assemblies available from the UCSC Genome Browser \cite{Rhead2010} during
-preparation.
+genome assemblies available from the UCSC Genome Browser \cite{Karolchik2008}
+during preparation.
 
@@
 
-\section{Availability}
-
-The method is implemented in the R/Bioconductor \cite{Gentleman2004} package \texttt{seqbias}.
-
-\bibliographystyle{plain}
+\bibliographystyle{natbib}
 \bibliography{seqbias-paper}

### 2011-06-12 build.sh

@@ -1,0 +1,18 @@
+#!/bin/sh
+
+# Beacuse the makefile is fucking hopeless.
+
+
+export TEXINPUTS=.:bioinformatics:
+export BSTINPUTS=.:bioinformatics:
+
+TEX=pdflatex
+BIB=bibtex
+
+
+$TEX seqbias-paper.tex
+$BIB seqbias-paper
+$TEX seqbias-paper.tex
+
+
+

### 2011-06-13 seqbias-paper.tex
overview figure, recreated

@@
 under reported. They render untrustworthy comparisons of relative abundance
-between genes, and any test of differential expression hangs on the assumption
+between genes, and any test of differential expression hangs on the assuBmption
 that these biases are identical between replicates.  Additionally, in many tests
@@
 
-\begin{comment}
 \begin{figure*}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{fig/freqs.pdf}
-\end{center}
-\caption{Nucleotide frequencies relative to the start of each read (labeled
-position 0), respecting strand. The divergence of the nucleotide distribution
-is summarized using the symmetrized Kullback-Leibler divergence (see Section
-3).}
+\centerline{\includegraphics[width=0.8\textwidth]{fig/freqs.eps}}
+\caption{Nucleotide frequencies are plotted relative to the start (labeled,
+position 0) of each mapped read, respecting strand. The sequence is taken frmo
+the genomic context surrounding the read, so that -40 to -1, for example, fall
+outside the read sequence itself. The symmetrized Killback-Leibler divergence is
+used to summarize the difference in nucleotide frequency compared to a fixed
+estimate of background necleotide frequencies made by sampling many positions
+nearby mapped reads.  Under the assumption that reads are sampled uniformly from
+transcripts, each of the plots should be essentially flat.}
 \label{fig:freqs}
 \end{figure*}
-\end{comment}
 
@@
 
-\section{Method}
+\begin{figure}
+\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
+\caption{An overview of the approach taken: (a) foreground sequences are sampled
+from the regions surrounding the starts of mapped reads, (b) background
+sequences are sampled by randomly offsetting foreground positions, (c) a
+Bayesian network is classifier is trained to descriminate between the set of
+sampled foreground and background sequences, (d) and the classifier is evaluated
+at each position within a locus, predicting bias.}
+\label{fig:overview}
+\end{figure}
+
+\section{Methods}
 

### 2011-06-13 results/plots/plot-dg.py
a new models graph figure

@@ -1,0 +1,213 @@
+#!/usr/bin/env python
+
+'''
+This a simple program to plot directed graphs in a compact, readable fashion,
+similar to circos (http://circos.ca/).
+'''
+
+import pydot
+import cairo
+import argparse
+from math        import pi, radians, sin, cos, atan2, sqrt
+from collections import defaultdict
+
+
+def guess_radius():
+    return min(args.width, args.height) / 2.0 - 50
+
+
+def draw_nodes(G, ctx):
+    '''
+    Draw labeled nodes in a big circle, given a pydot graph and a cairo context
+    object.
+    '''
+
+    us = [u for u in G.get_nodes() if u.get_label() is not None]
+    us.sort(key = lambda u: -int(u.get_label().strip('"')))
+
+    n = len(us)
+    r = guess_radius()
+    k = (2.0 * pi - radians(args.gap)) / n
+
+    for (i, u) in enumerate(us):
+        label = u.get_label().strip('"')
+        style = u.get_style().strip('"')
+
+        theta = (i + 0.5) * k + radians(args.gap) / 2.0
+
+        x, y = r * sin(theta), r * cos(theta)
+
+        ctx.arc(x, y, args.node_rad, 0.0, 2 * pi)
+
+        if style == 'solid':
+            ctx.stroke_preserve()
+            ctx.arc(x, y, args.node_rad, 0.0, 2 * pi)
+            ctx.fill()
+        else:
+            ctx.stroke()
+
+
+
+        xt = (r + 2 * args.node_rad) * sin(theta)
+        yt = (r + 2 * args.node_rad) * cos(theta)
+
+        ctx.save()
+
+        ctx.translate(xt, yt)
+        ctx.rotate(pi - theta)
+
+        ctx.move_to(-ctx.text_extents(label)[4] / 2.0, 0)
+        ctx.text_path(label)
+        ctx.fill()
+
+        ctx.restore()
+
+        # TODO: draw label
+
+
+def ed((x0, y0), (x1, y1)):
+    ''' Euclidean distance. '''
+    return sqrt((x0 - x1) ** 2 + (y0 - y1) ** 2)
+
+
+def draw_edges(G, ctx):
+    us = [u for u in G.get_nodes() if u.get_label() is not None]
+    us.sort(key = lambda u: -int(u.get_label().strip('"')))
+
+    node_pos = dict(zip([u.get_name() for u in us], range(len(us))))
+
+    n = len(us)
[…]

### 2011-06-14 results/plots/config.R
more plotting

@@ -1,0 +1,4 @@
+
+
+d <- '../20110610/'
+

### 2011-06-14 results/plots/plot-kl.R

@@ -1,0 +1,88 @@
+
+
+
+
+source('config.R')
+
+ws <- c('bullard.tab', 'katze.tab', 'mortazavi.tab',
+        'trapnell.tab', 'wetterbom.tab')
+
+xs <- NULL
+for (w in ws) {
+    x  <- read.table(paste(d, '/kl/', w, sep = ''), header = T)
+    xs <- rbind(xs, x)
+}
+
+
+# reorder this factor so the plot is drawn in a flattering manner
+xs$method <- factor(xs$method, levels = c(
+                                          'Unadjusted',
+                                          'Hansen/A',
+                                          'Li/GLM',
+                                          'Li/MART',
+                                          'Jones'
+                                          ))
+
+# reorder this factor so that it is drawn in the same order as every other plot
+xs$sample <- factor(xs$sample, levels = c(
+                                         'Wetterbom',
+                                         'Katze',
+                                         'Bullard',
+                                         'Mortazavi',
+                                         'Trapnell'))
+
+library(Cairo)
+library(ggplot2)
+source('theme_dcjstd.R')
+
+
+xs <- subset(xs, k == 1 | k == 4)
+xs <- subset(xs, -20 <= pos & pos <= 20)
+
+
+scl <- 2.8
+CairoSVG('kl.svg', width = 6 * scl, height = 1.8 * scl)
+
+xs$k <- factor(xs$k)
+
+p <- qplot(data = xs, x = pos, y = div, color = method,
+           geom = 'line')
+p <- p + facet_grid(k ~ sample, scale = 'free')
+
+
+p <- p + scale_color_manual(
+            name = 'Method',
+            labels = list(
+                          'Jones'    = 'BN',
+                          'Li/MART'  = 'MART',
+                          'Li/GLM'   = 'GLM',
+                          'Hansen/A' = '7mer',
+                          'Unadjusted' = 'Unadjusted'
+                          ),
+            values = c('grey70',
+                       '#f0c956ff',
+                       '#b8359eff',
+                       '#357db8ff',
+                       '#e31919ff'
+                       ))
+
+
+                             
+
+# get rid of grid lines, since they get fucket up on this plot anyway
+p <- p + theme_dcjstd()
+
+#p <- p + structure(list(panel.grid.minor = theme_blank(),
+                        #panel.grid.major = theme_blank()),
+                   #class = "options")
+
+print(p)
+
+dev.off()
+
+
+
+# TODO
+
+
+

### 2011-06-14 results/plots/plot-models.sh

@@ -1,0 +1,11 @@
+#!/bin/sh
+
+
+
+for f in `ls ../20110610/dot/*.dot`
+do
+    g="model."`basename $f | sed s/dot/svg/`
+    echo "$f -> $g"
+    ./plot-dg.py --output=$g $f
+done
+

### 2011-06-14 results/plots/plot-pois.R

@@ -1,0 +1,77 @@
+
+
+source('config.R')
+
+# build dataframe
+
+ws <- c('bullard.tab', 'katze.tab', 'mortazavi.tab',
+        'trapnell.tab', 'wetterbom.tab')
+
+xs <- NULL
+for (w in ws) {
+    x  <- read.table(paste(d, '/pois/', w, sep = ''), header = T)
+    xs <- rbind(xs, x)
+}
+
+
+
+# join with unadjusted likelihood for easy comparison
+ys <- subset(xs, method == 'Unadjusted')[,c('sample', 'gene', 'L')]
+xs <- merge(xs, ys, by = c('sample', 'gene'))
+names(xs) <- c('sample', 'gene', 'method', 'mu', 'L', 'L.base')
+
+
+
+# mcfadden's pseudo-R2
+r2 <- function(L, L.base) { 1 - L / L.base }
+
+
+
+
+# the boxplots are plotted seperately, then combined, so that they can be put no
+# seperate scales.
+
+library(Cairo)
+library(ggplot2)
+source('theme_dcjstd.R')
+
+
+#CairoPNG('pois-boxplot-simple.png',
+         #width = 600, height = 2000)
+
+#p <- qplot(data = xs, y = r2(L, L.base), x = method, geom = 'boxplot')
+#p <- p + facet_grid(sample ~ .)
+#p <- p + coord_flip()
+#print(p)
+
+#dev.off()
+
+
+# reorder
+xs$method <- factor(xs$method,
+                    levels = c('Hansen/A', 'Li/GLM', 'Li/MART', 'Jones'))
+
+
+
+for (w in levels(xs$sample)) {
+
+    ys <- subset(xs, sample == w & method != 'Unadjusted')
+    print(head(ys))
+
+    CairoSVG(paste('pois-boxplot', w, 'svg', sep = '.'),
+             width = 4, height = 1.5)
+
+    p <- qplot(data = ys, y = r2(L, L.base), x = method,
+               geom = 'boxplot', outlier.colour = alpha('black', 0.3))
+    p <- p + scale_x_discrete(name   = 'Method',
+                              labels = list('Hansen/A' = '7mer',
+                                            'Li/MART'  = 'MART',
+                                            'Li/GLM'   = 'GLM',
+                                            'Jones'    = 'BN'))
+    p <- p + coord_flip()
+    p <- p + theme_dcjstd()
+    print(p)
+
+    dev.off()
+}
+

### 2011-06-14 results/plots/theme_dcjstd.R

@@ -1,0 +1,106 @@
+
+
+
+theme_dcjstd <- function()
+{
+    structure(
+        list(
+            panel.background = theme_rect(fill   = '#f5f5f5',
+                                          colour = '#e6e6e6'),
+
+            panel.grid.major = theme_line(colour = "white"), 
+
+            panel.grid.minor = theme_line(colour = "white", size = 0.25)
+        ),
+
+        class = "options"
+    )
+}
+
+
+#theme_grey <-
+#function (base_size = 12, base_family = "") 
+#{
+    #structure(
+        #list(
+            #axis.line = theme_blank(),
+
+            #axis.text.x = theme_text(family     = base_family, 
+                                     #size       = base_size * 0.8,
+                                     #lineheight = 0.9,
+                                     #colour     = "grey50", 
+                                     #vjust      = 1),
+
+            #axis.text.y = theme_text(family     = base_family, 
+                                     #size       = base_size * 0.8,
+                                     #lineheight = 0.9,
+                                     #colour     = "grey50", 
+                                     #hjust      = 1),
+
+            #axis.ticks = theme_segment(colour = "grey50"), 
+
+            #axis.title.x = theme_text(family = base_family,
+                                      #size   = base_size, 
+                                      #vjust  = 0.5),
+
+            #axis.title.y = theme_text(family = base_family, 
+                                      #size   = base_size,
+                                      #angle  = 90,
+                                      #vjust  = 0.5),
+
+            #axis.ticks.length = unit(0.15, "cm"),
+
+            #axis.ticks.margin = unit(0.1, "cm"),
+            
+            #legend.background = theme_rect(colour = "white"), 
+
+            #legend.key = theme_rect(fill = "grey95", colour = "white"), 
+
+            #legend.key.size = unit(1.2, "lines"), legend.key.height = NA, 
+
+            #legend.key.width = NA,
+            
+            #legend.text = theme_text(family = base_family, 
+                                     #size   = base_size * 0.8),
+
+            #legend.text.align = NA, 
+
+            #legend.title = theme_text(family = base_family,
+                                      #size   = base_size * 0.8,
+                                      #face   = "bold",
+                                      #hjust  = 0),
+                                      
+            #legend.title.align = NA, 
+
+            #legend.position = "right",
+
+            #legend.direction = "vertical", 
+
+            #legend.box = NA,
+            
[…]

### 2011-06-14 seqbias-paper.tex
a new models graph figure
poisson boxplot
more plotting
text revisions

@@
 \section{Motivation:}
-
 Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
@@
 \section{Results:}
-
 We present a method to measure and correct for these influences using a simple
@@
 
-A significant example is the effect of transcript length on quantification.
-As noted by Oshlack, et. al., transcript length biases discovery of differentially
-expression, since more reads are aligned to longer transcripts, granting
-higher statistical significance in tests differential expression
-\cite{Oshlack2009}. Recently, some methods have been developed attempting to account for
-this effect in gene set enrichment tests \cite{Young2010} \cite{Gao2011}.
-Additionally, differing transcript lengths introduce ``edge
-effects'', as very short fragments from the end of a transcript will not be
-sequenced. NEUMA \cite{Lee2010} and FPKM \cite{Trapnell2010} are two methods
-that attempt to take this into account.
+A significant example is the effect of transcript length on quantification.  As
+noted by \citet{Oshlack2009} transcript length biases discovery of
+differentially expression, since more reads are aligned to longer transcripts,
+granting higher statistical significance in tests differential expression.
+Recently, some methods have been developed attempting to account for this effect
+in gene set enrichment tests \citep{Young2010,Gao2011}.
+
+%% TODO: do something with this sentance.
+Additionally,
+differing transcript lengths introduce ``edge effects'', as very short fragments
+from the end of a transcript will not be sequenced. NEUMA \cite{Lee2010} and
+FPKM \cite{Trapnell2010} are two methods that attempt to take this into account.
 
 The underlying nucleotide sequence also influences quantification in a manner
-that is complex and protocol specific. In an ideal experiment, the number of
+that is complex and protocol-specific. In an ideal experiment, the number of
 reads mapping to a particular position in the genome is a function of RNA
-abundance and should not be additionally dependent on the sequence. As pointed
-out recently, this is not the case \cite{Dohm2008} \cite{Li2010}
-\cite{Hansen2010}. As illustration, Figure \ref{fig:freqs} plots this
-nonuniformity in nucleotide frequencies on five data sets (see, Table
-\ref{tab:datasets}), each using a different protocol.
+abundance and should not be additionally dependent on the sequence at that
+position. As noted first by \citet{Dohm2008}, this is not the case.  As
+illustration, Figure \ref{fig:freqs} plots this nonuniformity in nucleotide
+frequencies on five data sets (see, Table \ref{tab:datasets}), each using a
+different protocol.
 
@@
 under reported. They render untrustworthy comparisons of relative abundance
-between genes, and any test of differential expression hangs on the assuBmption
-that these biases are identical between replicates.  Additionally, in many tests
-of differential expression---similarly the effects of transcript length---a higher
-read count will result in higher statistical confidence. It follows that the
-sensitivity of such a test will also be biased by sequence, affecting downstream
-systems biology analysis such as gene ontology enrichment tests.
+between genes or isoforms, and any test of differential expression hangs on the
+assumption that these biases are identical between replicates.  Additionally, in
+many tests of differential expression---similar to the effects of transcript
+length---a higher read count will result in higher statistical confidence. It
+follows that the sensitivity of such a test will also be biased by sequence,
+affecting downstream analysis such as gene ontology enrichment tests.
 
 This bias, though observed primarily in the 5' end of a read, is not resolved by
-trimming the reads prior to mapping (see, supplementary Figure 1), suggesting
-it is not a result of erroneous base calling, and that a more sophisticated
-means of correction is needed.
+trimming the reads prior to mapping \citep{Hansen2010} (see also, Supplementary
+Section 1), suggesting it is not a result of erroneous base calling, and that a
+more sophisticated means of correction is needed.
 
@@
 
-Some work has been done to model and correct for these biases.  Li, et. al.,
-\cite{Li2010} propose two models. The firsts is a Poisson linear model, in which
-read counts across a transcript follow inhomogeneous Poisson process. The read
-count at position $i$ within the transcript is Poisson distributed with
-parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of weights determined by the
-nucleotide at each position surrounding the read start, in addition to a term
-capturing the abundance of the transcript.
+Li, et. al., \cite{Li2010} propose two models. The first is a Poisson linear
+model, in which read counts across a transcript follow inhomogeneous Poisson
+process. The read count at position $i$ within the transcript is Poisson
+distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
+weights determined by the nucleotide at each position surrounding the read
+start, in addition to a term capturing the abundance of the transcript.
 
 The second model is based on multiple additive regression trees, or MART
-\cite{Friedman2003}.  In their tests, MART shows a moderate improvement over the
-Poisson linear model. Both models are fit to a number of abundant test genes,
-requiring existing gene annotations for the reference genome.
+\citep{Friedman2003}.  In their tests, the MART model shows a moderate
+improvement over the Poisson linear model. Both models are fit to a number of
+abundant test genes, requiring existing gene annotations for the reference
+genome.
 
-Another model, proposed by Hansen, et. al., \cite{Hansen2010} directly estimates
-the distribution of initial heptamers within reads, then estimates a presumed
-background heptamer distribution by sampling from the end of reads. Read counts
-are then adjusted by the ratio of the two. Specifying two distributions over
-heptamers (i.e. foreground and background distributions) requires $2(4^7-1) =
-32,766$ parameters, so while no gene annotations are needed to train such a
-model, a significant number of reads are required.
+Another model, proposed by \citet{Hansen2010}, directly estimates the
+distribution of initial heptamers within reads, then estimates a presumed
+background heptamer distribution, sampled from the end of reads. The read count
+at a given position is then adjusted by the ratio of the foreground and
+background heptamer probabilities. Specifying two distributions over heptamers
+(i.e. foreground and background distributions) requires $2(4^7-1) = 32,766$
+parameters, so while no gene annotations are needed to train such a model, a
+significant number of reads are required.
 
-Version 0.9.0 of Cufflinks \cite{Trapnell2010} also introduces a model to
-correct for sequence bias. No description of their method is yet available, nor
-is there any implementation of the model that can be run independently of
-Cufflinks to facilitate testing, so we have withheld any comparison in our
-tests.
+Lastly, \citet{Roberts2011} have very recently published a description of
+another approach, in which sequence probabilities are modeled by variable-order
+Markov chains. The structure of these Markov chains are hard-coded, chosen in
+advance using a hill-climbing algorithm on a representative dataset. This method
+is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
+tightly incorporated into its estimation of transcript abundance.
 
-Besides the direct influence of nucleotide composition on sequencing efficiency,
-the ``uniqueness'' of a sequence within the genome affects how likely it is to
-be mapped correctly, thus influencing quantification in a non-trivial way. Some
-methods exist attempting to quantify uniqueness \cite{Koehler2010} and to take
-it into account in the measure of transcript abundance \cite{Li2010b}
-\cite{Lee2010}. Though we make no attempt to address this problem, any
-probabilistic characterization of uniqueness or ``mapability'' could easily be
-combined with the probabilistic characterization of sequence bias that we
-described here.
+Here we propose a new approach, using Bayesian networks to model sequence
+probabilities. Unlike the methods of Roberts or Li, our model requires no gene
+annotations, nor even the assumption that the short reads are derived from RNA.
+In this sense, we build on the work done by \citet{Hansen2010}, generalizing
+their approach in a way we find to be more robust and effective at correcting
+for bias in a variety of protocols. Due to the weak assumptions required my our
+model, it is applicable and potentially useful in any setting in which short
+reads are aligned to a reference sequence.
 
+%% TODO: where should this figure be cited?
 \begin{figure}
@@
 
-We assume a simple model of an RNA-Seq experiment: the number of reads $x_i$
-aligned to genomic position $i$ is an unbiased estimate of RNA abundance.
-Furthermore, we assume reads may be treated as i.i.d. samples. That
-is, if $N$ reads are generated, and $m_i$ is the event that a generated read
-maps to position $i$,
-$$ E[ x_i ] = N \Pr[ m_i ] $$
+We begin with a natural model of an RNA-Seq experiment (and one that is often
+assumed, whether implicitly or otherwise). The number of reads $x_i$ aligned to
+genomic position $i$ is an unbiased estimate of RNA abundance. Furthermore, we
+assume reads may be treated as independent and identically distributed samples.
+That is, if $N$ reads are generated, and $m_i$ is the event that a generated
+read maps to position $i$.
+$$ E[x_i] = N \Pr[m_i] $$
 
-The experiment may be considered unbiased  with regards to sequence if having
+The experiment may be considered unbiased  with regards to sequence if, having
 observed the nucleotide sequence $s_i$ surrounding position $i$,
@@
 
-This suggests a natural reweighing scheme. First, define the \emph{sequence
-bias} $b_i$ at position $i$ as,
+This suggests a natural scheme in which observations may be reweighted to
+correct for bias.  First, define the \emph{sequence bias} $b_i$ at position $i$
+as,
 $$ b_i = \frac{ \Pr[ s_i ] } { \Pr[ s_i | m_i ] } $$
 
-Now, if we reweight the read count at position $i$ by $b_i$, we
+Now, if we reweight the read count $x_i$ at position $i$ by $b_i$, we
 have,
@@
 \end{align*}
-Thus the reweighted read counts are made to be unbiased.
+Thus the reweighted read counts are made unbiased.
 
@@
 probability $\Pr[s_i]$ and the foreground sequence probability $\Pr[ s_i | m_i
-]$, the later being the probability of the sequence given a read being sampled
-from its position. Any estimation of sequence probabilities presents a challenge
-of finding a model that sufficiently complex to capture the common features of
-the training data but avoids overfitting.
+]$, the latter being the probability of the sequence given a read being sampled
+from its position. Estimating bias is therefore a problem of finding a model of
+sequence probability that sufficiently complex to capture the common features of
+the training data yet avoids overfitting.
 
@@
 The Bayesian network can then used to evaluate sequence probability, and thus
-bias at any genomic position.
+bias, at any genomic position.
 
-This discussion so far ignores one complication. The RNA abundance that we wish to
-estimate is not itself independent of of the nucleotide sequence. Notably,
-exonic DNA is tends to be more GC-rich than intergenic DNA. If background
-sequences are sampled uniformly from the genome we run the risk of incorrectly
-adjusting for biological sequence bias, rather than technical sequence bias. To
-avoid this, we propose using paired training data. Each foreground training
-sequence is paired with a background sequence taken from nearby position that is
-likely to have similar abundance and general nucleotide composition.
+However, we have so for ignored one complication: the RNA abundance that we wish
+to estimate is not itself independent of the nucleotide sequence. Notably,
+exonic DNA tends to be more GC-rich than intergenic DNA. If background sequences
+are sampled uniformly from the genome we run the risk of incorrectly adjusting
+for biological sequence bias, rather than techncical sequence bias.  To avoid
+this, we propose using paired training data. Each foreground training
+sequence is paired with a background sequence taken from nearby position
+that is likely to have similar abundance and general nucleotide composition.
 Alternatively, we could pair foreground samples with background samples from
-within the same transcript, but we prefer to avoid dependence on existing gene
-annotations as such a model would be less suitable for de-novo discovery or
-annotation applications.
+within the same transcript, but we prefer to avoid dependence on existing
+gene annotations as such a model would be limited to RNA-Seq analysis in which
+trustworthy and reasonably complete gene annotations are available.
 
-Hansen, et. al., \cite{Hansen2010} implicitly characterizes and corrects for
-bias in the same manner proposed here. Their approach differs primarily in how
-sequence probabilities are estimated. Li, et. al., \cite{Li2010} offers a
-different perspective, taking advantage of gene annotations to train a model
-that optimally explains variance in base level counts across transcripts.
+The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
+correction as a problem of estimating foreground and background sequence
+probabilities. They differ primarily in how these sequence probabilities are
+estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
+notation) directly, given training data consistig of long, highly-expressed
+transcripts.
 
@@
 each node represents a position in the sequence, relative to the read start, and
-edges encode dependency between positions. Bayesian networks have been
-applied to recognize motifs in nucleotide sequences in the past, in particular
-in modeling splice cites \cite{Cai2000} \cite{Chen2005} and transcription factor
-binding sites \cite{Ben-Gal2005} \cite{Grau2006} \cite{Pudimat2005}.
+edges encode dependency between positions.  Bayesian networks have been applied
+to recognize motifs in nucleotide sequences in the past, in particular in
+modeling splice cites \citep{Cai2000, Chen2005} and transcription factor binding
+sites \citep{Ben-Gal2005, Grau2006, Pudimat2005}.
 
@@
 position $i_k$ as $s_{i_k}$.
+
+%% TODO: leaving it here for now
+%% -------------------------------------------------------
 
@@
 
-\begin{comment}
-\begin{figure*}
-\begin{center}
-\includegraphics[width=0.75\textwidth]{fig/structure2.pdf}
-\end{center}
-\caption{The network structure learned on the Trapnell data set, where 0 is the
-read start. Dotted boxes represent positions not included in the model (i.e.
-positions deemed uninformative).}
-\label{fig:structure}
-\end{figure*}
-\end{comment}
-
 Figure \ref{fig:structure} shows an example of the structure learned when this
@@
 \section{Results}
+
+\begin{figure*}
+\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
+\caption{Blah blah blah.}
+\label{fig:models}
+\end{figure*}
 
@@
 
-%% TODO
-\begin{comment}
-\begin{figurehere}
-\begin{center}
-\includegraphics[width=0.5\textwidth]{fig/kl.pdf}
-\end{center}
+\begin{figure*}
+\centerline{\includegraphics[width=0.9\textwidth]{fig/kl.eps}}
 \caption{Plots of the KL divergence over overlapping $k$-mers in the Trapnell
@@
 \label{fig:kl}
-\end{figurehere}
-\end{comment}
+\end{figure*}
 
@@
 %% TODO
-\begin{comment}
-\begin{figurehere}
-\begin{center}
-\includegraphics[width=0.4\textwidth]{fig/pois.pdf}
-\end{center}
+\begin{figure}
+\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
 \caption{Plotted is the change in fit of each of the test exons. A
@@
 \label{fig:pois}
-\end{figurehere}
-\end{comment}
+\end{figure}
 

### 2011-06-14 build.sh

@@ -13,6 +13,7 @@ BIB=bibtex
 $TEX seqbias-paper.tex
 $BIB seqbias-paper
 $TEX seqbias-paper.tex
+$TEX seqbias-paper.tex
 
 
 

### 2011-06-14 seqbias-paper.tex
more revisions

@@
 to recognize motifs in nucleotide sequences in the past, in particular in
-modeling splice cites \citep{Cai2000, Chen2005} and transcription factor binding
+modeling splice sites \citep{Cai2000, Chen2005} and transcription factor binding
 sites \citep{Ben-Gal2005, Grau2006, Pudimat2005}.
@@
 
-To begin, we generate $n$ training examples, where $n = 200,000$ is typical, half of
-which labeled as foreground, the other background. To do so, we sample the
-positions of $n/2$ reads. Duplicate reads are ignored, to avoid overfitting to
-any exceedingly highly expressed locus.  For each foreground position $i$
-sampled, we then sample a background position by drawing a random offset $\delta
-\sim \text{Normal}( 0, \sigma )$, where by default $\sigma = 50$ is chosen, and
-taking $i' = \lceil i + \delta \rceil$. By using such a scheme we attempt to
-mitigate the effects of biological sequence bias, sampling positions that are
-more likely to be biologically similar.
+The model is trained on $n$ sequences, one half labeled as foreground, the other
+background, sampled from the reference genome. To obtain the foreground
+sequences, we take sequences surrounding the starts positions of $n/2$ aligned
+reads. We wish to avoid overfitting to any short locus with a large abundance of reads,
+yet we must also capture the bias introduced by PCR amplification. We therefore
+choose the $n/2$ unique reads with the highest number of duplicates,
+yet count each read only once, ignoring duplicates
 
-This sampling procedure produces a set of $n$ genomic positions with
-accompanying labels $T = \{ (i_1,x_1), (i_2,x_2), \dots, (i_n,x_n) \}$. The
-label $x_k$ is binary, indicating classification as background ($x_{i_k} =
-0$) or foreground ($x_{i_k} = 1$). We will denote the nucleotide sequence at
-position $i_k$ as $s_{i_k}$.
+To obtain background training sequences, we randomly offset the positions from
+which the foreground sequences were sampled.  The offset is drawn from a
+zero-mean gaussian, and rounded to the nearest integer, away from zero.
+By using such a scheme, we attempt to mitigate the effects of biological
+sequence bias, sampling positions that are more likely to be biologically
+similar.
 
-%% TODO: leaving it here for now
-%% -------------------------------------------------------
+This procuder produces a training set of $n$ sequences with accompanying labels
+$T = \{ (s_1, x_1), (s_2, x_2), \dots, (s_n, x_n) \}$. The label $x_i$ is
+binary, indicating classification as background ($x_i = 0$) or foreground ($x_i
+= 1$).
 
@@
 To determine the structure and parameters of the Bayesian network, we use a
-hill-climbing approach similar to BNC algorithm \cite{Grossman2004}.
+hill-climbing approach similar to algorithm described by \citet{Grossman2004}.
 The network structure is determined by greedily optimizing the conditional
@@
 \begin{align*}
-\ell &= \sum_{k=1}^{n} \log \Pr( x_{i_k} | s_{i_k} ) \\
+\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
 &=
-\sum_{k=1}^{n} \log \frac{ \Pr(s_{i_k} | x_{i_k}) \Pr( x_{i_k} ) }{
-\sum_{x \in \{0,1\}} \Pr( s_{i_k} | x ) \Pr(x) } \\
-%&= \sum_{k=1}^{n} \log \frac{ \Pr( s_{i_k} | x_{i_k} ) }
-%{ \sum_{x \in \{0,1\}} \Pr( s_{i_k} | x ) }
+\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
+\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
 \end{align*}
@@
 set of samples, simply maximizing the likelihood would severely overfit the
-training set. We thus penalize model complexity using the Akaike information
-criterion with the second order sample size correction \cite{Hurvich1989}.
-Where $m$ is the number of parameters needed to specify the model, we maximize,
-$$ \ell' = \ell - m - \frac{(m + 1)(m + 2)}{n - m - 2} $$
+training set. We thus penalize model complexity heuristically using the Akaike
+information criterion. Where $m$ is the number of parameters needed to specify
+the model, we maximize,
+
+$$ \ell' = \ell - m $$
+
+Some benifit would be obtained from a more highly tuned complexity penalty.
+However, since the model is trained greedily, additional parameters will be
+decreasingly informitive, and increasingly similar between foreground and
+background. Adding more parameters will have little effect.  Only when $m$ is
+allowed to grow exponentially does the prediction become polluted by small
+deviations between thousands of uninformitive parameters.
 
 At each step of the optimization procedure, every possible edge or position
-addition, removal, or edge reversal that produces a valid network (i.e. does not
+addition, removal, or edge reversal that produces a valid network (i.e., does not
 introduce a cycle) is evaluated, and the alteration that increases the score
-$\ell'$ the most is kept.  This process is repeated until a local maxima is
+$\ell'$ the most is kept.  This process is repeated until a local maximum is
 found, in which no single alteration to the network will increase the score.
@@
 Given the network structure, the parameters are estimated directly from the
-observed nucleotide frequencies. This choice of parameters maximizes the
-likelihood of the model, but not necessarily the conditional likelihood. They
-are equivalent asymptotically, thus, as per the BNC algorithm, we adopt this
-approximation rather than  resorting to numerical methods which might render the
-algorithm too slow to be practical.
+observed nucleotide frequencies in the training data.
 
@@
 The second of these encodes the assumption that distant nucleotides are
-effectively independent.
+effectively independent. We choose $p_{\text{max}} = 4$ and $d_{\text{max}} =
+10$, as reasonable default values.
 
-Figure \ref{fig:structure} shows an example of the structure learned when this
-procedure is applied to 200,000 reads from the Trapnell data set
-\cite{Trapnell2010}. The structure learned is non-trivial, and
-positions that uninformative are excluded automatically.
+Figure \ref{fig:models} shows examples of the structure learned when this
+procedure is applied to several data sets, using 200,000 reads from each.
 
@@
 \begin{enumerate}
-\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}) measured
-from reads within exons should not differ greatly from frequencies measured by
+\item Positional nucleotide frequencies (as in Figure \ref{fig:freqs}), measured
+from reads within exons, should not differ greatly from frequencies measured by
 sampling uniformly within the same exons.
-\item Read counts across exons should follow approximately a Poisson process.
+\item Read counts across a single exon should follow, approxmiately, a Poisson
+process.
 \end{enumerate}
 
-We applied our procedure (labeled ``BN'') as well as those of Li, et. al, and
-Hansen, et. al., which are implemented in the R packages \texttt{mseq} and
-\texttt{Genominator}, respectively. The procedures were applied to four publicly
-available data sets \cite{Bullard2010} \cite{Mortazavi2008} \cite{Trapnell2010}
-\cite{Wetterbom2010}, as well as an unpublished data set of our own (see, Table
-\ref{tab:datasets}).
+We applied our procedure (labeled ``BN'') as well as those as well as those of
+\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
+are implemented in the R packages \texttt{mseg} and \texttt{Genominator},
+respectively. Thhe procedures were applied to four publically available data
+sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
+an unpublished data sets of our own (Table \ref{tab:datasets}).
 
-%% TODO
-\begin{comment}
-\begin{tablehere}
-\begin{center}
-\footnotesize{
-\begin{tabular}{lllr}
-\textbf{Experiment} & \textbf{Species} & \textbf{Platform} & \textbf{Read Length} \\ \hline
-Wetterbom \cite{Wetterbom2010} & Chimpanzee & ABI & 33 \\
+\begin{table}
+\processtable{Datasets on which the methods are evaluated.\label{tab:datasets}}
+{
+\begin{tabular}{lllr}\toprule
+Experiment & Species & Platform & Read Length \\\midrule
+\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
 Katze (unpublished) & Macaque & ABI & 50 \\
-Bullard \cite{Bullard2010} & Human & Illumina & 35 \\
-Mortazavi \cite{Mortazavi2008} & Mouse & Illumina & 33 \\
-Trapnell \cite{Trapnell2010} & Mouse & Illumina & 75
+\citet{Bullard2010} & Human & Illumina & 35 \\
+\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
+\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
 \end{tabular}
-}
-\end{center}
-\caption{Data sets on which the methods are evaluated.}
-\label{tab:datasets}
-\end{tablehere}
-\end{comment}
+}{}
+\end{table}
 
@@
 from chromosomes 1--8 of the genome from which the reads were mapped (including
-chromosomes 2a and 2b in the Chimpanzee genome). For evaluation, we drew a set
-long, highly expressed exons from the remaining chromosomes. In particular, for
+chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
+of long, highly expressed exons from the remaining chromosomes. In particular, for
 each reference sequence, beginning with the set of exons annotated by Ensembl
@@
 
-Li recommends that their MART and GLM models be trained using the 100 most abundant
-genes. We used 1000 exons from chromosomes 1--8, otherwise chosen in a manner
-identical to that which was used to select the test exons.  Both the GLM and
-MART model were trained using the default parameters.
+\citet{Li2010} recommends that their MART and GLM models be trained using the
+100 most abundant genes. We used 1000 exons from chromosomes 1--8, otherwise
+chosen in a manner identical to that which was used to select the test exons.
+Both the GLM and MART models were trained using the default parameters.
 
-Hansen recommends using all of the reads to estimate heptamer frequencies used by
-their model. The training procedure works by simple tallying of frequencies and
-is thus relatively fast. We choose to use all of the reads from chromosomes 1--8.
-We evaluated three variations of this scheme. The first uses the default
-parameters, evaluating 7-mer frequencies averaged over the first two positions
-of the read (labeled ``Avg-7mer''). We also evaluated estimating 7-mer
-frequencies over only the first position, as well as 4-mer frequencies over only
-the first position. Though many variations of this scheme may be used, we found these
-additional two to be particularly informative. The background frequencies in
-each variation were were estimated from positions 18--23 within each read.
+\citet{Hansen2010} recommends using all of the reads to estimate heptamer
+frequencies used by their model. The training procedure works by simple tallying
+of frequencies. The implementation of this model in the Genominator package we
+found to use a great deal of memory, and were unable to train with the volume of
+data we wished, so we reimplemented the model and trained it on all of the reads
+aligned to chromosomes 1-8.
 
-All data sets were mapped using Bowtie \cite{Langmead2009} against the latest
-genome assemblies available from the UCSC Genome Browser \cite{Karolchik2008}
+We evaluated several variations of the heptamer model. The suggested method
+involved averaging the frequencies of the first two heptamers of each read. Yet,
+we found that in every case, this performed worse than simply counting the
+frequencies of the initial heptamer, and thus we report only the latter. The
+background frequences are estimated from positions 18--23 in each read.
+
+All data sets were mapped using Bowtie \citep{Langmead2009} against the latest
+genome assemblies available from the UCSC Genome Browser \citep{Karolchik2008}
 during preparation.
+
+%% leaving off here for now
+%% --------------------------------------------------------
 

### 2011-06-15 seqbias-paper.tex
more revisions

@@
 \centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
+%% TODO: caption
 \caption{Blah blah blah.}
@@
 
-%% leaving off here for now
-%% --------------------------------------------------------
-
 \subsection{Kullback-Leibler Divergence}
@@
 obvious bias. To quantify the non-uniformity observed in these plots, we use the
-symmetrized Kullback-Leibler (KL) divergence \cite{Kullback1951}. Comparing only
-nucleotide frequencies may miss dinucleotide or higher order bias. Thus, for a fair
-comparison, we compute the KL divergence for $k$-mers, with $k \in \{ 1 \dots 6
-\}$.
 
 If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
-frequency, the KL divergence is computed as,
-$$D( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
-Where the sum is over all $k$-mers. We compute the divergence at each
-position relative to the read start, by tallying (overlapping) $k$-mer
-frequencies at each position, weighted by the bias coefficient at that position,
-and comparing to a background distribution. The background distribution is
-estimated by averaging the $k$-mer frequencies at positions $-60 \dots -20$ and
-$20 \dots 60$, where the observed bias was very low in each data set.  We further
-take the precaution of ignoring any positions with more than 10,000 reads, to
-avoid the measure being dominated by any small group of outliers. The only
-qualitative effect this was seen to have was in the Katze data set, in which the
-frequency plots are otherwise quite skewed.
+frequency, the KL divergence is computed as, symmetrized Kullback-Leibler (KL)
+divergence \citep{Kullback1951}.
+%Comparing only nucleotide frequencies may miss
+%dinucleotide or higher order bias. Thus, for a fair comparison, we compute the
+%KL divergence for $k$-mers, with $k \in \{ 1 \dots 6 \}$.
+$$D_k( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
+where the sum is over all $k$-mers.
 
-The result of this this analysis is plotted in Figure
-\ref{fig:kl} for the Trapnell data set. Similar plots for the other data sets are
-provided in the supplement.
+%% --------------------
+
+To avoid the risk of a small number of reads with many duplicates dominating the
+measure, we do not count duplicate reads when computing the KL divergence. Since we
+explicitly chose exons with high coverage, not counting duplicates may not
+fully capturing the effects of the bias correction methods. Thus, for each exon,
+we compute the number of duplicates for each read, and count only reads in the
+upper half. Under the assumption of uniform sampling, the set of reads falling
+in the upper half should not depend on sequence, and we should expect the KL
+divergence to be low.
+
+We compute the divergence by reweighting the read counts using the predicted
+bias coefficient, choosing the upper half in each exon, discarding duplicate
+reads, and then tallying frequencies of overlapping $k$-mers. The $k$-mer
+distribution obtained are then compared to a background distribution obtained by
+redistributing reads uniformly at random within their exons.
+
+We repeated the procedure for $k \in \{1, 2, 3, 4, 5, 6\}$. The results of this
+analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
+remaining cases are plotted in Supplementary Section 3.
+%% TODO: what section, for real?
 
@@
 \centerline{\includegraphics[width=0.9\textwidth]{fig/kl.eps}}
-\caption{Plots of the KL divergence over overlapping $k$-mers in the Trapnell
-experiment.}
+\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
+(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
+frquencies expected under the assumption of unifom sampling from within exons.
+A large divergence indicates significant bias. Plotted here is the divergence
+from unadjusted reads counts as well as after adjusting reads counts using each
+method.}
 \label{fig:kl}
 \end{figure*}
-
-To summarize these plots, we summed the divergence from position -20
-through 20, for each of the methods, then normalized this number by dividing by
-the summed divergence before adjustment. The results of this analysis are plotted in
-Figure \ref{fig:kl-delta}.
-
-%% TODO
-\begin{comment}
-\begin{figurehere}
-\begin{center}
-\includegraphics[width=0.35\textwidth]{fig/kl-delta.pdf}
-\end{center}
-\caption{Summed KL divergence, normalized by unadjusted divergence, so that a
-number less than 1 indicates a decrease in divergence, or an increase in
-uniformity.}
-\label{fig:kl-delta}
-\end{figurehere}
-\end{comment}
 
@@
 In this comparison, we measure how well the counts conform to a Poisson process.
-The assumption of positional read counts following a Poisson distribution has
-been shown to be a poor fit \cite{Srivastava2010}, yet measuring the improvement in
-fit derived from correcting for technical bias remains a principled and easily
-interpreted criteria.
+The assumption of positional read counts following a Poisson distribution is
+known to be a poor fit \citep{Srivastava2010}, but measuring the improvement in
+the fit derived from correcting for bias remains a principled and easily
+interpreted criterion.
 
@@
 Poisson rate is fixed for each exon. That is, for position $j$ within
-gene $i$, the rate is,
+exon $i$, the rate is
 $$ \lambda_{ij} = a_i $$
-Where $a_i$ is the parameter being fit. For comparison, we then fit a model in
-which the rate is proportional to the predicted bias coefficients.
+where $a_i$ is the parameter being fit. For comparison, we then fit a model in
+which the rate is also proportional to the predicted bias coefficients:
 $$ \lambda'_{ij} = a_i b_{ij} $$
 
-If the null model has log-likelihood $L$ and the bias corrected model
-$L'$, we measure the change in fit as,
-$$ \log \left( \frac{ L_{i} } { L'_{i} } \right) $$
-When the fit improves, this number is positive, and negative when it is
-worsened. The values computed from our test are listed in Table \ref{tab:pois}.
+If the null model has log-likelihood $L$, and the bias-corrected model $L'$, a
+simple goodness of fit measure is the improvement in log-likelihood (a statistic
+commonly known as McFadden's pesudo-coefficient of determination
+\citep{McFadden1974}), defined as,
+$$R^2 = 1 - \frac{L'}{L}$$
 
-\begin{comment}
-\begin{tablehere}
-\begin{center}
-\tiny{
-\begin{tabular}{lrrrrrr}
-         &\textbf{Avg-7mer}&\textbf{7mer}&\textbf{4mer}&\textbf{GLM}&\textbf{MART}&\textbf{BN} \\
-\textbf{Katze}            &  0.004  & 0.017  & 0.055 &0.166  &0.248&0.305 \\
-\textbf{Wetterbom}        & -0.076  &-0.058  &-0.002 &0.240  &0.070&0.557 \\
-\textbf{Bullard}          &  0.124  & 0.192  & 0.139 &0.285  &0.198&0.380 \\
-\textbf{Mortazavi}        &  0.059  & 0.123  & 0.099 &0.284  &0.310&0.401 \\
-\textbf{Trapnell}         &  0.094  & 0.163  & 0.104 &0.265  &0.332&0.349 \\
-\end{tabular}
-}
-\end{center}
-\caption{Change in fit, compared to the null model of uniform Poisson rate across exons.}
-\label{tab:pois}
-\end{tablehere}
-\end{comment}
+This measure can be interpreted as the improvement in fit over the null model,
+with $R^2 = 1$ indicating a prefect fit, and a negitve numbers, an increasingly
+worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
+proportion of variability eplained by the model.
 
-Computing one number to summarize the fit of the model ignores variance in the
-likelihood from exon to exon. This number could be dominated by a few test
-regions. Figure \ref{fig:pois} plots the distribution of the change in fit
-across the 1000 test exons, revealing a similar, but more nuanced result.
+We compue $R^2$ for each of the test exons, giving us a sense of the variability
+of the effectiveness of each model. The results of this analysis are plotted in
+Figure \ref{fig:pois}.  To summarize each model with a single number, we can
+examine the median $R^2$ value, as listed in Table \ref{tab:pois}.
 
-%% TODO
 \begin{figure}
 \centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
-\caption{Plotted is the change in fit of each of the test exons. A
-positive value indicates a better fit in the bias corrected model.}
+\caption{For each of the 1000 test exons, we measure a pseudo-coefficients of
+determination $R^2$, equivalent to the improvement in log-likelihood under the
+bias corrected model. The statistic is positive when uniformity is increased,
+and negative when decreased. Boxes are plotted to mark the 25\%, 50\%, and 75\%
+quantiles, with whiskers extending to 1.5 times the inter-quartile range (i.e.,
+the span between tho 25\% and 75\% quantiles).
+}
+
     \label{fig:pois}
 \end{figure}
+
+\begin{table}
+\processtable{
+    The $R^2$ goodness of fit statistic measures increased uniformity in read
+    coverage, after correcting for bias. Here the median $R^2$ across the test
+    exons is listed for each method and simple. A positive $R^2$ indicates a
+    better fit.\label{tab:pois}}{
+    \begin{tabular}{lcccc}\toprule
+          & BN    & MART  & GLM   & 7mer \\\midrule
+Wetterbom & 0.168 & 0.016 & 0.066 & -0.079 \\
+Katze     & 0.283 & 0.243 & 0.158 & 0.033 \\
+Bullard   & 0.271 & 0.163 & 0.224 & 0.157 \\
+Mortazavi & 0.225 & 0.210 & 0.197 & 0.091 \\
+Trapnell  & 0.298 & 0.289 & 0.248 & 0.138
+    \end{tabular}
+}{}
+\end{table}
 
@@
 
-In our results, estimating initial 4-mer or 7-mer frequencies was not seen to be
-as effective as the other models, even when data generated using random hexamer
-priming was used. Given the large number of parameters needed to estimate 7-mer
-frequencies, a likely explanation is that the model is overfit to the training
-set. Though the training set consisted of at least 600,000 reads in
+In our results, estimating initial heptamer frequencies was not seen to be as
+effective as the other models, even when data generated using random hexamer
+priming was used. Given the large number of parameters needed to estimate
+heptamer frequencies, a likely explanation is that the model is overfit to the
+training set. Though the training set consisted of at least 600,000 reads in
 each experiment, it seems more reads, or a greater diversity of reads are needed
-to produce an accurate estimate.
+to produce an accurate or useful estimate.
 
@@
 using an estimation of sequence probability that is more robust to overfitting
-and can account for bias beyond the initial heptamer. This approach is
-competitive with that of Li, et. al., despite not requiring gene annotations.
+and can account for bias beyond the initial heptamer. In all our tests, this
+approach was at least as effective as those of \citet{Li2010}, despite not
+requiring gene annotations, or manual selection of training examples.
+
+We have withheld any direct comparison to the method described by
+\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
+this method is superficially similar to our own, a proper comparison is
+difficult, as the software can not be applied independently of estimating
+transcript abundance in FPKM using Cufflinks. Fairness would dictate that
+competing methods be substituted in FPKM estimation, or that a seperate
+interface be written to the Cufflinks bias correction method---both comparison
+requiring significant effort.
+
+Though the Cufflinks method and our own both use graphical models to estimate
+sequence probabilities, we make no restriction on the graph other than
+acyclicity, and go to considerable effort to efficiently approximate the optimal
+structure for each data set, rathor than using a fixed structure, as in
+Cufflinks. A ``one size fits all'' approach likely works quite well in many
+cases, yet the observed specificity of the bias to protocal and platform argues
+against it. For example, the structures learned by our method (Figure
+\ref{fig:models}), are considerably different between those sequenced on an
+Illumina platform versus an AB platform.
+
+%% TODO: discussion of run-time, etc
 
 Because we do not require annotations, CHiP-Seq, and other high-throughput
-sequencing experiments, may also benefit from our model. In a preliminary
+sequencing experiments may also benefit from our model. In a preliminary
 investigation, we found the sequence bias in one CHiP-Seq experiment
@@
 evaluated, however our method is still able effectively correct for the bias
-that was observed (see, supplementary). Protocol differences, as we have seen,
+that was observed (see, Supplementary Section 4).
+%% TODO: which actual section ?
+Protocol differences, as we have seen,
 can result in significant differences in observed nucleotide frequencies, so we
@@
 
-Our method can also very naturally capture dependency between nucleotides in the
-sequence. This of particular importance when working with reads generated from
-the AB SOLiD platform, in which the di-base, ``colorspace'' coding scheme is
-used, encoding two adjacent nucleotides as a single color. A color bias at one
-position will manifest itself as a nucleotide bias at two positions. Capturing
-this effect requires a model that considers the dependency between nearby
-positions. In our results, the MART model, as well as our own, perform
-significantly better than the GLM model, illustrating this proposition.
-
 RNA-Seq is most often used to compare levels of expression, and so a natural
@@
 the bias appears to be largely, but not entirely consistent (see,
-supplementary). Similarly, in Figure \ref{fig:freqs}, the three data sets
+Supplementary Section 5).
+%% TODO: which actual section ?
+Similarly, in Figure \ref{fig:freqs}, the three data sets
 sequenced on the Illumina platform display similar patterns of nonuniformity,

### 2011-06-15 results/plots/plot-pois.R

@@ -75,3 +75,12 @@ for (w in levels(xs$sample)) {
     dev.off()
 }
 
+
+
+# TODO: compute median values.
+
+xs$r2 <- r2(xs$L, xs$L.base)
+med <- tapply(xs$r2, INDEX = list(xs$sample, xs$method), FUN = median)
+print(round(med, digits=3))
+
+

### 2011-06-16 seqbias-paper.tex
new data and plots

@@
 
+Our own method was trained on the 200,000 reads from chromosomes 1--8, with the
+highest number of duplicates.
+
 All data sets were mapped using Bowtie \citep{Langmead2009} against the latest
@@
           & BN    & MART  & GLM   & 7mer \\\midrule
-Wetterbom & 0.168 & 0.016 & 0.066 & -0.079 \\
-Katze     & 0.283 & 0.243 & 0.158 & 0.033 \\
-Bullard   & 0.271 & 0.163 & 0.224 & 0.157 \\
-Mortazavi & 0.225 & 0.210 & 0.197 & 0.091 \\
-Trapnell  & 0.298 & 0.289 & 0.248 & 0.138
+Wetterbom & 0.173 & 0.016 & 0.066 & -0.079 \\
+Katze     & 0.286 & 0.243 & 0.158 & 0.033 \\
+Bullard   & 0.276 & 0.163 & 0.224 & 0.157 \\
+Mortazavi & 0.248 & 0.210 & 0.197 & 0.091 \\
+Trapnell  & 0.308 & 0.289 & 0.248 & 0.138
     \end{tabular}

### 2011-06-16 results/plots/config.R

@@ -1,4 +1,5 @@
 
 
-d <- '../20110610/'
+d <- '../20110615/'
+d_old <- '../20110610/'
 

### 2011-06-16 results/plots/plot-dg.py

@@ -99,8 +99,10 @@ def draw_edges(G, ctx):
         d = abs(node_pos[u] - node_pos[v])
 
         # totally ad-hoc formlae, here:
-        delta = r * (1.0 - (8.0 * (d) / n))
-        w = ed((xu, yu), (xv, yv)) / 2.0
+        delta = r * (1.0 - (7.0 * (d) / n))
+
+        #w = ed((xu, yu), (xv, yv)) / 2.0
+        w =  5.0 / d
 
         omega = atan2(w, delta)
         xa, ya = delta * sin(theta + omega), delta * cos(theta + omega)

### 2011-06-16 results/plots/plot-kl.R

@@ -9,7 +9,10 @@ ws <- c('bullard.tab', 'katze.tab', 'mortazavi.tab',
 
 xs <- NULL
 for (w in ws) {
-    x  <- read.table(paste(d, '/kl/', w, sep = ''), header = T)
+    x  <- read.table(paste(d_old, '/kl/', w, sep = ''), header = T)
+    x  <- subset(x, method != 'Jones')
+    x  <- rbind(x, read.table(paste(d, '/kl/', w, sep = ''), header = T))
+
     xs <- rbind(xs, x)
 }
 

### 2011-06-16 results/plots/plot-models.sh

@@ -2,7 +2,7 @@
 
 
 
-for f in `ls ../20110610/dot/*.dot`
+for f in `ls ../20110615/dot/*.dot`
 do
     g="model."`basename $f | sed s/dot/svg/`
     echo "$f -> $g"

### 2011-06-16 results/plots/plot-pois.R

@@ -7,9 +7,13 @@ source('config.R')
 ws <- c('bullard.tab', 'katze.tab', 'mortazavi.tab',
         'trapnell.tab', 'wetterbom.tab')
 
+
 xs <- NULL
 for (w in ws) {
-    x  <- read.table(paste(d, '/pois/', w, sep = ''), header = T)
+    x  <- read.table(paste(d_old, '/pois/', w, sep = ''), header = T)
+    x <- subset(x, method != 'Jones')
+    x <- rbind(x, read.table(paste(d, '/pois/', w, sep = ''), header = T))
+
     xs <- rbind(xs, x)
 }
 

### 2011-06-17 seqbias-paper.tex
all new data, plots

@@
 \centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
-%% TODO: caption
-\caption{Blah blah blah.}
+\caption{
+The network structures learned on each of the data sets are examined. Positions
+are relative to the read start, which is labeled 0. Hollow circles indicate
+position that were not included in the model, being deemed uninformative, give
+the other positions and edges. The number of parameters of parameters needed to
+specify each model is listed in parenthesis below.}
 \label{fig:models}
@@
           & BN    & MART  & GLM   & 7mer \\\midrule
-Wetterbom & 0.173 & 0.016 & 0.066 & -0.079 \\
-Katze     & 0.286 & 0.243 & 0.158 & 0.033 \\
-Bullard   & 0.276 & 0.163 & 0.224 & 0.157 \\
-Mortazavi & 0.248 & 0.210 & 0.197 & 0.091 \\
-Trapnell  & 0.308 & 0.289 & 0.248 & 0.138
+Wetterbom & 0.174 & 0.016 & 0.066 & -0.079 \\
+Katze     & 0.280 & 0.243 & 0.158 & 0.033 \\
+Bullard   & 0.267 & 0.163 & 0.224 & 0.157 \\
+Mortazavi & 0.240 & 0.210 & 0.197 & 0.091 \\
+Trapnell  & 0.289 & 0.289 & 0.248 & 0.138
     \end{tabular}
@@
 that should be evaluated and corrected for.
+%% TODO: better end sentence
 

### 2011-06-17 results/plots/config.R

@@ -1,5 +1,5 @@
 
 
-d <- '../20110615/'
+d <- '../20110616/'
 d_old <- '../20110610/'
 

### 2011-06-17 results/plots/plot-kl.R

@@ -48,8 +48,8 @@ CairoSVG('kl.svg', width = 6 * scl, height = 1.8 * scl)
 
 xs$k <- factor(xs$k)
 
-p <- qplot(data = xs, x = pos, y = div, color = method,
-           geom = 'line')
+p <- qplot(data = xs, x = pos, y = div, color = method, geom = 'line')
+p <- p + geom_line(size = 0.7)
 p <- p + facet_grid(k ~ sample, scale = 'free')
 
 

### 2011-06-17 results/plots/plot-models.sh

@@ -2,7 +2,7 @@
 
 
 
-for f in `ls ../20110615/dot/*.dot`
+for f in `ls ../20110616/dot/*.dot`
 do
     g="model."`basename $f | sed s/dot/svg/`
     echo "$f -> $g"

### 2011-06-17 seqbias-paper.tex
paper: finished

@@
 Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
-often conflated with protocal-specific bias, influenced by PCR amplification,
-and differing primer affinities and mixtures, for example. The result is
-potentially inaccurate quantification, affecting de novo gene annotation and
-isoform quantification in RNA-Seq, and peak calling in ChiP-Seq.
+often conflated with protocol-specific sequence bias, influenced by PCR
+amplification, and differing primer affinities and mixtures, for example.  The
+result is decreased accuracy in many applications, such as de novo gene
+annotation and isoform quantification in RNA-Seq, and peak calling in
+ChIP-Seq.
 
@@
 applicable to any high throughput sequencing data. We evaluate our method on
-several datasets, and by multiple criteria, demonstrating that it effectively
+several data sets, and by multiple criteria, demonstrating that it effectively
 decreases bias and increases uniformity.
@@
 The method is implemented in the \texttt{seqbias} R/Bioconductor package,
-available freely under the LGPL licence from
+available freely under the LGPL license from
 \href{http://bioconductor.org}{http://bioconductor.org}.
@@
 differentially expression, since more reads are aligned to longer transcripts,
-granting higher statistical significance in tests differential expression.
+granting higher statistical significance in tests of differential expression.
 Recently, some methods have been developed attempting to account for this effect
-in gene set enrichment tests \citep{Young2010,Gao2011}.
-
-%% TODO: do something with this sentance.
-Additionally,
-differing transcript lengths introduce ``edge effects'', as very short fragments
-from the end of a transcript will not be sequenced. NEUMA \cite{Lee2010} and
-FPKM \cite{Trapnell2010} are two methods that attempt to take this into account.
+in gene set enrichment tests \citep{Young2010,Gao2011}, leading to more accurate
+downstream analysis.
 
@@
 \caption{Nucleotide frequencies are plotted relative to the start (labeled,
-position 0) of each mapped read, respecting strand. The sequence is taken frmo
+position 0) of each mapped read, respecting strand. The sequence is taken from
 the genomic context surrounding the read, so that -40 to -1, for example, fall
-outside the read sequence itself. The symmetrized Killback-Leibler divergence is
+outside the read sequence itself. The symmetrized Kullback-Leibler divergence is
 used to summarize the difference in nucleotide frequency compared to a fixed
-estimate of background necleotide frequencies made by sampling many positions
+estimate of background nucleotide frequencies made by sampling many positions
 nearby mapped reads.  Under the assumption that reads are sampled uniformly from
@@
 Markov chains. The structure of these Markov chains are hard-coded, chosen in
-advance using a hill-climbing algorithm on a representative dataset. This method
+advance using a hill-climbing algorithm on a representative data set. This method
 is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
@@
 reads are aligned to a reference sequence.
-
-%% TODO: where should this figure be cited?
-\begin{figure}
-\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
-\caption{An overview of the approach taken: (a) foreground sequences are sampled
-from the regions surrounding the starts of mapped reads, (b) background
-sequences are sampled by randomly offsetting foreground positions, (c) a
-Bayesian network is classifier is trained to descriminate between the set of
-sampled foreground and background sequences, (d) and the classifier is evaluated
-at each position within a locus, predicting bias.}
-\label{fig:overview}
-\end{figure}
 
@@
 The Bayesian network can then used to evaluate sequence probability, and thus
-bias, at any genomic position.
+bias, at any genomic position. Figure \ref{fig:overview} gives a brief overview
+of the proposed model.
+
+\begin{figure}
+\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
+\caption{An overview of the approach taken: (a) foreground sequences are sampled
+from the regions surrounding the starts of mapped reads, (b) background
+sequences are sampled by randomly offsetting foreground positions, (c) a
+Bayesian network is classifier is trained to discriminate between the set of
+sampled foreground and background sequences, (d) and the classifier is evaluated
+at each position within a locus, predicting bias.}
+\label{fig:overview}
+\end{figure}
 
@@
 are sampled uniformly from the genome we run the risk of incorrectly adjusting
-for biological sequence bias, rather than techncical sequence bias.  To avoid
+for biological sequence bias, rather than technical sequence bias.  To avoid
 this, we propose using paired training data. Each foreground training
@@
 estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
-notation) directly, given training data consistig of long, highly-expressed
+notation) directly, given training data consisting of long, highly-expressed
 transcripts.
@@
 which the foreground sequences were sampled.  The offset is drawn from a
-zero-mean gaussian, and rounded to the nearest integer, away from zero.
+zero-mean Gaussian, and rounded to the nearest integer, away from zero.
 By using such a scheme, we attempt to mitigate the effects of biological
@@
 
-This procuder produces a training set of $n$ sequences with accompanying labels
+This procedure produces a training set of $n$ sequences with accompanying labels
 $T = \{ (s_1, x_1), (s_2, x_2), \dots, (s_n, x_n) \}$. The label $x_i$ is
@@
 set of samples, simply maximizing the likelihood would severely overfit the
-training set. We thus penalize model complexity heuristically using the Akaike
-information criterion. Where $m$ is the number of parameters needed to specify
-the model, we maximize,
+training set. We thus penalize model complexity heuristically using the Bayesian
+Information Criterion \citep{Schwarz1978a}. Where $m$ is the number of
+parameters needed to specify the model, we maximize,
 
-$$ \ell' = \ell - m $$
+$$ \ell' = 2 \ell - m \log n $$
 
-Some benifit would be obtained from a more highly tuned complexity penalty.
+Some benefit would be obtained from a more highly tuned complexity penalty.
 However, since the model is trained greedily, additional parameters will be
-decreasingly informitive, and increasingly similar between foreground and
+decreasingly informative, and increasingly similar between foreground and
 background. Adding more parameters will have little effect.  Only when $m$ is
 allowed to grow exponentially does the prediction become polluted by small
-deviations between thousands of uninformitive parameters.
+deviations between thousands of uninformative parameters.
 
@@
 Figure \ref{fig:models} shows examples of the structure learned when this
-procedure is applied to several data sets, using 200,000 reads from each.
+procedure is applied to several data sets, using 100,000 reads from each.
 
@@
 sampling uniformly within the same exons.
-\item Read counts across a single exon should follow, approxmiately, a Poisson
+\item Read counts across a single exon should follow, approximately, a Poisson
 process.
@@
 \citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
-are implemented in the R packages \texttt{mseg} and \texttt{Genominator},
-respectively. Thhe procedures were applied to four publically available data
+are implemented in the R packages \texttt{mseq} and \texttt{Genominator},
+respectively. The procedures were applied to four publicly available data
 sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
@@
 \begin{table}
-\processtable{Datasets on which the methods are evaluated.\label{tab:datasets}}
+\processtable{Data sets on which the methods are evaluated.\label{tab:datasets}}
 {
@@
 frequencies of the initial heptamer, and thus we report only the latter. The
-background frequences are estimated from positions 18--23 in each read.
+background frequencies are estimated from positions 18--23 in each read.
 
-Our own method was trained on the 200,000 reads from chromosomes 1--8, with the
+Our own method was trained on the 100,000 reads from chromosomes 1--8, with the
 highest number of duplicates.
@@
 where the sum is over all $k$-mers.
-
-%% --------------------
 
@@
 \begin{figure*}
-\centerline{\includegraphics[width=0.9\textwidth]{fig/kl.eps}}
+\centerline{\includegraphics[width=1.00\textwidth]{fig/kl.eps}}
 \caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
 (here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
-frquencies expected under the assumption of unifom sampling from within exons.
+frequencies expected under the assumption of uniform sampling from within exons.
 A large divergence indicates significant bias. Plotted here is the divergence
@@
 simple goodness of fit measure is the improvement in log-likelihood (a statistic
-commonly known as McFadden's pesudo-coefficient of determination
+commonly known as McFadden's pseudo-coefficient of determination
 \citep{McFadden1974}), defined as,
@@
 This measure can be interpreted as the improvement in fit over the null model,
-with $R^2 = 1$ indicating a prefect fit, and a negitve numbers, an increasingly
+with $R^2 = 1$ indicating a prefect fit, and a negative numbers, an increasingly
 worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
-proportion of variability eplained by the model.
+proportion of variability enplaned by the model.
 
-We compue $R^2$ for each of the test exons, giving us a sense of the variability
+We compute $R^2$ for each of the test exons, giving us a sense of the variability
 of the effectiveness of each model. The results of this analysis are plotted in
@@
 difficult, as the software can not be applied independently of estimating
-transcript abundance in FPKM using Cufflinks. Fairness would dictate that
-competing methods be substituted in FPKM estimation, or that a seperate
-interface be written to the Cufflinks bias correction method---both comparison
-requiring significant effort.
+transcript abundance in FPKM (fragments per exonic kilobase, per million mapped
+reads) using Cufflinks. Fairness would dictate that competing methods be
+substituted in FPKM estimation, or that a separate interface be written to the
+Cufflinks bias correction method---both comparison requiring significant effort.
 
@@
 acyclicity, and go to considerable effort to efficiently approximate the optimal
-structure for each data set, rathor than using a fixed structure, as in
+structure for each data set, rather than using a fixed structure, as in
 Cufflinks. A ``one size fits all'' approach likely works quite well in many
-cases, yet the observed specificity of the bias to protocal and platform argues
+cases, yet the observed specificity of the bias to protocol and platform argues
 against it. For example, the structures learned by our method (Figure
@@
 
-%% TODO: discussion of run-time, etc
+Training our model on more reads leads to more accurate estimation of bias, but
+increasingly long training time. For example, in our tests, fitting our model to
+100,000 reads from the Mortazavi data, training time was approximately 45
+minutes, running on one core of a 3Ghz Intel Xeon processor. However, limiting
+the training to 25,000 reads leads to a model that is nearly as accurate while
+requiring under 4 minutes to train. A full discussion of this trade-off is in
+Supplementary Section 6.
+%% TODO: what actual section?
 
-Because we do not require annotations, CHiP-Seq, and other high-throughput
+Because we do not require annotations, ChIP-Seq, and other high-throughput
 sequencing experiments may also benefit from our model. In a preliminary
-investigation, we found the sequence bias in one CHiP-Seq experiment
+investigation, we found the sequence bias in one ChIP-Seq experiment
 \cite{Cao2010} was less than that observed in any of the RNA-Seq data we
@@
 can result in significant differences in observed nucleotide frequencies, so we
-may not safely assert that bias in CHiP-Seq data is always low.  Given the weak
+may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
 assumptions made by our model, our estimation of bias could easily be
-incorporated into CHiP-Seq peak calling algorithms, and potentially improve
+incorporated into ChIP-Seq peak calling algorithms, and potentially improve
 accuracy.
@@
 %% TODO: which actual section ?
-Similarly, in Figure \ref{fig:freqs}, the three data sets
-sequenced on the Illumina platform display similar patterns of nonuniformity,
-yet differ in magnitude. Batch effects in RNA-Seq remain a legitimate concern,
-that should be evaluated and corrected for.
-%% TODO: better end sentence
+Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced on the
+Illumina platform display similar patterns of nonuniformity, yet differ in
+magnitude, suggesting that batch effects in RNA-Seq remain a legitimate concern
+that should not be dismissed without evaluation.
+
+In summary, we have demonstrated a relatively simple graphical model that
+effectively corrects for sequence bias pervasive in RNA-Seq, and to a lesser
+extent, ChIP-Seq experiments. In our tests, this model performs at least as
+well, and often better than existing methods, and involves fewer requirements or
+assumptions. Our model leads to more accurate quantification, and would likely
+provide a positive benefit when incorporated into downstream analysis.
 

### 2011-06-17 seqbias-paper.tex
supplement: finished

@@
 
-\title{Correcting for bias in high throughput sequencing data}
+\title{Correcting for bias in high-throughput sequencing data}
 \author[Jones \textit{et~al}]
@@
 analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
-remaining cases are plotted in Supplementary Section 3.
-%% TODO: what section, for real?
+remaining cases are plotted in Supplementary Section 4.
 
@@
 requiring under 4 minutes to train. A full discussion of this trade-off is in
-Supplementary Section 6.
-%% TODO: what actual section?
+Supplementary Section 5.
 
@@
 evaluated, however our method is still able effectively correct for the bias
-that was observed (see, Supplementary Section 4).
-%% TODO: which actual section ?
+that was observed (see, Supplementary Section 2).
 Protocol differences, as we have seen,
@@
 concern is the consistency of the bias between samples. In the data we examined,
-the bias appears to be largely, but not entirely consistent (see,
-Supplementary Section 5).
-%% TODO: which actual section ?
-Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced on the
-Illumina platform display similar patterns of nonuniformity, yet differ in
-magnitude, suggesting that batch effects in RNA-Seq remain a legitimate concern
-that should not be dismissed without evaluation.
+the bias appears to be largely, but not entirely consistent (see, Supplementary
+Section 3).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
+on the Illumina platform display similar patterns of nonuniformity, yet differ
+in magnitude, suggesting that batch effects in RNA-Seq remain a legitimate
+concern that should not be dismissed without evaluation.
 

### 2011-06-17 seqbias-supplement.tex
compiling in bioinformatics template
supplement: finished

@@
 \documentclass[letterpaper]{article}
-\usepackage{qcjstd}
 \usepackage{lscape}
@@
 \usepackage{graphicx}
-%\usepackage{palatcm}
+\usepackage{fontspec}
 
-\geometry{left=1in,right=1in, top=1in, bottom=1in}
+%\setmainfont{Helvetica}
+
+\geometry{left=1.5in,right=1.5in, top=1in, bottom=1in}
+
+\title{SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data}
+\author{Daniel C. Jones}
 
@@
 
-\qcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}
+\maketitle
+
+%\qcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}
 
@@
 
-%\begin{figure}[b]
 \begin{center}
@@
 \end{center}
-%\caption{Nucleotide frequencies for trimmed and untrimmed reads from the
-%Mortazavi dataset.}
-%\label{fig:trimmed}
-%\end{figure}
 
@@
 
-%\begin{figure}[b]
 \begin{center}
-\includegraphics[width=0.5\textwidth]{fig/freqs_cao.pdf}
+\includegraphics[width=0.7\textwidth]{fig/freqs_cao.pdf}
 \end{center}
-%\caption{Nucleotide frequencies from CHiP-Seq reads.}
-%\label{fig:freqs_cao}
-%\end{figure}
 
@@
 
-We did however repeat our analysis using the Kullback-Leibler divergence.  We
-trained each method using reads from reads from chromosomes 1--8. The remaining
-chromosomes were segmented into 500nt bins, and the KL divergence was sampled
-from the 50,000 segments with the highest read counts.
+We did however repeat our analysis using the Kullback-Leibler divergence.
+We used several variations of the method described by Hanset, et. al.,
+\cite{Hansen2010}. ``7mer'' estimated the initial heptamer, ``Avg-7mer''
+averages the intial two heptamers, and ``4mer'' averages the initial 4mer.
 
-First we plot directly the positional KL divergence, computed identically
+We trained each method using reads from reads from chromosomes 1--8. The
+remaining chromosomes were segmented into 500nt bins, and the KL divergence was
+sampled from the 50,000 segments with the highest read counts.
+
+Below, we plot directly the positional KL divergence, computed identically
 
@@
 \includegraphics[width=0.5\textwidth]{fig/cao-kl.pdf}
-\end{center}
-
-Secondly, we summarize these plots, computing one normalized divergence
-estimate by summing the KL divergence of positions -20 through 20, and
-divided by the unadjusted summed divergence to normalize.
-
-\begin{center}
-\includegraphics[width=0.5\textwidth]{fig/cao-kl-delta.pdf}
 \end{center}
@@
 
-\section{Additional Plots}
+\section{Additional Kullback-Leibler Divergence Plots}
 
-Below we plot the adjusted and unadjusted KL divergence of each sample.
+Below we plot the adjusted and unadjusted KL divergence plots, supplementing
+those in the results section of the main paper.
 
-%\begin{figure}[b]
-%\begin{center}
 \includegraphics[width=\textwidth]{fig/kl-all.pdf}
-%\end{center}
-%\caption{Positional KL divergence for each dataset and method.}
-%\label{fig:kl-all}
-%\end{figure}
+
+\section{Runtime-Accuracy Tradeoff}
+
+Training our model on more reads results in more accurate estimation of bias,
+but a longer training time. Here we investigate that trade-off by training our
+model on progressively more reads from the Mortazavi data set. We use the
+median pseuo-coefficient of determination ($R^2$) over exons selected for our
+test set. Each point is additionally labeled with the time used to train the
+model on one core of a 3Ghz Inteal Xeon processor.
+
+\includegraphics[width=\textwidth]{fig/pois-n.eps}
 
 \bibliographystyle{plain}
-\bibliography{seqbias-supplement}
+\bibliography{seqbias-paper}
 

### 2011-06-21 seqbias-paper.tex
added funding note, corrections in the supplement

@@
 \address{
-$^{1}$Deportment of Computer Science and Engineering, University of
+$^{1}$Department of Computer Science and Engineering, University of
 Washington, Seattle, WA 98195-2350, USA\\
@@
 Experiment & Species & Platform & Read Length \\\midrule
-\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
-Katze (unpublished) & Macaque & ABI & 50 \\
+\citet{Wetterbom2010} & Chimpanzee & AB & 33 \\
+Katze (unpublished) & Macaque & AB & 50 \\
 \citet{Bullard2010} & Human & Illumina & 35 \\
@@
 Training our model on more reads leads to more accurate estimation of bias, but
-increasingly long training time. For example, in our tests, fitting our model to
+an increasingly long training time. For example, in our tests, fitting our model to
 100,000 reads from the Mortazavi data, training time was approximately 45
@@
 
+%\section*{Acknowledgement}
+%Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text text text text
+
+\paragraph{Funding\textcolon} This project has been funded in whole or in part
+with Federal funds from the National Institute of Allergy and Infectious
+Diseases, National Institutes of Health, Department of Health and Human Services,
+under Contract No.  HHSN272200800060C and by the Public Health Service grant
+P51RR000166 from the National Institutes of Health.
+
 \bibliographystyle{natbib}

### 2011-06-21 seqbias-supplement.tex
added funding note, corrections in the supplement

@@
 \documentclass[letterpaper]{article}
+\usepackage{dcj}
 \usepackage{lscape}
@@
 
-\maketitle
+%\maketitle
 
-%\qcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}
+\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}
 
@@
 
-\section{CHiP-Seq}
+\section{ChIP-Seq}
 
-Though we found the MART model \cite{Li2010} to be competitive with our own,
-yet ours offers the advantage that no gene annotations are needed for training. In
+Though the MART model \cite{Li2010} performed very well in several cases, ours
+offers the advantage that no gene annotations are required for training.  In
 RNA-Seq this is useful in applications of de-novo gene discovery, but it also
-allows our method to be applied to CHiP-Seq and another short read data. Here we
-examine one publicly available CHiP-Seq data set from Cao, et. al.
-\cite{Cao2010}.
+allows our method to be applied to ChIP-Seq and other short read data. Here we
+examine one publicly available ChIP-Seq data set from Cao, et. al.
+\cite{Cao2010}. Specifically, we used one run, with Sequence Read Archive
+accession number SRR034831, containing 3,873,192 reads. These were mapped to the
+reference genome using Bowtie \cite{Langmead2009}, resulting in 1,382,867
+uniquely mapped reads, which were used for this analysis.
 
@@
 transcription across annotated exons to evaluate the efficacy of the models, as
-well as to train the GLM and MART models. Evaluating bias correction on CHiP-Seq
+well as to train the GLM and MART models. Evaluating bias correction on ChIP-Seq
 data necessitates dropping the Poisson regression test, as well as the GLM and
@@
 We did however repeat our analysis using the Kullback-Leibler divergence.
-We used several variations of the method described by Hanset, et. al.,
+We used several variations of the method described by Hansen, et. al.,
 \cite{Hansen2010}. ``7mer'' estimated the initial heptamer, ``Avg-7mer''
-averages the intial two heptamers, and ``4mer'' averages the initial 4mer.
+averages the initial two heptamers, and ``4mer'' averages the initial 4mer.
 
-We trained each method using reads from reads from chromosomes 1--8. The
+We trained each method using reads from chromosomes 1--8. The
 remaining chromosomes were segmented into 500nt bins, and the KL divergence was
@@
 
-Below, we plot directly the positional KL divergence, computed identically
+Below, we plot directly the positional KL divergence, computed in tho same
+manner described in the Results section of the paper.
 
@@
 
-We see that our method is effective at reducing the bias in this ChiP-Seq data.
+We see that our method is effective at reducing the bias in this ChIP-Seq data.
 
@@
 
-\section{Runtime-Accuracy Tradeoff}
+\section{Runtime-Accuracy Trade-off}
 
@@
 model on progressively more reads from the Mortazavi data set. We use the
-median pseuo-coefficient of determination ($R^2$) over exons selected for our
+median pseudo-coefficient of determination ($R^2$) over exons selected for our
 test set. Each point is additionally labeled with the time used to train the
-model on one core of a 3Ghz Inteal Xeon processor.
+model on one core of a 3Ghz Intel Xeon processor.
 

### 2011-06-21 seqbias-paper.tex
various corrections from markus korth

@@
 Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
-often conflated with protocol-specific sequence bias, influenced by PCR
-amplification, and differing primer affinities and mixtures, for example.  The
+often conflated by protocol-specific sequence bias, influenced by PCR
+amplification, or differing primer affinities and mixtures, for example.  The
 result is decreased accuracy in many applications, such as de novo gene
@@
 noted by \citet{Oshlack2009} transcript length biases discovery of
-differentially expression, since more reads are aligned to longer transcripts,
+differential expression, since more reads are aligned to longer transcripts,
 granting higher statistical significance in tests of differential expression.
@@
 
-This bias, though observed primarily in the 5' end of a read, is not resolved by
+This bias, though observed primarily in the $5'$ end of a read, is not resolved by
 trimming the reads prior to mapping \citep{Hansen2010} (see also, Supplementary
@@
 
-Li, et. al., \cite{Li2010} propose two models. The first is a Poisson linear
+\citet{Li2010} propose two models. The first is a Poisson linear
 model, in which read counts across a transcript follow inhomogeneous Poisson
@@
 from its position. Estimating bias is therefore a problem of finding a model of
-sequence probability that sufficiently complex to capture the common features of
-the training data yet avoids overfitting.
+sequence probability that is sufficiently complex to capture the common features
+of the training data yet avoids overfitting.
 
@@
 estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
-notation) directly, given training data consisting of long, highly-expressed
+notation) directly, given training data consisting of long, highly expressed
 transcripts.
@@
 trees), and instead approximate the NP-Hard problem of determining the optimal
-network structure using a fast hill climbing algorithm. Furthermore, we
+network structure using a fast hill-climbing algorithm. Furthermore, we
 train our model \emph{discriminatively}; only parameters that are deemed
@@
 To determine the structure and parameters of the Bayesian network, we use a
-hill-climbing approach similar to algorithm described by \citet{Grossman2004}.
+hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
 The network structure is determined by greedily optimizing the conditional
@@
 are relative to the read start, which is labeled 0. Hollow circles indicate
-position that were not included in the model, being deemed uninformative, give
-the other positions and edges. The number of parameters of parameters needed to
+positions that were not included in the model, being deemed uninformative, given
+the other positions and edges. The number of parameters needed to
 specify each model is listed in parenthesis below.}
@@
 
-We applied our procedure (labeled ``BN'') as well as those as well as those of
+We applied our procedure (labeled ``BN'') as well as those of
 \citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
@@
 sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
-an unpublished data sets of our own (Table \ref{tab:datasets}).
+an unpublished data set of our own (Table \ref{tab:datasets}).
 
@@
 sites, then chose the top 1000 exons by read count, restricting ourselves those
-at least 100nt long.
+at least 100 nt long.
 
@@
 frequencies used by their model. The training procedure works by simple tallying
-of frequencies. The implementation of this model in the Genominator package we
-found to use a great deal of memory, and were unable to train with the volume of
+of frequencies. The implementation of this model in the Genominator package
+uses a great deal of memory, and we were unable to train with the volume of
 data we wished, so we reimplemented the model and trained it on all of the reads
-aligned to chromosomes 1-8.
+aligned to chromosomes 1--8.
 
@@
 obvious bias. To quantify the non-uniformity observed in these plots, we use the
+symmetrized Kullback-Leibler (KL) divergence \citep{Kullback1951}.
 
 If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
-frequency, the KL divergence is computed as, symmetrized Kullback-Leibler (KL)
-divergence \citep{Kullback1951}.
-%Comparing only nucleotide frequencies may miss
-%dinucleotide or higher order bias. Thus, for a fair comparison, we compute the
-%KL divergence for $k$-mers, with $k \in \{ 1 \dots 6 \}$.
+frequency, the KL divergence is computed as
 $$D_k( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
@@
 explicitly chose exons with high coverage, not counting duplicates may not
-fully capturing the effects of the bias correction methods. Thus, for each exon,
-we compute the number of duplicates for each read, and count only reads in the
+fully capture the effects of the bias correction methods. Thus, for each exon,
+we compute the number of duplicates of each read and count only reads in the
 upper half. Under the assumption of uniform sampling, the set of reads falling
@@
 A large divergence indicates significant bias. Plotted here is the divergence
-from unadjusted reads counts as well as after adjusting reads counts using each
+from unadjusted reads counts as well as after adjusting read counts using each
 method.}
@@
 quantiles, with whiskers extending to 1.5 times the inter-quartile range (i.e.,
-the span between tho 25\% and 75\% quantiles).
+the span between the 25\% and 75\% quantiles).
 }
@@
 \processtable{
-    The $R^2$ goodness of fit statistic measures increased uniformity in read
-    coverage, after correcting for bias. Here the median $R^2$ across the test
-    exons is listed for each method and simple. A positive $R^2$ indicates a
-    better fit.\label{tab:pois}}{
+    The median $R^2$ goodness fo fit statistic across test exons.
+\label{tab:pois}}{
     \begin{tabular}{lcccc}\toprule
@@
     \end{tabular}
-}{}
+}
+{
+    The $R^2$ statistic measures increased uniformity in read coverage, after
+    correcting for bias.  Here the median $R^2$ across the test exons is listed
+    for each method and sample. A positive $R^2$ indicates a better fit.
+}
 \end{table}
@@
 approach was at least as effective as those of \citet{Li2010}, despite not
-requiring gene annotations, or manual selection of training examples.
+requiring gene annotations or manual selection of training examples.
 
@@
 this method is superficially similar to our own, a proper comparison is
-difficult, as the software can not be applied independently of estimating
+difficult, as the software cannot be applied independently of estimating
 transcript abundance in FPKM (fragments per exonic kilobase, per million mapped
@@
 substituted in FPKM estimation, or that a separate interface be written to the
-Cufflinks bias correction method---both comparison requiring significant effort.
+Cufflinks bias correction method---both comparisons requiring significant effort.
 
@@
 sequence probabilities, we make no restriction on the graph other than
-acyclicity, and go to considerable effort to efficiently approximate the optimal
-structure for each data set, rather than using a fixed structure, as in
+acyclicity. We go to considerable effort to efficiently approximate the optimal
+structure for each data set rather than using a fixed structure, as in
 Cufflinks. A ``one size fits all'' approach likely works quite well in many
@@
 against it. For example, the structures learned by our method (Figure
-\ref{fig:models}), are considerably different between those sequenced on an
+\ref{fig:models}) are considerably different between those sequenced on an
 Illumina platform versus an AB platform.
@@
 100,000 reads from the Mortazavi data, training time was approximately 45
-minutes, running on one core of a 3Ghz Intel Xeon processor. However, limiting
+minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
 the training to 25,000 reads leads to a model that is nearly as accurate while
@@
 Because we do not require annotations, ChIP-Seq, and other high-throughput
-sequencing experiments may also benefit from our model. In a preliminary
+sequencing experiments, may also benefit from our model. In a preliminary
 investigation, we found the sequence bias in one ChIP-Seq experiment
 \cite{Cao2010} was less than that observed in any of the RNA-Seq data we
-evaluated, however our method is still able effectively correct for the bias
+evaluated; however, our method is still able to effectively correct for the bias
 that was observed (see, Supplementary Section 2).
@@
 assumptions made by our model, our estimation of bias could easily be
-incorporated into ChIP-Seq peak calling algorithms, and potentially improve
+incorporated into ChIP-Seq peak-calling algorithms, and potentially improve
 accuracy.

### 2011-06-21 build.sh

@@ -6,7 +6,7 @@
 export TEXINPUTS=.:bioinformatics:
 export BSTINPUTS=.:bioinformatics:
 
-TEX=pdflatex
+TEX=latex
 BIB=bibtex
 
 
@@ -15,5 +15,5 @@ $BIB seqbias-paper
 $TEX seqbias-paper.tex
 $TEX seqbias-paper.tex
 
-
+dvipdf seqbias-paper.dvi
 

### 2011-07-08 seqbias-paper.tex
small typo
move figures, to placate bioinformatics upload tool
final revisions before submission

@@
 \footnote{to whom correspondence should be addressed}\hspace{0.5em}
-Walter L. Ruzzo\,$^{1,2}$
-Xinxia Peng\,$^{3}$
-Michael G. Katze$^{3}$
+Walter L. Ruzzo\,$^{1,2,3}$
+Xinxia Peng\,$^{4}$
+Michael G. Katze$^{4}$
 }
@@
 Washington, Seattle, WA 98195-2350, USA\\
-$^{2}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA\\
-$^{3}$Department of Microbiology, University of Washington, Seattle, WA
+$^{2}$Department of Genome Sciences, University of Washington, Seattle, Wa
+98195-5065, USA\\
+$^{3}$Fred Hutchinson Cancer Research Center, Seattle, WA 98109, USA\\
+$^{4}$Department of Microbiology, University of Washington, Seattle, WA
 98195-7242, USA}
@@
 These biases may adversely effect transcript discovery, as low level noise may
-be over reported in some regions, and in others, active transcription may be
-under reported. They render untrustworthy comparisons of relative abundance
+be over-reported in some regions, and in others, active transcription may be
+under-reported. They render untrustworthy comparisons of relative abundance
 between genes or isoforms, and any test of differential expression hangs on the
@@
 \begin{figure*}
-\centerline{\includegraphics[width=0.8\textwidth]{fig/freqs.eps}}
-\caption{Nucleotide frequencies are plotted relative to the start (labeled,
-position 0) of each mapped read, respecting strand. The sequence is taken from
-the genomic context surrounding the read, so that -40 to -1, for example, fall
-outside the read sequence itself. The symmetrized Kullback-Leibler divergence is
-used to summarize the difference in nucleotide frequency compared to a fixed
-estimate of background nucleotide frequencies made by sampling many positions
-nearby mapped reads.  Under the assumption that reads are sampled uniformly from
-transcripts, each of the plots should be essentially flat.}
+\centerline{\includegraphics[width=0.9\textwidth]{freqs.eps}}
+\caption{Nucleotide frequencies are plotted relative to the start (labeled
+position 0) of each mapped read, respecting strand, and grouped by platform
+(Illumina or ABI SOLiD).  The data sets plotted here are those used
+for evaluation, listed in Table \ref{tab:datasets}.
+The sequence is taken from the genomic context surrounding the read, so that -40
+to -1, for example, fall outside the read sequence itself. The symmetrized
+Kullback-Leibler divergence is used to summarize the difference in nucleotide
+frequency compared to a fixed estimate of background nucleotide frequencies made
+by sampling many positions near mapped reads.  Under the assumption that reads
+are sampled uniformly from transcripts, each of the plots should be essentially
+flat}
 \label{fig:freqs}
@@
 \citet{Li2010} propose two models. The first is a Poisson linear
-model, in which read counts across a transcript follow inhomogeneous Poisson
+model, in which read counts across a transcript follow an inhomogeneous Poisson
 process. The read count at position $i$ within the transcript is Poisson
@@
 distribution of initial heptamers within reads, then estimates a presumed
-background heptamer distribution, sampled from the end of reads. The read count
+background heptamer distribution, sampled from the ends of reads. The read count
 at a given position is then adjusted by the ratio of the foreground and
 background heptamer probabilities. Specifying two distributions over heptamers
-(i.e. foreground and background distributions) requires $2(4^7-1) = 32,766$
+(i.e. foreground and background distributions) requires \mbox{$2(4^7-1) =
+32,766$}
 parameters, so while no gene annotations are needed to train such a model, a
@@
 is implemented in the latest version of Cufflinks \citep{Trapnell2010}, and
-tightly incorporated into its estimation of transcript abundance.
+tightly incorporated into its estimation of transcript abundance, requiring
+either predicted or existing gene annotations.
 
@@
 their approach in a way we find to be more robust and effective at correcting
-for bias in a variety of protocols. Due to the weak assumptions required my our
+for bias in a variety of protocols. Due to the weak assumptions required by our
 model, it is applicable and potentially useful in any setting in which short
@@
 discriminating between foreground and background are not included in the model.
-The Bayesian network can then used to evaluate sequence probability, and thus
+The Bayesian network can then be used to evaluate sequence probability, and thus
 bias, at any genomic position. Figure \ref{fig:overview} gives a brief overview
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.35\textwidth]{fig/overview.eps}}
+\centerline{\includegraphics[width=0.35\textwidth]{overview.eps}}
 \caption{An overview of the approach taken: (a) foreground sequences are sampled
@@
 this, we propose using paired training data. Each foreground training
-sequence is paired with a background sequence taken from nearby position
+sequence is paired with a background sequence taken from a nearby position
 that is likely to have similar abundance and general nucleotide composition.
@@
 
-The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
+The methods proposed by \citet{Hansen2010} and \cite{Roberts2011} also treat bias
 correction as a problem of estimating foreground and background sequence
@@
 
-In our model, we do not rely on constraining the set of networks (e.g. to
+In our model, we do not rely on constraining the set of networks (e.g., to
 trees), and instead approximate the NP-Hard problem of determining the optimal
@@
 background, sampled from the reference genome. To obtain the foreground
-sequences, we take sequences surrounding the starts positions of $n/2$ aligned
-reads. We wish to avoid overfitting to any short locus with a large abundance of reads,
-yet we must also capture the bias introduced by PCR amplification. We therefore
-choose the $n/2$ unique reads with the highest number of duplicates,
-yet count each read only once, ignoring duplicates
+sequences, we take sequences surrounding (extending 20 nt to either side, by
+default) the starts positions of $n/2$ aligned reads. We wish to avoid
+overfitting to any short locus with a large abundance of reads, yet we must also
+capture the bias introduced by, for exapmle, PCR amplification. We therefore
+choose the $n/2$ unique reads with the highest number of duplicates, yet count
+each read only once, ignoring duplicates
 
@@
 which the foreground sequences were sampled.  The offset is drawn from a
-zero-mean Gaussian, and rounded to the nearest integer, away from zero.
-By using such a scheme, we attempt to mitigate the effects of biological
-sequence bias, sampling positions that are more likely to be biologically
-similar.
+zero-mean Gaussian (with $\sigma^2 = 10$, by default), and rounded to the
+nearest integer, away from zero.  By using such a scheme, we attempt to mitigate
+the effects of biological sequence bias, sampling positions that are more likely
+to be biologically similar.
 
@@
 The network structure is determined by greedily optimizing the conditional
-log-likelihood.
+log-likelihood:
 \begin{align*}
@@
 \end{align*}
-
-Where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
+where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
 0.5$) if we sample foreground and background positions equally.
@@
 \begin{figure*}
-\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
+\centerline{\includegraphics[width=\textwidth]{models.eps}}
 \caption{
-The network structures learned on each of the data sets are examined. Positions
+The network structures learned on each of the data sets are displayed. Positions
 are relative to the read start, which is labeled 0. Hollow circles indicate
@@
 Experiment & Species & Platform & Read Length \\\midrule
-\citet{Wetterbom2010} & Chimpanzee & AB & 33 \\
-Katze (unpublished) & Macaque & AB & 50 \\
+\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
+Katze (unpublished) & Macaque & ABI & 50 \\
 \citet{Bullard2010} & Human & Illumina & 35 \\
@@
 Our own method was trained on the 100,000 reads from chromosomes 1--8, with the
-highest number of duplicates.
+highest number of duplicate reads.
 
-All data sets were mapped using Bowtie \citep{Langmead2009} against the latest
-genome assemblies available from the UCSC Genome Browser \citep{Karolchik2008}
-during preparation.
+All data sets were mapped using Bowtie \citep{Langmead2009} against,
+respectively, the hg19, mm9, and panTro2 genome assemblies obtained from the
+UCSC Genome Browser \citep{Karolchik2008}.
 
@@
 reads, and then tallying frequencies of overlapping $k$-mers. The $k$-mer
-distribution obtained are then compared to a background distribution obtained by
+distribution obtained is then compared to a background distribution obtained by
 redistributing reads uniformly at random within their exons.
@@
 analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
-remaining cases are plotted in Supplementary Section 4.
+remaining cases are plotted in Supplementary Section 2.
 
 \begin{figure*}
-\centerline{\includegraphics[width=1.00\textwidth]{fig/kl.eps}}
+\centerline{\includegraphics[width=1.00\textwidth]{kl.eps}}
 \caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
@@
 A large divergence indicates significant bias. Plotted here is the divergence
-from unadjusted reads counts as well as after adjusting read counts using each
+from unadjusted read counts as well as after adjusting read counts using each
 method.}
@@
 This measure can be interpreted as the improvement in fit over the null model,
-with $R^2 = 1$ indicating a prefect fit, and a negative numbers, an increasingly
+with $R^2 = 1$ indicating a prefect fit, and smaller numbers, an increasingly
 worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
-proportion of variability enplaned by the model.
+proportion of variability explained by the model.
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
-\caption{For each of the 1000 test exons, we measure a pseudo-coefficients of
+\centerline{\includegraphics[width=0.35\textwidth]{pois-boxplot.eps}}
+\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
 determination $R^2$, equivalent to the improvement in log-likelihood under the
-bias corrected model. The statistic is positive when uniformity is increased,
-and negative when decreased. Boxes are plotted to mark the 25\%, 50\%, and 75\%
-quantiles, with whiskers extending to 1.5 times the inter-quartile range (i.e.,
-the span between the 25\% and 75\% quantiles).
+bias corrected model. The statistic is positive, and increases as uniformity is
+increased, and negative when uniformity is decreased. Boxes are plotted to mark
+the 25\%, 50\%, and 75\% quantiles, with whiskers extending to 1.5 times the
+inter-quartile range (i.e., the span between the 25\% and 75\% quantiles), and
+dots marking more extreme values.
 }
@@
 \processtable{
-    The median $R^2$ goodness fo fit statistic across test exons.
+    The median $R^2$ goodness of fit statistic across test exons.
 \label{tab:pois}}{
@@
     correcting for bias.  Here the median $R^2$ across the test exons is listed
-    for each method and sample. A positive $R^2$ indicates a better fit.
+    for each method and sample. A higher $R^2$ indicates a better fit.
 }
@@
 heptamer frequencies, a likely explanation is that the model is overfit to the
-training set. Though the training set consisted of at least 600,000 reads in
-each experiment, it seems more reads, or a greater diversity of reads are needed
+training set. Though we trained the 7mer model on at least 600,000 reads in each
+experiment, it seems more reads, or a greater diversity of reads are needed
 to produce an accurate or useful estimate.
@@
 
-We have withheld any direct comparison to the method described by
+We have not performed any direct comparison to the method described by
 \citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
@@
 \ref{fig:models}) are considerably different between those sequenced on an
-Illumina platform versus an AB platform.
+Illumina platform versus an ABI platform.
 
@@
 the training to 25,000 reads leads to a model that is nearly as accurate while
-requiring under 4 minutes to train. A full discussion of this trade-off is in
-Supplementary Section 5.
+requiring less than 4 minutes to train. A full discussion of this trade-off is in
+Supplementary Section 3.
 
@@
 evaluated; however, our method is still able to effectively correct for the bias
-that was observed (see, Supplementary Section 2).
+that was observed (see, Supplementary Section 4).
 Protocol differences, as we have seen,
@@
 the bias appears to be largely, but not entirely consistent (see, Supplementary
-Section 3).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
+Section 5).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
 on the Illumina platform display similar patterns of nonuniformity, yet differ

### 2011-07-08 seqbias-supplement.tex
various corrections from markus korth
final revisions before submission

@@
 
-%\setmainfont{Helvetica}
-
 \geometry{left=1.5in,right=1.5in, top=1in, bottom=1in}
@@
 
-%\maketitle
-
-\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{A Model for Sequencing Bias in RNA-Seq}}{Daniel C. Jones}
-
-%\begin{center}
-%\large{
-%\sc{SUPPLEMENTARY}
-
-%\sc{A model for Sequence Bias in RNA-Seq}
-%}
-%\end{center}
+\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{Correcting for Bias in High-Throughput
+Sequencing Data}}{Daniel C. Jones}
 
@@
 Observing the nonuniform distribution of nucleotide frequencies surrounding the
-5' end reads, a natural step to take would be to trim the 5' end before mapping,
+5' end of reads, a natural step to take would be to trim the 5' end before mapping,
 in the hope that simply removing the portion of the read in which the bias
@@
 \end{center}
+
+\section{Additional Kullback-Leibler Divergence Plots}
+
+Below we plot the adjusted and unadjusted KL divergence plots, supplementing
+those in the results section of the main paper.
+
+\includegraphics[width=\textwidth]{fig/kl-all.pdf}
+
+\section{Runtime-Accuracy Trade-off}
+
+Training our model on more reads results in more accurate estimation of bias,
+but a longer training time. Here we investigate that trade-off by training our
+model on progressively more reads from the Mortazavi data set. We use the
+median pseudo-coefficient of determination ($R^2$) over exons selected for our
+test set. Each point is additionally labeled with the time used to train the
+model on one core of a 3Ghz Intel Xeon processor.
+
+\includegraphics[width=\textwidth]{fig/pois-n.eps}
 
@@
 \cite{Hansen2010}. ``7mer'' estimated the initial heptamer, ``Avg-7mer''
-averages the initial two heptamers, and ``4mer'' averages the initial 4mer.
+averages the initial two heptamers, and ``4mer'' averages the initial two 4mers.
 
@@
 
-\section{Additional Kullback-Leibler Divergence Plots}
-
-Below we plot the adjusted and unadjusted KL divergence plots, supplementing
-those in the results section of the main paper.
-
-\includegraphics[width=\textwidth]{fig/kl-all.pdf}
-
-\section{Runtime-Accuracy Trade-off}
-
-Training our model on more reads results in more accurate estimation of bias,
-but a longer training time. Here we investigate that trade-off by training our
-model on progressively more reads from the Mortazavi data set. We use the
-median pseudo-coefficient of determination ($R^2$) over exons selected for our
-test set. Each point is additionally labeled with the time used to train the
-model on one core of a 3Ghz Intel Xeon processor.
-
-\includegraphics[width=\textwidth]{fig/pois-n.eps}
-
 \bibliographystyle{plain}

### 2011-08-18 response/response.tex
notes on revision

\documentclass{article}
\usepackage{dcj}

\begin{document}

%\dcjtitle{One}{Two}{Daniel Jones}
\textbf{Notes on Reviewers Comments} \\
\textbf{\today}

\begin{quote}
Reviewer: 1
Comments to the Author
(1) Page3, 2.2.1 mentions a correction for bias introduced by PCR amplification.  RNA sequencing methods are available where PCR amplification is not used.  Correcting for bias in these cases removes true duplication.  Could the authors please comment on this?
\end{quote}

Because model complexity is penalized, a very sparse, or even empty model will
be trained on a data set without bias.  If PCR amplification is responsible for
a majority of the bias, and an RNA-Seq method without PCR amplification is used
producing largely bias-free data, then using the method to correct for bias
will have little if any effect.

That said, evaluating PCR-free methods of RNA-Seq would be informative to
determine if they truly are without bias.

\begin{quote}
Reviewer: 2
Comments to the Author
The ms presents an approach to estimating the effect of local sequence
context on the rate of seeing aligned reads at each position within a
transcript in RNA-Seq data. The method is inspired by Hansen et al.'s,
which simply determines this bias by comparing heptamer frequencies.

Compared to the brute force approach of Hansen, who estimated model
parameters (the density of a joint distribution) directly from
heptamer frequencies, here a Bayesian network approach is taken, which
allows for a sparser parameterisation. However, the cost for this are
a plethora of tuning parameters (e.g. the complexity penalty, data
sampling, $p_{max}$, $d_{max}$) and a greedy algorithm with apparently no
control of the quality of its solution.
\end{quote}

The method has precisely the following parameters,
\begin{enumerate}
\item The model complexity penality.
\item The variance of the distance at which background sequences are sampled.
\item The number of reads the method is trained on.
\item The size of the window surrounding read starts over which the model should
be trained.
\item $p_{\text{max}}, d_{\text{max}}$
\end{enumerate}

Parameters (3), (4), and (5) are introduced only to control the CPU time needed
to train the algorithm, and the all have an extremely simple interpretation:
bigger is better, but slower.

The important parameters here are (1) and (2).  The fact that we demonstrate the
effectiveness of our method across a diverse group of data sets using identical
parameters, with no specific tuning whatsoever argues that the quality of the
solution is not particularity sensitive to any of these. The paper could make
this fact more explicit.

What I might also make explicit is that, while it is true that the simple
brute-force method has fewer parameters, the two other methods compared here
actually have more parameters. For example the in the MART method, one must set
the complexity penalty, the interaction depth, the number of trees, a lower
bound threshold for read counts, as well as selecting training sequences and
coercing them into very specific input format. In fact, the actual training
procedure has even more parameters that are even more cryptic, but the software
does not provide the means to change these from the default values that the
authors have somehow divined.

\begin{quote}
To justify this cost, the
authors make the argument that their approach avoids the potential
overfitting of a brute force approach such as Hansen's. However, I am
not convinced: they evaluate both methods, theirs and Hansen's, on
'more than 600,000 reads', yet this seems unrealistic: current
datasets have hundreds of millions of reads, and that may just solve
the estimation precision problem in the simple brute force approach.

(I appreciate that the implementation of the brute force approach in
the Genominator package is ridiculously inefficient, but that can
easily be fixed by more competent programming.)
\end{quote}

The paper was a little unclear is this regard. At least 600,000 reads were used
on each data set, but on several of the data sets far more reads were used.  (I
did reimplement the Genominator method in order to do this, which I note in the
paper.) The brute-force solution performs very poorly on every data set,
suggesting that throwing more reads at the problem will not solve it.  The
results as presented are fairly unambiguous: the other methods perform far
better with far fewer reads.

Moreover, even if the brute-force method were effective if trained with hundreds
of millions of reads, many data sets are not this large, rendering
it impractical.

\begin{quote}
The major weakness of the ms, in my view, is the choice of 'figures of
merit' that the authors made (Fig.1, Sections 3.1, 3.2). What I would
have expected to see are

1.) Plots closer to the data such as Supplementary Fig.5 of
reference [1]. That figure shows highly uneven coverage of reads
within exons.
- Does this unevenness go away with the proposed adjustment?
- Also, as is evident from that figure, for some of the experimental
protocols, some regions within highly expressed exons get vanishingly
small read coverage.  These are the really hard situations, and much
more worrying than instances where coverage just fluctuates a bit.
How does the proposed scaling adjustment of Section 2.1 work in these
places (i.e. when $s_{i}$ is very small)?
\end{quote}

I chose not to include such a figure because I assumed it would not be taken
seriously: it would be far too easy to cherry-pick a minuscule genomic interval
that makes even a very shoddy method appear favorable.

Instead I implemented Poisson-regression metric and performed a large
cross-validation experiment to demonstrate that this method is quantifiably more
effective at reducing the observed unevenness across five data sets and
thousands of exons, rather than just qualitatively on one exon. I believe the
figure that the reviewer suggests would be largely uninformative, representing a
minute, unreliable subset of the analysis already presented.  However, it
makes some sense to include it anyway, since such a figure is so easily interpreted
that the point of the analysis can be understood without having to read the text
of results section.

\begin{quote}
2.) A 'bottom-line', biologically relevant figure of merit: E.g., does
the identification of differentially expressed genes improve? The
clustering of gene expression profiles? The identification of
transcript isoforms?
\end{quote}

I believe that the paper effectively argues that the accuracy of quantification
is increased, but it is true that it does not demonstrate what effect this has
on downstream analysis. This would take considerable time and effort to
establish this rigorously with a large sample size, but it could be worthwhile
to include some token examples, which might help address what is admittedly a
pretty natural question.

\begin{quote}
Minor points:
------------
1. The ms seems confused between classification and regression.
Reference [2] might be helpful to resolve some of that.
\end{quote}

I used the term ``classification'' in the context of noting that the algorithm I
use is adapted from one typically used for classification. It is correct to say
that the method here performs regression, not classification, and I was maybe
unclear.

\begin{quote}
2. The chosen Bayesian network is rather complicated, and a spectrum
of alternative methods exists between that and the brute force
approach of Hansen (say, lasso or svm regression). In absence of a
more meaningful benchmark, the choice of method seems arbitrary.
\end{quote}

This seems to ask, ``might there exist an even better method?'' Which is
possible, yet not possile to know without implementing and evaluating many
methods, which is unrealistic. Bayesian networks have a very direct
probabilistic interpretation in this context, which is why they are used here
rather than SVMs or Lasso regression, methods whose output would be somewhat
more opaque.

\end{document}

### 2011-08-18 response/revision_plan.tex
notes on revision

\documentclass{article}
\usepackage{dcj}

\begin{document}

\dcjtitle{Seqbias Paper}{Plan for Revision}{Daniel Jones}

\begin{enumerate}

\item Demonstrated the results are not particularity sensitive to the
complexity penalty, drawing a plot similar to the one is Supplementary Section
3, but varying the penalty rather than the number or reads.

\item Point out that the fact that the method is effective over diverse sets of
data argues that the results are not sensitive to any of the parameters used.

\item Note that the Hansen's brute-force method was trained using many more than
600,000 reads on some of the data sets.

\item (Maybe) Draw a curve showing how the performance of Hansen's brute-force
method changes as a function of the number of reads fed into it.

\item Make very clear that in there is no bias present in the data, a sparse or
empty model will be trained.

\item Include a figure from the genome browser showing an example uniformity of
coverage being increased over a gene or exon.

\item Use the MAQC set to demonstrate the following:
\begin{enumerate}
\item Increased agreement between bias corrected sequencing data and qPCR.
\item Increased agreement between replicates.
\end{enumerate}

\end{enumerate}

\end{document}

### 2011-08-19 seqbias-paper.tex
added frt-seq data

@@
 evaluated; however, our method is still able to effectively correct for the bias
-that was observed (see, Supplementary Section 4).
-Protocol differences, as we have seen,
-can result in significant differences in observed nucleotide frequencies, so we
-may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
-assumptions made by our model, our estimation of bias could easily be
-incorporated into ChIP-Seq peak-calling algorithms, and potentially improve
-accuracy.
+that was observed (see, Supplementary Section 4).  Protocol differences, as we
+have seen, can result in significant differences in observed nucleotide
+frequencies, so we may not safely assert that bias in ChIP-Seq data is always
+low.  Given the weak assumptions made by our model, our estimation of bias could
+easily be incorporated into ChIP-Seq peak-calling algorithms, and potentially
+improve accuracy.
+
+To determine the extent to which PCR amplification is responsible for the
+observed bias, we evaluated data from the FRT-Seq method proposed by
+\citet{Mamanova2010}.  FRT-Seq avoids the PCR amplification step during library
+preparation with reverse transcription occuring on the flowcell surface. We
+observed that this data is not free from sequence bias, yet unlike other data
+generated on the Illumina platform, it appears to be effected only by relatively
+few positions adjacent to the read start (see, Supplementary Section 5).
 
@@
 the bias appears to be largely, but not entirely consistent (see, Supplementary
-Section 5).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
+Section 6).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
 on the Illumina platform display similar patterns of nonuniformity, yet differ

### 2011-08-19 seqbias-supplement.tex
added frt-seq data

@@
 
+\section{Bias in Amplification-Free Sequencing}
+
+To evaluate the extent to which the observed bias is caused by PCR amplification
+in the library preparation stage, we analyzed data from the FRT-Seq protocol
+developed by Mamanova, et. al. \cite{Mamanova2010}. FRT-Seq avoids the
+amplification step during library preparation with reverse transcription
+occuring on the flowcell surface. We obtained data generated by Mamanova, et.
+al., from the European Nucleotide Archive with accession code ERR007689, and
+mapped them to the hg19 assembly of the human genome with Bowtie
+\cite{Langmead2009}. The nucleotide frequencies and divergence are plotted
+below.
+
+\begin{center}
+\includegraphics[width=0.6\textwidth]{fig/frt-all.pdf}
+\end{center}
+
+It is clear from this analysis that the FRT-Seq method, as implemented to
+generate this data, is not without bias, yet there is significant divergence at
+only a small number of positions surrounding the read start. When our method is
+trained on this data we obtain the suitably sparse model pictured below.
+
+\begin{center}
+\includegraphics[width=0.4\textwidth]{fig/frt-graph.pdf}
+\end{center}
+
 \section{Variability Between Replicates}

### 2011-08-19 results/plots/plot-frt.R

@@ -1,0 +1,26 @@
+
+library(Cairo)
+library(ggplot2)
+source('theme_dcjstd.R')
+
+xs <- read.table('../frtseq/freqs.tab', header = T)
+xs$kmer <- factor(xs$kmer, levels = c('a', 'c', 'g', 't', 'div'))
+
+scale = 0.9
+
+CairoSVG('frt.svg', width = scale * 12, height = scale * 6)
+ys <- subset(xs, kmer != 'div')
+p <- qplot(data = ys, x = pos, y = freq, geom = 'line')
+p <- p + facet_grid(kmer ~ .)
+p <- p + theme_dcjstd()
+print(p)
+dev.off()
+
+
+CairoSVG('frt-div.svg', width = scale * 12, height = scale * 2)
+ys <- subset(xs, kmer == 'div')
+p <- qplot(data = ys, x = pos, y = freq, geom = 'line')
+p <- p + facet_grid(kmer ~ ., scales = 'free')
+p <- p + theme_dcjstd()
+print(p)
+

### 2011-08-19 results/plots/plot-kl.R

@@ -72,7 +72,7 @@ p <- p + scale_color_manual(
 
                              
 
-# get rid of grid lines, since they get fucket up on this plot anyway
+# get rid of grid lines, since they get fucked up on this plot anyway
 p <- p + theme_dcjstd()
 
 #p <- p + structure(list(panel.grid.minor = theme_blank(),

### 2011-08-19 results/plots/theme_dcjstd.R

@@ -1,16 +1,38 @@
 
 
 
-theme_dcjstd <- function()
+theme_dcjstd <- function(base_size = 12, base_family = "")
 {
     structure(
         list(
-            panel.background = theme_rect(fill   = '#f5f5f5',
-                                          colour = '#e6e6e6'),
+            panel.background = theme_rect(fill   = '#eaeaea',
+                                          colour = '#d0d0d0'),
 
             panel.grid.major = theme_line(colour = "white"), 
 
-            panel.grid.minor = theme_line(colour = "white", size = 0.25)
+            panel.grid.minor = theme_line(colour = "white", size = 0.25),
+
+
+            axis.text.x = theme_text(family     = base_family, 
+                                     size       = base_size * 0.8,
+                                     lineheight = 0.9,
+                                     colour     = "black", 
+                                     vjust      = 1),
+
+            axis.text.y = theme_text(family     = base_family, 
+                                     size       = base_size * 0.8,
+                                     lineheight = 0.9,
+                                     colour     = "black", 
+                                     hjust      = 1),
+
+            axis.title.x = theme_text(family = base_family,
+                                      size   = base_size, 
+                                      vjust  = 1),
+
+            axis.title.y = theme_text(family = base_family, 
+                                      size   = base_size,
+                                      angle  = 90,
+                                      vjust  = 0.2)
         ),
 
         class = "options"

### 2011-08-23 seqbias-paper.tex
parameter sensitivity explored in the suplement

@@
 sequences are sampled by randomly offsetting foreground positions, (c) a
-Bayesian network is classifier is trained to discriminate between the set of
-sampled foreground and background sequences, (d) and the classifier is evaluated
+Bayesian network is trained to discriminate between the set of
+sampled foreground and background sequences, (d) and the model is evaluated
 at each position within a locus, predicting bias.}
@@
 overfitting to any short locus with a large abundance of reads, yet we must also
-capture the bias introduced by, for exapmle, PCR amplification. We therefore
+capture the bias introduced by, for example, PCR amplification. We therefore
 choose the $n/2$ unique reads with the highest number of duplicates, yet count

### 2011-08-23 seqbias-supplement.tex
parameter sensitivity explored in the suplement

@@
 \includegraphics[width=\textwidth]{fig/pois-n.eps}
+
+\section{Sensitivity of Parameters}
+
+The performance of our method is dependent on two parameters: the standard
+deviation at which background sequences are sampled, and the degree to which
+model complexity in penalized. We have also introduced parameters limiting the
+number of parents a node may have ($p_{\text{max}}$) as well as the distance
+that an edge may traverse ($d_{\text{max}}$), but these exist only to control
+the amount of CPU time used and have a much simpler interpretation: bigger is
+better, but slower.
+
+Here we retrain the model on 50,000 reads from the Mortazavi data
+set, varying these first two parameters to assess the sensitivity of the
+observed results to their values.
+
+\begin{center}
+\includegraphics[width=0.8\textwidth]{fig/pois-sampstd.pdf}
+\end{center}
+
+The above plot show the standard deviation used to draw background samples
+varied from 1 to 10,000. Apparent from this plot is that, while the optimal
+choice lies somewhere between 10 and 100, the model performs competitively for
+any reasonable value.
+
+\begin{center}
+\includegraphics[width=0.8\textwidth]{fig/pois-comppen.pdf}
+\end{center}
+
+Here we multiply the complexity penalty term of the BIC by constants varying
+from $10^{-6}$ to $10^{3}$. (Recall that the BIC is $2\ell - m \log n$ where
+$\ell$ is the log-likelihood, $m$ is the number of parameters needed to specify
+the model, and $n$ is the number of training examples. Here we compute $2 \ell -
+c m \log n$, varying $c$.)
+
+It is clear from the above figure, that a very severe complexity penalty will
+result in a poor model. Here, for constants larger than 100, an empty model is
+trained, resulting in the goodness of fit of 0. Conversely, if this constant is
+very small, a overly-dense model will be trained.  In this data, for constants
+less than 0.01, the maximally dense model is trained (given the restraints on
+in-degree and edge distance). This result is a sub-optimal but entirely adequate
+solution.
 

### 2011-08-23 response/revision_plan.tex
added frt-seq data
parameter sensitivity explored in the suplement
comments regarding parameter sensitivity

@@
 \usepackage{dcj}
+\usepackage{ulem}
 
@@
 
-\item Demonstrated the results are not particularity sensitive to the
+\item \sout{Demonstrated the results are not particularity sensitive to the
 complexity penalty, drawing a plot similar to the one is Supplementary Section
-3, but varying the penalty rather than the number or reads.
+3, but varying the penalty rather than the number or reads.}
 
-\item Point out that the fact that the method is effective over diverse sets of
-data argues that the results are not sensitive to any of the parameters used.
+\item \sout{Point out that the fact that the method is effective over diverse sets of
+data argues that the results are not sensitive to any of the parameters used.}
 
@@
 
-\item (Maybe) Draw a curve showing how the performance of Hansen's brute-force
-method changes as a function of the number of reads fed into it.
+%\item (Maybe) Draw a curve showing how the performance of Hansen's brute-force
+%method changes as a function of the number of reads fed into it.
 
+% This is sort of implied by the FRT-seq data
 \item Make very clear that in there is no bias present in the data, a sparse or
@@
 
-\item Include a figure from the genome browser showing an example uniformity of
-coverage being increased over a gene or exon.
+\item \sout{Include a figure from the genome browser showing an example uniformity of
+coverage being increased over a gene or exon.}
 
@@
 \item Increased agreement between bias corrected sequencing data and qPCR.
-\item Increased agreement between replicates.
+%\item Increased agreement between replicates. % Won't work here
 \end{enumerate}
+
+\item \sout{Examine the bias in new methods without a PCR-amplification step.}
+
+\item \sout{Typos mentioned by reviewr 1.}
 

### 2011-08-23 seqbias-paper.tex
comments regarding parameter sensitivity

@@
 \begin{figure*}
-\centerline{\includegraphics[width=0.9\textwidth]{freqs.eps}}
+\centerline{\includegraphics[width=0.85\textwidth]{freqs.eps}}
 \caption{Nucleotide frequencies are plotted relative to the start (labeled
@@
 correct for bias.  First, define the \emph{sequence bias} $b_i$ at position $i$
-as,
-$$ b_i = \frac{ \Pr[ s_i ] } { \Pr[ s_i | m_i ] } $$
+as $ b_i = \Pr[ s_i ] /  \Pr[ s_i | m_i ]  $.
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.35\textwidth]{overview.eps}}
+\centerline{\includegraphics[width=0.30\textwidth]{overview.eps}}
 \caption{An overview of the approach taken: (a) foreground sequences are sampled
@@
 
+Along with $p_{\text{max}}$ and $d_{\text{max}}$, the time required to train the
+model is controlled by the number of reads used (i.e., $n$) and the number of
+positions surrounding the read start to consider. Increasing any of these
+parameters results in a solution that is no worse, and potentially better, at
+the expense of increased training time.
+
+The quality of the solution depends also on two other parameters: the standard
+deviation at which background sequences are sampled, and the weight applied to
+the penalty term of the BIC, yet it is not particularly sensitive to their values
+(see, Supplementary Section 4). In all our evaluation, no special tuning of the
+parameters was performed, suggesting it can be used effectively across data sets
+without any intervention.
+
 Figure \ref{fig:models} shows examples of the structure learned when this
@@
 \begin{figure*}
-\centerline{\includegraphics[width=1.00\textwidth]{kl.eps}}
+\centerline{\includegraphics[width=0.95\textwidth]{kl.eps}}
 \caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
@@
 
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{uniformity.eps}}
+\caption{Read counts (i.e., the number of reads starting at each position)
+from the Mortazavi data set are plotted along the $3'$ UTR of Apoa2. The upper
+plot shows unadjusted counts, the lower, counts adjusted by our method.}
+\label{fig:unif}
+\end{figure}
+
 In this comparison, we measure how well the counts conform to a Poisson process.
@@
 the fit derived from correcting for bias remains a principled and easily
-interpreted criterion.
+interpreted criterion. This increase in uniformity is illustrated in Figure
+\ref{fig:unif}.
 
 We perform maximum-likelihood fitting of two models. In the null model, the
-Poisson rate is fixed for each exon. That is, for position $j$ within
-exon $i$, the rate is
-$$ \lambda_{ij} = a_i $$
-where $a_i$ is the parameter being fit. For comparison, we then fit a model in
-which the rate is also proportional to the predicted bias coefficients:
-$$ \lambda'_{ij} = a_i b_{ij} $$
+Poisson rate is fixed for each exon. That is, for position $j$ within exon $i$,
+the rate is $ \lambda_{ij} = a_i $ where $a_i$ is the parameter being fit. For
+comparison, we then fit a model in which the rate is also proportional to the
+predicted bias coefficients: $ \lambda'_{ij} = a_i b_{ij} $.
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.35\textwidth]{pois-boxplot.eps}}
+\centerline{\includegraphics[width=0.30\textwidth]{pois-boxplot.eps}}
 \caption{For each of the 1000 test exons, we compute a pseudo-coefficients of

### 2011-08-25 response/response.tex
work on the response

@@
 %\dcjtitle{One}{Two}{Daniel Jones}
-\textbf{Notes on Reviewers Comments} \\
-\textbf{\today}
+%\textbf{Notes on Reviewers Comments} \\
+%\textbf{\today}
+
+We appreciate the time and effort extended by the anonymous reviewers and the
+editorial staff to assess the suitability of the original version of this
+manuscript for publication in Bioinformatics. Upon review, the second of the two
+reviewers raised substantial concerns, compelling the its rejection.  However,
+we found the bulk of the these concerns to be either incongruous with the
+content of the manuscript or addressable with very little effort, and therefore
+not warranting rejection. Here we have provided a detailed response to these
+comments, and have included additional material in the manuscript in an effort
+to clarify these points. Under these circumstances we respectively request that
+the manuscript, in this revised form, be reconsidered for publication.
+
+\subsection*{Response to Reviewer 1}
 
 \begin{quote}
-Reviewer: 1
-Comments to the Author
 (1) Page3, 2.2.1 mentions a correction for bias introduced by PCR amplification.
@@
 
-That said, evaluating PCR-free methods of RNA-Seq would be informative to
-determine if they truly are without bias.
+We have included new data examining the FRT-Seq protocol of Mamanova, et. al.,
+in which no PCR-amplification in performed during library preparation. We show
+that bias is reduced, but not eliminated in this method, and that our method
+accounts for this reduced bias by constructing a much sparser model, relative to
+the other RNA-Seq data sets we have examined.
+
+\subsection*{Response to Reviewer 2}
 
 \begin{quote}
-Reviewer: 2
-Comments to the Author
 The ms presents an approach to estimating the effect of local sequence
@@
 \begin{enumerate}
-\item The model complexity penality.
-\item The variance of the distance at which background sequences are sampled.
 \item The number of reads the method is trained on.
@@
 be trained.
-\item $p_{\text{max}}, d_{\text{max}}$
+\item The maximum in-degree (i.e., $p_{\text{max}}$)
+\item The maximum edge distance (i.e., $d_{\text{max}}$).
+\item The standard deviation at which background sequences are sampled.
+\item The model complexity penalty.
 \end{enumerate}
 
-Parameters (3), (4), and (5) are introduced only to control the CPU time needed
-to train the algorithm, and the all have an extremely simple interpretation:
-bigger is better, but slower.
+Parameters (1), (2), (3), and (4) are introduced only to reduce the CPU time
+needed to train the algorithm. Each has the same very simple interpretation: a
+larger value leads to solution that is no worse and potentially better, at the
+cost of increased training time.  We mention these parameters in the manuscript
+only for the sake of transparency.  They generally require no intervention or
+tuning, being essentially an implementation detail.
 
-The important parameters here are (1) and (2).  The fact that we demonstrate the
-effectiveness of our method across a diverse group of data sets using identical
-parameters, with no specific tuning whatsoever argues that the quality of the
-solution is not particularity sensitive to any of these. The paper could make
-this fact more explicit.
+Parameters (5) and (6) are more important to the quality of the solution, yet we
+have demonstrated, with no special tuning, the method is effective across a
+diverse collection of data sets. In this revision we have included new data
+showing explicitly that the quality of the solution is remarkably insensitive to
+values of these two parameters.
 
-What I might also make explicit is that, while it is true that the simple
-brute-force method has fewer parameters, the two other methods compared here
-actually have more parameters. For example the in the MART method, one must set
-the complexity penalty, the interaction depth, the number of trees, a lower
-bound threshold for read counts, as well as selecting training sequences and
-coercing them into very specific input format. In fact, the actual training
-procedure has even more parameters that are even more cryptic, but the software
-does not provide the means to change these from the default values that the
-authors have somehow divined.
+It should also be noted that, while the brute-force method of Hansen, et. al.,
+has fewer parameters. The other methods evaluated in the manuscript (the GLM and
+MART methods of Li, et. al.), involve more more parameters that the method we
+have described. (For example training the MART method of Li, et al., involves
+parameters controlling the complexity penalty, the number of trees, the lower
+bound threshold for read counts, in addition to others that are not described or
+easily changed from the values divined by the authors.) Our method therefore
+presents a simplification over competing methods.
 
@@
 
-The paper was a little unclear is this regard. At least 600,000 reads were used
-on each data set, but on several of the data sets far more reads were used.  (I
-did reimplement the Genominator method in order to do this, which I note in the
-paper.) The brute-force solution performs very poorly on every data set,
-suggesting that throwing more reads at the problem will not solve it.  The
-results as presented are fairly unambiguous: the other methods perform far
+We regret that we more not more explicit on this point. At least 600,000 reads
+on each data set, but as many as 23 million were used. (As noted in the paper,
+we did reimplement the Genominator method in order to do this.) The brute-force
+method performed very poorly on every data set, suggesting that simply throwing
+more data at a suboptimal solution will not adequately solve the problem. The
+results as they were presented are unambiguous: the other methods perform far
 better with far fewer reads.
 
-Moreover, even if the brute-force method were effective if trained with hundreds
-of millions of reads, many data sets are not this large, rendering
-it impractical.
+Moreover, even if the brute-force method were effective if trained with many
+hundreds of millions of reads, many (possibly, most) existing data sets are not
+this large, rendering it generally impractical.
 
@@
 
-I chose not to include such a figure because I assumed it would not be taken
-seriously: it would be far too easy to cherry-pick a minuscule genomic interval
-that makes even a very shoddy method appear favorable.
+During preparation we choose not to include a figure such as the one suggested
+by the reviewer as we believed it would not be legitimately informative. It
+would be far too easy to cherry-pick a minuscule genomic interval that makes
+even the very shoddy method appear favorable.
 
-Instead I implemented Poisson-regression metric and performed a large
-cross-validation experiment to demonstrate that this method is quantifiably more
-effective at reducing the observed unevenness across five data sets and
-thousands of exons, rather than just qualitatively on one exon. I believe the
-figure that the reviewer suggests would be largely uninformative, representing a
-minute, unreliable subset of the analysis already presented.  However, it
-makes some sense to include it anyway, since such a figure is so easily interpreted
-that the point of the analysis can be understood without having to read the text
-of results section.
+Instead, we presented data (in Section 3.2) that answers the reviewers question
+(i.e., ``does this unevenness go away with the proposed adjustment?'') in
+rigorous and comprehensive manner. We implemented a Poisson-regression metric
+and performed a large cross-validation study proving the effectiveness of our
+method at increasing uniformity quantitatively, across thousands of exons drawn
+from a diverse group of data sets, rather than qualitatively across one exon.
+
+In this revision, we have included a figure, as suggested, showing an example of
+this increase in uniformity. Yet, we must stress, this figure represents a
+minute, unreliable subset of the analysis that had previously been presented,
+and is included only as an illustration.
 
@@
 
-I believe that the paper effectively argues that the accuracy of quantification
-is increased, but it is true that it does not demonstrate what effect this has
-on downstream analysis. This would take considerable time and effort to
-establish this rigorously with a large sample size, but it could be worthwhile
-to include some token examples, which might help address what is admittedly a
-pretty natural question.
+In Section 3, we demonstrate, after correction, a significant improvement in
+the two obvious anomalies associated with the issue of sequence bias:
+non-uniform read coverage, and divergent nucleotide frequencies. We assess these
+aspects in a statistically rigorous manner and with a large sample size. We
+believe that this analysis presents a compelling case that bias correction is
+worthwhile, and that the method we present is competitive with, and quite often
+superior to existing methods.
+
+In this revision, we have included more data, showing that in addition to the
+improvements in uniformity and nucleotide frequencies, increased correlation
+between RNA-Seq and qPCR, after adjusting for bias.
 
@@
 
-I used the term ``classification'' in the context of noting that the algorithm I
-use is adapted from one typically used for classification. It is correct to say
-that the method here performs regression, not classification, and I was maybe
-unclear.
+The term ``classification'' was used in the context of noting that the algorithm
+is adapted from one typically used in classification. It is correct to say that
+the method here performs regression, and we have tried to clarify this language
+in this revision.
 
@@
 
-This seems to ask, ``might there exist an even better method?'' Which is
-possible, yet not possile to know without implementing and evaluating many
-methods, which is unrealistic. Bayesian networks have a very direct
-probabilistic interpretation in this context, which is why they are used here
-rather than SVMs or Lasso regression, methods whose output would be somewhat
-more opaque.
+Our motivation for adopting a Bayesian network model is the very natural
+probabilistic interpretation that we have described in Section 2.1. We
+disagree with the reviewer's opinion that Bayesian networks are inherently more
+complicated than Lasso regression or support vector machines.  In this
+application, the latter two methods would provide output that is arguably less
+transparent and interpretable, and neither be significantly simpler to
+implement, nor involve fewer parameters.
+
+Naturally, it is possible that another model might perform as well or better
+than the one presented here, yet it is unrealistic or impossible to prove that
+this is not case. What we have demonstrated is a model that is well reasoned,
+efficiently implemented, and effective at reducing sequence bias in RNA-Seq.
 

### 2011-08-25 seqbias-paper.tex
work on the response

@@
 
-To estimate sequencing bias, we train a Bayesian network classifier in which
-each node represents a position in the sequence, relative to the read start, and
-edges encode dependency between positions.  Bayesian networks have been applied
-to recognize motifs in nucleotide sequences in the past, in particular in
-modeling splice sites \citep{Cai2000, Chen2005} and transcription factor binding
-sites \citep{Ben-Gal2005, Grau2006, Pudimat2005}.
+To estimate sequencing bias, we train a Bayesian network in which each node
+represents a position in the sequence, relative to the read start, and edges
+encode dependency between positions.  Bayesian networks have been applied to
+recognize motifs in nucleotide sequences in the past, in particular in modeling
+splice sites \citep{Cai2000, Chen2005} and transcription factor binding sites
+\citep{Ben-Gal2005, Grau2006, Pudimat2005}.
 
@@
 the other positions and edges. The number of parameters needed to
-specify each model is listed in parenthesis below.}
+specify each model is listed in parenthesis below. Applied to data with less
+bias, a sparser model is trained, as evinced by the Wetterbom data set.
+}
 \label{fig:models}

### 2011-08-31 seqbias-paper.tex
finishing touches

@@
 \begin{figure}
-\centerline{\includegraphics[width=0.30\textwidth]{overview.eps}}
+\centerline{\includegraphics[width=0.25\textwidth]{overview.eps}}
 \caption{An overview of the approach taken: (a) foreground sequences are sampled
@@
 \begin{align*}
-\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
-&=
+\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) 
+=
 \sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
@@
 parameters needed to specify the model, we maximize, 
-
 $$ \ell' = 2 \ell - m \log n $$
@@
 \citep{McFadden1974}), defined as,
-$$R^2 = 1 - \frac{L'}{L}$$
+$R^2 = 1 - L'/L$.
 
@@
 method to account for this bias without the need of existing gene annotations.
+The analysis provided demonstrates that our method shows significant improvement
+in two aspects: uniformity of read coverage, and consistency of nucleotide
+frequencies. These results strongly suggest an increase in accuracy, and we show
+in Supplementary Section 8, that this conclusion is further justified by
+comparison to qRT-PCR.
 
@@
 heptamer frequencies, a likely explanation is that the model is overfit to the
-training set. Though we trained the 7mer model on at least 600,000 reads in each
-experiment, it seems more reads, or a greater diversity of reads are needed
-to produce an accurate or useful estimate.
+training set.  We trained the 7mer model on a minimum of 600,000 reads (with the
+Bullard data set), and up to 23 million reads (with the Wetterbom data set), yet
+it seems more reads, or a greater diversity of reads are needed to produce an
+accurate or useful estimate.
 

### 2011-08-31 seqbias-supplement.tex
finishing touches

@@
 
+\section{Comparison to qRT-PCR}
+
+We used sequencing data previously published by Au, et al, \cite{Au2010} to
+evalute the effect bias correction has on correlation to measurements made by
+TaqMan RT-PCR, made available by the Microarray Quality Control project
+\cite{Shi2006}. The RNA-Seq data shows a pattern of bias similar to that seen in
+other samples sequenced on an Illumina platform, as plotted below.
+
+\centerline{\includegraphics[width=0.6\textwidth]{fig/au-freqs.pdf}}
+
+To evaluate the efficacy of each of the bias correction methods considered,
+
+We counted reads overlapping each gene, defining the gene by the union of every
+transcript annotated in release 62 of the Ensembl gene annotations, excluding
+UTRs, and 100 additional bases from both ends of the gene. Counts were then
+normalized (divided) by the adjusted length of these genes.  We then remove any
+genes with reads fewer than 10 reads, or without a unique TaqMan probe, leaving
+592 genes.
+
+Each bias correction method was trained using a procedure identical to the one
+described in the results section. After adjusting read counts according to the
+predicted sequence bias, we compute the Pearson correlation between log read counts
+and log TaqMan expression values, which are averaged across three replicates. The
+expression values
+
+These correlations are listed below.
+
+\begin{center}
+\begin{tabular}{ll}
+\textbf{Method} & \textbf{Correlation} \\ \hline
+Unadjusted &  0.6665 \\
+7mer &  0.6677 \\
+GLM & 0.6854 \\
+MART & 0.6921 \\
+BN & 0.7016
+\end{tabular}
+\end{center}
+
+These results are consistent with with the KL-divergence and Poisson regression
+analysis presented in Section 3. We see only very small gains with the
+brute-force 7mer method. The GLM and MART models show much better performance,
+and our Baysian network (BN) method shows moderate increase correlation over these.
+
 \bibliographystyle{plain}

### 2011-09-09 response/response.tex
additions to the response letter

@@
 \documentclass{article}
-\usepackage{dcj}
+\usepackage{amsmath}
+%\usepackage{dcj}
 
@@
 
-%\dcjtitle{One}{Two}{Daniel Jones}
-%\textbf{Notes on Reviewers Comments} \\
-%\textbf{\today}
+We appreciate the time and effort extended by the anonymous reviewers and the
+editorial staff to evaluate the original version of this manuscript, which was
+submitted on July 8. Upon review, the second of the two reviewers raised some
+concerns, compelling its rejection. However, we believe that the bulk of these
+concerns are either incongruous with the content of the manuscript or
+addressable with very little effort, and therefore do not warrant its rejection.
+Here we have provided a detailed response to these comments, and have included
+additional material in an effort to clarify these points.
 
-We appreciate the time and effort extended by the anonymous reviewers and the
-editorial staff to assess the suitability of the original version of this
-manuscript for publication in Bioinformatics. Upon review, the second of the two
-reviewers raised substantial concerns, compelling the its rejection.  However,
-we found the bulk of the these concerns to be either incongruous with the
-content of the manuscript or addressable with very little effort, and therefore
-not warranting rejection. Here we have provided a detailed response to these
-comments, and have included additional material in the manuscript in an effort
-to clarify these points. Under these circumstances we respectively request that
-the manuscript, in this revised form, be reconsidered for publication.
+The second reviewer raised three major concerns. The first concern raised is
+that the method we present is unjustifiably complex, involving a ``plethora of
+tuning parameters'', and that the much simpler method of Hansen, et al., would
+perform as well is trained on enough data.  In the data we presented, the simple
+method developed by Hansen, et al., performed poorly regardless of the number of
+reads used in training, which varied between hundreds of thousands to tens of
+millions.  The methods presented by Li, et al., are more effective, yet involve
+more parameters, and require more time to train, so that our method, while often
+being more effective, also offers a significant simplification.
+
+Secondly, it is asserted that we did not demonstrate that uniformity of read
+coverage is increased ofter correcting for bias using our method.  Yet, in
+section 3.2 we had presented a rigorous statistical analysis over a large and
+diverse collection of data sets showing that this is indeed the case. This
+section of the paper was evidently overlooked by the reviewer, as it received
+no mention in this criticism.
+
+Finally, the reviewer doubts the biological relevance of the bias correction
+methods. Having demonstrated an increase in uniformity (section 3.2) and a
+decrease in nucleotide frequency divergence (section 3.1), we believe a
+compelling case was made that correcting for bias increases the accuracy of
+quantification. In this revision we have supplied another line of evidence:
+analysis showing that correlation between RNA-Seq and qRT-PCR is increased after
+correcting for bias.
+
+Under these circumstances we respectively request that the manuscript, in this
+revised form, be reconsidered for publication.
 
@@
 RNA sequencing methods are available where PCR amplification is not used.
-Correcting for bias in these cases removes true duplication.
-Could the authors please comment on this?
+Correcting for bias in these cases removes true duplication.  Could the authors
+please comment on this?
 \end{quote}
@@
 It should also be noted that, while the brute-force method of Hansen, et. al.,
-has fewer parameters. The other methods evaluated in the manuscript (the GLM and
-MART methods of Li, et. al.), involve more more parameters that the method we
-have described. (For example training the MART method of Li, et al., involves
-parameters controlling the complexity penalty, the number of trees, the lower
-bound threshold for read counts, in addition to others that are not described or
+has fewer parameters, the other methods evaluated in the manuscript (the GLM and
+MART methods of Li, et. al.), involve more parameters that the method we
+have described. (For example training the MART method involves parameters
+controlling the complexity penalty, the number of trees, the lower bound
+threshold for read counts, in addition to others that are not described or
 easily changed from the values divined by the authors.) Our method therefore
@@
 Moreover, even if the brute-force method were effective if trained with many
-hundreds of millions of reads, many (possibly, most) existing data sets are not
+hundreds of millions of reads, many (possibly most) existing data sets are not
 this large, rendering it generally impractical.
@@
 (i.e., ``does this unevenness go away with the proposed adjustment?'') in
-rigorous and comprehensive manner. We implemented a Poisson-regression metric
+rigorous and comprehensive manner. We implemented a Poisson regression metric
 and performed a large cross-validation study proving the effectiveness of our
@@
 improvements in uniformity and nucleotide frequencies, increased correlation
-between RNA-Seq and qPCR, after adjusting for bias.
+between RNA-Seq and qRT-PCR, after adjusting for bias in RNA-Seq. These results
+are consistent with, and reinforce our previous results.
 
@@
 complicated than Lasso regression or support vector machines.  In this
-application, the latter two methods would provide output that is arguably less
+application, the latter two methods would generate output that is arguably less
 transparent and interpretable, and neither be significantly simpler to

### 2011-09-09 seqbias-paper.tex
additions to the response letter

@@
 the penalty term of the BIC, yet it is not particularly sensitive to their values
-(see, Supplementary Section 4). In all our evaluation, no special tuning of the
+(see, Supplementary Section 2). In all our evaluation, no special tuning of the
 parameters was performed, suggesting it can be used effectively across data sets
@@
 analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
-remaining cases are plotted in Supplementary Section 2.
+remaining cases are plotted in Supplementary Section 3.
 
@@
 frequencies. These results strongly suggest an increase in accuracy, and we show
-in Supplementary Section 8, that this conclusion is further justified by
+in Supplementary Section 4, that this conclusion is further justified by
 comparison to qRT-PCR.
@@
 requiring less than 4 minutes to train. A full discussion of this trade-off is in
-Supplementary Section 3.
+Supplementary Section 5.
 
@@
 evaluated; however, our method is still able to effectively correct for the bias
-that was observed (see, Supplementary Section 4).  Protocol differences, as we
+that was observed (see, Supplementary Section 6).  Protocol differences, as we
 have seen, can result in significant differences in observed nucleotide
@@
 generated on the Illumina platform, it appears to be effected only by relatively
-few positions adjacent to the read start (see, Supplementary Section 5).
+few positions adjacent to the read start (see, Supplementary Section 7).
 
@@
 the bias appears to be largely, but not entirely consistent (see, Supplementary
-Section 6).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
+Section 8).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
 on the Illumina platform display similar patterns of nonuniformity, yet differ
@@
 
-%\section*{Acknowledgement}
-%Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text text text text
-
 \paragraph{Funding\textcolon} This project has been funded in whole or in part

### 2011-09-23 seqbias-paper.tex
changes

@@
 Because we cannot observe directly the underlying RNA-abundance, our evaluation
-strategy relies on testing two assumptions we make of an ideal, unbiased experiment.
+strategy relies on testing three assumptions we make of an ideal, unbiased experiment.
 \begin{enumerate}
@@
 process.
+\item Adjusting for bias should increase the agreement between RNA-Seq and
+qRT-PCR.
 \end{enumerate}
@@
 
+\subsection{qRT-PCR Correlation}
+
+We used sequencing data previously published by \citet{Au2010} to evaluate the
+effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
+made available by the the Microarray Quality Control project \cite{Shi2006}. The
+RNA-Seq data shows a pattern of bias similar to that seen in the other samples
+sequenced on an Illumina platform.
+%% TODO: refer to supplement for figure of bias
+
+To evaluate the efficacy of each of the bias correction methods considered, we
+counted reads overlapping each gene, defining the gene by the union of every
+transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
+100 additional bases from both ends of the gene. Counts were then normalized
+(divided) by the length of these genes. We then removed any genes with a read
+count less than 10, or that did not correspond to a unique TaqMan probe, leaving
+592 genes.
+
+%% TODO: something goes here
+After adjusting read counts according to the predicted sequence bias, we
+computed the Pearson correlation between log read counts and log TaqMan
+expression values, which are averaged across three replicates.
+
+\begin{figure}
+\centerline{\includegraphics[width=0.4\textwidth]{pcrcor.eps}}
+\caption{
+    Pearson correlation is computed betweer TaqMan values and (bias-corrected)
+    read counts, on a logarithmic scale.
+    %%TODO
+    Error bars show a 95\% confidence interval, estimated using the Fisher
+    transformation, as computed by the R function \texttt{cor.test}.
+}
+\label{fig:pcrcor}
+\end{figure}
+
+%\begin{table}
+%\processtable{
+    %Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
+    %values.\label{tab:pcr}}{
+%\begin{tabular}{ll} \toprule
+%Method & Correlation \\ \midrule
+%Unadjusted &  0.6665 \\
+%7mer &  0.6677 \\
+%GLM & 0.6854 \\
+%MART & 0.6921 \\
+%BN & 0.7016
+%\end{tabular}
+%}
+%{
+%}
+%\end{table}
+
 \section{Discussion}

### 2011-09-24 seqbias-paper.tex
significance tests, minor tweaks

@@
 {
-\begin{tabular}{lllr}\toprule
-Experiment & Species & Platform & Read Length \\\midrule
-\citet{Wetterbom2010} & Chimpanzee & ABI & 33 \\
-Katze (unpublished) & Macaque & ABI & 50 \\
-\citet{Bullard2010} & Human & Illumina & 35 \\
-\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
-\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
+\begin{tabular}{lllcc}\toprule
+ & & & Read & Num. of \\
+Experiment & Species & Platform & length & reads \\\midrule
+\citet{Wetterbom2010} & Chimpanzee & ABI & 33 & $3.9 \times 10^7$ \\
+Katze (unpublished) & Macaque & ABI & 50  & $4.8 \times 10^7$ \\
+\citet{Bullard2010} & Human & Illumina & 35 & $1.2 \times 10^7$ \\
+\citet{Mortazavi2008} & Mouse & Illumina & 33 & $3.9 \times 10^6$ \\
+\citet{Trapnell2010} & Mouse & Illumina & 75 & $6.6 \times 10^7$ \\\botrule
 \end{tabular}
@@
 bias corrected model. The statistic is positive, and increases as uniformity is
-increased, and negative when uniformity is decreased. Boxes are plotted to mark
-the 25\%, 50\%, and 75\% quantiles, with whiskers extending to 1.5 times the
-inter-quartile range (i.e., the span between the 25\% and 75\% quantiles), and
-dots marking more extreme values.
+increased, and negative when uniformity is decreased.  Marked with stars are
+methods over which our approach showed a statistically significant improvement
+when applied to the same data, according to a one-sided Wilcoxon signed-rank
+test. In each of those marked, we observed p-values less than $10^{-23}$.  Boxes
+are plotted to mark the 25\%, 50\%, and 75\% quantiles, with whiskers extending
+to 1.5 times the inter-quartile range (i.e., the span between the 25\% and 75\%
+quantiles), and dots marking more extreme values.
 }
@@
           & BN    & MART  & GLM   & 7mer \\\midrule
-Wetterbom & 0.174 & 0.016 & 0.066 & -0.079 \\
-Katze     & 0.280 & 0.243 & 0.158 & 0.033 \\
-Bullard   & 0.267 & 0.163 & 0.224 & 0.157 \\
-Mortazavi & 0.240 & 0.210 & 0.197 & 0.091 \\
-Trapnell  & 0.289 & 0.289 & 0.248 & 0.138
+Wetterbom & \textbf{0.174} & 0.016 & 0.066 & -0.079 \\
+Katze     & \textbf{0.280} & 0.243 & 0.158 & 0.033 \\
+Bullard   & \textbf{0.267} & 0.163 & 0.224 & 0.157 \\
+Mortazavi & \textbf{0.240} & 0.210 & 0.197 & 0.091 \\
+Trapnell  & \textbf{0.289} & \textbf{0.289} & 0.248 & 0.138
     \end{tabular}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{pcrcor.eps}}
-\caption{
-    Pearson correlation is computed betweer TaqMan values and (bias-corrected)
-    read counts, on a logarithmic scale.
-    %%TODO
-    Error bars show a 95\% confidence interval, estimated using the Fisher
-    transformation, as computed by the R function \texttt{cor.test}.
+%\begin{figure}
+%\centerline{\includegraphics[width=0.4\textwidth]{pcrcor.eps}}
+%\caption{
+    %Pearson correlation is computed betweer TaqMan values and (bias-corrected)
+    %read counts, on a logarithmic scale.
+    %%%TODO
+    %Error bars show a 95\% confidence interval, estimated using the Fisher
+    %transformation, as computed by the R function \texttt{cor.test}.
+%}
+%\label{fig:pcrcor}
+%\end{figure}
+
+\begin{table}
+
+\processtable{
+    Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
+    values.\label{tab:pcr}}{
+    \centerline{
+\begin{tabular*}{0.4\textwidth}{lrl}
+\toprule
+Method & Correlation & \\ \midrule
+Unadjusted &  0.6665 & $\ast\ast$ \\
+7mer &  0.6677 & $\ast\ast$ \\
+GLM & 0.6854 & $\ast\ast$ \\
+MART & 0.6921 & $\ast$ \\
+BN & \textbf{0.7016}
+\end{tabular*}}
+
 }
-\label{fig:pcrcor}
-\end{figure}
-
-%\begin{table}
-%\processtable{
-    %Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
-    %values.\label{tab:pcr}}{
-%\begin{tabular}{ll} \toprule
-%Method & Correlation \\ \midrule
-%Unadjusted &  0.6665 \\
-%7mer &  0.6677 \\
-%GLM & 0.6854 \\
-%MART & 0.6921 \\
-%BN & 0.7016
-%\end{tabular}
-%}
-%{
-%}
-%\end{table}
+{
+We estimated the statistical significance of the improvement in correlation
+using the BN method over the other methods using a simple boostrap procedure. A
+bootstrap sample is formed by sampling, with replacement, 592 genes from the
+original set of the same size. The correlation is then computed for this set,
+using the adjusted count from each method. We repeated this procedure one
+million times, and counted the number of times each of the competing method achieves a
+higher correlation than the BN method.  Those marked with a sigle asterisk
+achieved a higher correlation fewer than 1000 times, resulting in a p-value of
+less than $10^{-3}$ Those marked with two asterisks achieved a higher
+correlation in none of the bootstrap samples, indicating a p-value less than
+$10^{-6}$.
+}
+\end{table}
 

### 2011-09-24 results/plots/plot-pois.R

@@ -24,12 +24,42 @@ ys <- subset(xs, method == 'Unadjusted')[,c('sample', 'gene', 'L')]
 xs <- merge(xs, ys, by = c('sample', 'gene'))
 names(xs) <- c('sample', 'gene', 'method', 'mu', 'L', 'L.base')
 
+xs$sample <- factor(tolower(xs$sample))
+
+
+# read test regions (so we no how many positions were included the
+# log-likelihood calculation)
+ks.all <- NULL
+for (w in c('bullard', 'katze', 'mortazavi', 'trapnell', 'wetterbom')) {
+    fn <- paste('../exons/', w, '01.test.bed', sep = '')
+    ks <- read.table(fn, header = F)
+    names(ks) <- c('seqname', 'start', 'end', 'name', 'score', 'strand')
+    ks$gene   <- factor(as.character(0:(length(ks[,1]) - 1)))
+    ks$sample <- w
+    ks$k      <- ks$end - ks$start
+    ks <- ks[,c('sample', 'gene', 'k')]
+    ks.all <- rbind(ks.all, ks)
+}
+
+xs <- merge(xs, ks.all, by = c('sample', 'gene'))
 
 
 # mcfadden's pseudo-R2
-r2 <- function(L, L.base) { 1 - L / L.base }
+r2.mcfadden <- function(L, L.base, n) { (1 - L / L.base) }
+
+# nagelkerke's R2
+r2.nagelkerke <- function(L, L.base, n) { 1 - exp( (2/n) * (L.base - L)) }
+
+r2 <- r2.mcfadden
+#r2 <- r2.nagelkerke
 
 
+# reorder
+xs$method <- factor(xs$method,
+                    levels = c('Hansen/A', 'Li/GLM', 'Li/MART', 'Jones'))
+
+xs$r2 <- r2(xs$L, xs$L.base, xs$k)
+med <- tapply(xs$r2, INDEX = list(xs$sample, xs$method), FUN = median)
 
 
 # the boxplots are plotted seperately, then combined, so that they can be put no
@@ -40,32 +70,24 @@ library(ggplot2)
 source('theme_dcjstd.R')
 
 
-#CairoPNG('pois-boxplot-simple.png',
+#cairopng('pois-boxplot-simple.png',
          #width = 600, height = 2000)
 
-#p <- qplot(data = xs, y = r2(L, L.base), x = method, geom = 'boxplot')
+#p <- qplot(data = xs, y = r2(l, l.base), x = method, geom = 'boxplot')
[…]

### 2011-09-25 seqbias-paper.tex
combined two figures

@@
 \begin{figure}
-\centerline{\includegraphics[width=0.25\textwidth]{overview.eps}}
+\centerline{\includegraphics[width=0.4\textwidth]{overview.eps}}
 \caption{An overview of the approach taken: (a) foreground sequences are sampled
@@
 sampled foreground and background sequences, (d) and the model is evaluated
-at each position within a locus, predicting bias.}
+at each position within a locus, predicting bias. The predicted bias can then be
+used to adjust read counts, as in (e). In (d) and (e) we show the results of
+this method applied to the $3'$ UTR of Apoa2, using data from
+\citet{Mortazavi2008}.}
 \label{fig:overview}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{uniformity.eps}}
-\caption{Read counts (i.e., the number of reads starting at each position)
-from the Mortazavi data set are plotted along the $3'$ UTR of Apoa2. The upper
-plot shows unadjusted counts, the lower, counts adjusted by our method.}
-\label{fig:unif}
-\end{figure}
-
 In this comparison, we measure how well the counts conform to a Poisson process.
@@
 interpreted criterion. This increase in uniformity is illustrated in Figure
-\ref{fig:unif}.
+\ref{fig:overview}.
 
@@
 effect bias correction has on correlation to measurements made by TaqMan RT-PCR,
-made available by the the Microarray Quality Control project \cite{Shi2006}. The
+made available by the the Microarray Quality Control project \citep{Shi2006}. The
 RNA-Seq data shows a pattern of bias similar to that seen in the other samples

### 2011-09-25 seqbias-paper.tex
description of qpcr analysis

@@
 
-We applied our procedure (labeled ``BN'') as well as those of
-\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
-are implemented in the R packages \texttt{mseq} and \texttt{Genominator},
-respectively. The procedures were applied to four publicly available data
-sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
-an unpublished data set of our own (Table \ref{tab:datasets}).
+To evaluate the first two assumption, we applied our procedure (labeled ``BN'')
+as well as those of \citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010}
+(``7mer''), which are implemented in the R packages \texttt{mseq} and
+\texttt{Genominator}, respectively, to four publicly available data sets
+\citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as an
+unpublished data set of our own (Table \ref{tab:datasets}).
 
@@
 chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
-of long, highly expressed exons from the remaining chromosomes. In particular, for
-each reference sequence, beginning with the set of exons annotated by Ensembl
-release 60 \cite{Hubbard2009}, we removed any exons with known alternate splice
-sites, then chose the top 1000 exons by read count, restricting ourselves those
-at least 100 nt long.
+of long, highly expressed exons from the remaining chromosomes. In particular,
+for each reference sequence, beginning with the set of exons annotated by
+Ensembl release 60 \cite{Hubbard2009}, we removed any exons with known alternate
+splice sites, then chose the top 1000 exons by read count, restricting ourselves
+to those at least 100 nt long.
 
@@
 All data sets were mapped using Bowtie \citep{Langmead2009} against,
-respectively, the hg19, mm9, and panTro2 genome assemblies obtained from the
-UCSC Genome Browser \citep{Karolchik2008}.
+respectively, the hg19, mm9, rheMac2, and panTro2 genome assemblies obtained
+from the UCSC Genome Browser \citep{Karolchik2008}.
 
@@
 RNA-Seq data shows a pattern of bias similar to that seen in the other samples
-sequenced on an Illumina platform.
-%% TODO: refer to supplement for figure of bias
+sequenced on an Illumina platform (see, Supplementary Section 5).
 
@@
 transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
-100 additional bases from both ends of the gene. Counts were then normalized
-(divided) by the length of these genes. We then removed any genes with a read
-count less than 10, or that did not correspond to a unique TaqMan probe, leaving
-592 genes.
+100 additional bases from both ends of the gene. Counts were then normalized by
+dividing by the length of these genes. We then removed any genes with a read
+count less than 10, or that did not correspond to
 
-%% TODO: something goes here
-After adjusting read counts according to the predicted sequence bias, we
-computed the Pearson correlation between log read counts and log TaqMan
-expression values, which are averaged across three replicates.
-
-%\begin{figure}
-%\centerline{\includegraphics[width=0.4\textwidth]{pcrcor.eps}}
-%\caption{
-    %Pearson correlation is computed betweer TaqMan values and (bias-corrected)
-    %read counts, on a logarithmic scale.
-    %%%TODO
-    %Error bars show a 95\% confidence interval, estimated using the Fisher
-    %transformation, as computed by the R function \texttt{cor.test}.
-%}
-%\label{fig:pcrcor}
-%\end{figure}
+Each method was trained in a manner identical to that used in the analysis of
+section 3.1 and 3.2, but without restricting the training data to the first
+eight chromosomes.  After adjusting read counts according to the predicted
+sequence bias, we computed the Pearson correlation between log read counts and
+log TaqMan expression values, which are averaged across three replicates. These
+correlations are listed in Table \ref{tab:pcr}.
 
 \begin{table}
-
 \processtable{
@@
 \end{tabular*}}
-
 }
@@
 higher correlation than the BN method.  Those marked with a sigle asterisk
-achieved a higher correlation fewer than 1000 times, resulting in a p-value of
-less than $10^{-3}$ Those marked with two asterisks achieved a higher
+achieved a higher correlation fewer than 1000 times, resulting in a p-value
+less than $10^{-3}$.  Those marked with two asterisks achieved a higher
 correlation in none of the bootstrap samples, indicating a p-value less than

### 2011-09-27 seqbias-paper.tex
minor corrections from r's notes

@@
 Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
-often conflated by protocol-specific sequence bias, influenced by PCR
-amplification, or differing primer affinities and mixtures, for example.  The
-result is decreased accuracy in many applications, such as de novo gene
-annotation and isoform quantification in RNA-Seq, and peak calling in
-ChIP-Seq.
+often conflated by protocol-specific sequence bias. The sources of the bias are
+unknown, but may influenced by PCR amplification, or differing primer affinities
+and mixtures, for example.  The result is decreased accuracy in many
+applications, such as de novo gene annotation and isoform quantification in
+RNA-Seq, and peak calling in ChIP-Seq.
 
@@
 are sampled uniformly from transcripts, each of the plots should be essentially
-flat}
+flat.}
 \label{fig:freqs}
@@
 The experiment may be considered unbiased  with regards to sequence if, having
-observed the nucleotide sequence $s_i$ surrounding position $i$,
+observed the nucleotide sequence $s_i$ surrounding position $i$, the expected
+number of reads sampled from position $i$ is independent of $s_i$, i.e., if
+
 $$ E[ x_i | s_i ] = N \Pr[ m_i | s_i ] = N \Pr[ m_i ] = E[ x_i ] $$
-
 From Bayes' rule,
 $$ \Pr[ m_i | s_i ] = \frac{ \Pr[ s_i | m_i ] \Pr[ m_i ] }{ \Pr[ s_i ] } $$
-
 This suggests a natural scheme in which observations may be reweighted to
@@
 The Bayesian network can then be used to evaluate sequence probability, and thus
-bias, at any genomic position. Figure \ref{fig:overview} gives a brief overview
+bias, at any genomic position. Figure \ref{fig:overview} gives a schematic overview
 of the proposed model.
@@
 
-However, we have so for ignored one complication: the RNA abundance that we wish
+However, we have so far ignored one complication: the RNA abundance that we wish
 to estimate is not itself independent of the nucleotide sequence. Notably,
@@
 
-The methods proposed by \citet{Hansen2010} and \cite{Roberts2011} also treat bias
-correction as a problem of estimating foreground and background sequence
+The methods proposed by \citet{Hansen2010} and \cite{Roberts2011} also treat
+bias correction as a problem of estimating foreground and background sequence
 probabilities. They differ primarily in how these sequence probabilities are
 estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
-notation) directly, given training data consisting of long, highly expressed
-transcripts.
+notation) directly, given training data consisting of long annotated, highly
+expressed transcripts.
 
@@
 sequences, we take sequences surrounding (extending 20 nt to either side, by
-default) the starts positions of $n/2$ aligned reads. We wish to avoid
-overfitting to any short locus with a large abundance of reads, yet we must also
-capture the bias introduced by, for example, PCR amplification. We therefore
-choose the $n/2$ unique reads with the highest number of duplicates, yet count
-each read only once, ignoring duplicates
+default) the start positions of a randomly sampled set of $n/2$ aligned reads,
+ignoring duplicate reads.
 
@@
 where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
-0.5$) if we sample foreground and background positions equally.
+0.5$) since we sample foreground and background positions equally.
 
@@
 
-Some benefit would be obtained from a more highly tuned complexity penalty.
+Some benefit might be obtained from a more highly tuned complexity penalty.
 However, since the model is trained greedily, additional parameters will be
@@
 We perform maximum-likelihood fitting of two models. In the null model, the
-Poisson rate is fixed for each exon. That is, for position $j$ within exon $i$,
+Poisson rate is fixed for each test exon. That is, for position $j$ within exon $i$,
 the rate is $ \lambda_{ij} = a_i $ where $a_i$ is the parameter being fit. For
@@
 bias corrected model. The statistic is positive, and increases as uniformity is
-increased, and negative when uniformity is decreased.  Marked with stars are
-methods over which our approach showed a statistically significant improvement
+increased, and negative when uniformity is decreased.  Marked with asterisks are
+methods over which the BN approach showed a statistically significant improvement
 when applied to the same data, according to a one-sided Wilcoxon signed-rank
@@
 using an estimation of sequence probability that is more robust to overfitting
-and can account for bias beyond the initial heptamer. In all our tests, this
+and can account for bias outside of the initial heptamer. In all our tests, this
 approach was at least as effective as those of \citet{Li2010}, despite not
@@
 \ref{fig:models}) are considerably different between those sequenced on an
-Illumina platform versus an ABI platform.
+Illumina platform versus an ABI platform, and even vary within platform.
 

### 2011-09-27 results/plots/plot-dg.py

@@ -173,7 +173,10 @@ def compute_num_params(G):
         name = e.get_destination()
         node_params[name] *= 4
 
-    return sum(node_params.values())
+    for name in node_params:
+        node_params[name] -= 1
+
+    return 2 * sum(node_params.values())
 
 
 def main():

### 2011-09-28 seqbias-supplement.tex
additions to the response letter
significance tests, minor tweaks
description of qpcr analysis
moste of r's notes on kl divergence

--- 2011-08-23 seqbias-supplement.tex
@@
 \title{SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data}
-\author{Daniel C. Jones}
+\author{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}
 
@@
 \dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{Correcting for Bias in High-Throughput
-Sequencing Data}}{Daniel C. Jones}
+Sequencing Data}}{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}
 
@@
 \end{center}
-
-\section{Additional Kullback-Leibler Divergence Plots}
-
-Below we plot the adjusted and unadjusted KL divergence plots, supplementing
-those in the results section of the main paper.
-
-\includegraphics[width=\textwidth]{fig/kl-all.pdf}
-
-\section{Runtime-Accuracy Trade-off}
-
-Training our model on more reads results in more accurate estimation of bias,
-but a longer training time. Here we investigate that trade-off by training our
-model on progressively more reads from the Mortazavi data set. We use the
-median pseudo-coefficient of determination ($R^2$) over exons selected for our
-test set. Each point is additionally labeled with the time used to train the
-model on one core of a 3Ghz Intel Xeon processor.
-
-\includegraphics[width=\textwidth]{fig/pois-n.eps}
 
@@
 solution.
+
+\section{Additional Kullback-Leibler Divergence Plots}
+
+Below we plot the adjusted and unadjusted KL divergence plots, supplementing
+those in the results section of the main paper.
+
+\includegraphics[width=\textwidth]{fig/kl-all.pdf}
+
+\section{Sequence bias in data from Au, et al., 2010}
+
+\centerline{\includegraphics[width=0.6\textwidth]{fig/au-freqs.pdf}}
+
+\section{Runtime-Accuracy Trade-off}
+
+Training our model on more reads results in more accurate estimation of bias,
+but a longer training time. Here we investigate that trade-off by training our
+model on progressively more reads from the Mortazavi data set. We use the
+median pseudo-coefficient of determination ($R^2$) over exons selected for our
+test set. Each point is additionally labeled with the time used to train the
+model on one core of a 3Ghz Intel Xeon processor.
+
+\includegraphics[width=\textwidth]{fig/pois-n.eps}
 
@@
 
+\section{KL Divergence}
+
+\subsection{TODO: Title}
+
+Suppose $P$ and $Q$ are probability distributions defined on a discrete sample
+space. E.g., in the context of this paper, think of them as the probabilities of
+DNA $k$-mers for some fixed $k$. The \emph{Kullback-Leibler divergence}, also
+known as \emph{relative entropy}, of $Q$ with respect to $P$ is defined as
+$$H(Q||P) = \sum_{i} q_i \ln \frac{q_{i}}{p_{i}}$$
+where $q_{i}$ ($p_{i}$) is the probability of observing the i$^{\text{th}}$
+event according to the distribution $Q$ (resp., $P$), and the summation is taken
+over all events in the sample space (e.g., all $k$-mers). In some sense, this is
+a measure of the dissimilarity between the distributions: if $p_{i} \approx
+q_{i}$ everywhere, their log rations will be near zero and $H$ will be small; as
+$q_{i}$ and $p_{i}$ diverge, their log rotions will deviate from zero and $H$
+will increase.
+
+For a more quantitative and (perhaps) intuitive interpretation, consider the
+following hypothesis-testing scenario. Given $m$ independent samples from either
+$P$ (arbitrarily called the ``null'') or from $Q$ (the ``alternative''), how
+large should $m$ be to cenfidently choose between these cases? A natural
+approach is to use a likelihood ratio test: letting $Y_i$ be the number of times
+event $i$ is observed in $m$ trials ($\sum Y_i = m$), the likelihood of the
+sequence of observations under model $P$ is $$\prod_{i} p_{i}^{Y_{i}}$$
+and similarily for $Q$. So, the logarithm of the likelihood ratios is
+$$
+\text{LLR} =
+\ln \frac{\prod_{i} q_{i} ^ {Y_{i}}}{\prod_{i} p_{i} ^ {Y_{i}}}
+= \sum_{i} Y_{i} \ln \frac{q_{i}}{p_{i}}
+= m \sum_{i} \frac{Y_{i}}{m} \ln \frac{q_{i}}{p_{i}} $$
+
+If the sample is drawn from the alternative distribution $Q$, then the
+expectation of $frac{Y_{i}}{m}$ is $q_i$, so the expected value of the LLR is
+$$ m \sum_{i} q_{i} \ln \frac{q_{i}}{p_{i}} = m H(Q || P) $$
+That is, the KL divergence is exactly the expected per-sample contribution to
+the log-likelihood ratio.
+%% when comparing these two hypotheses ??
+So, assuming the null hypothesis is false, in order for it to be rejected with
+$1000:1$ odds, one should choose $m$ to be inversely proportional to $H(Q ||
+P)$:
+\begin{align*}
+m H(Q||P) \ge \ln 1000 \\
+m \ge \frac{\ln 1000}{H(Q || P)}
+\end{align*}
+
+As a concrete example, all of the RNA-Seq data sets examined in this paper show
+a KL divergence between the 1$^{\text{st}}$ positon of reads versus the
+transcriptomic background nucleotide distribution of $0.05$ or greater. Thus,
+one can confidently reject the hypothesis that reads are beeing uniformly
+sampled across the transcriptome by examining only a few hundred randomly
+selected reads (i.e., $\ln 1000 / 0.05 \approx 140$).
+
+\subsection{Accuracy of Multinomial Parameter Estimation}
+
+Continuing the notation above, suppose $P$ as an unknown distribution with
+parameters $p_1, \dots, p_r$, $\sum p_i = 1$ where $r$ is the number of points
+in the sample space (e.g. $r = 4^{k}$ in the case of $k$-mers). Given a random
+ample $X_1, X_2, \dots, X_r$ of size $n = \sum_{i} X_i$ from $P$, it is well
+known that the maximum likelihood estimators for the parameters are $q_i =
+\frac{X_i}{n} \approx p_i$. How good an estimate for $P$ is this distribution
+$Q$?
+
+$$E[q_i] = E\left[\frac{X_i}{n}\right] = \frac{E[X_i]}{n} = \frac{np_i}{n} =
+p_i$$
+And the standard deviation of each estimate is proportional to $1/\sqrt{n}$. So
+these estimates are increasingly accurate as the sample size increases. A more
+quantitative assessment of the accuracy of the estimator is obtained by
+evaluating the KL divergence:
+$$H(Q||P)
+= \sum_{i = 1}^{r} q_i \ln \frac{q_i}{p_i}
+= \sum_{i = 1}^{r} q_i \ln \left(1 + \frac{q_i - p_i}{p_i} \right) $$
+Using the first two terms of he Taylor series for $\ln (1 + x)$, this is
+\begin{align*}
+H(Q||P) &\approx \sum_{i = 1}^{r} q_i \left( \frac{q_i - p_i}{p_i} -
+\frac{1}{2} \left( \frac{q_i - p_i}{p_i} \right)^2 \right ) \\
+&= \sum_{i = 1}^{r} q_i \frac{q_i - p_i}{p_i} -
+\frac{q_i}{2 p_i} \frac{(q_i - p_i)^2} {p_i}
+\end{align*}
+Since $\sum_{i = 1}^{r} q_i = \sum_{i = 1}^{r} pi = 1$,
+$\sum_{i = 1}^{r} p_i \frac{q_i - p_i}{p_i} = 0$, so
+\begin{align*}
+H(Q||P) &\approx \sum_{i=1}^{r} q_i \frac{q_i - p_i}{p_i} - p_i \frac{q_i -
+p_i}{p_i} - \frac{q_i}{2 p_i} \frac{(q_i - p_i)^2}{p_i} \\
+&= \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}\left(1 - \frac{q_i}{2 p_i}\right) \\
+&\approx \frac{1}{2} \sum_{i=1}^{r} \frac{(q_i - p_i)^2}{p_i}
+\end{align*}
+since $q_i \approx p_i$. Multiplying by $n^2 / n^2$ we have,
+\begin{align*}
+H(Q||P) &\approx \frac{1}{2n} \sum_{i=1}^{r} \frac{n q_i - n p_i)^2}{n p_i} \\
+        &= \frac{1}{2n} \sum_{i=1}^{r} \frac{(X_i - E[X_i])^2}{E[X_i]}
+\end{align*}
+
+The summation is the test statistic for the $\chi^2$ goodness-of-fit test for a
+multinomial distribution, and as $n \rightarrow \infty$ is known to follow a
+$\chi^2$ distribution with $r - 1$ degrees of freedom. Finally, the expected
+value of such a random variable is $r - 1$, hence the expected KL divergence of
+the MLE inferred distribution $Q$ with respect to the true distribution $P$ is
+$$E[H(Q||P)] = \frac{r - 1}{2n}$$
+
+Stochastic simulations of this proves when $P$ is uniform show that the above
+formula is a very good fit when $n > r$.
+%% TODO: see figure X
+
 \bibliographystyle{plain}

### 2011-09-29 response/response.tex
more revisions

@@
 \usepackage{amsmath}
-%\usepackage{dcj}
+\usepackage{dcj}
 
@@
 submitted on July 8. Upon review, the second of the two reviewers raised some
-concerns, compelling its rejection. However, we believe that the bulk of these
+concerns leading to its rejection. However, we believe that the bulk of these
 concerns are either incongruous with the content of the manuscript or
@@
 
-The second reviewer raised three major concerns. The first concern raised is
+The second reviewer raised three major concerns. We summarize our
+responses to these issues here.  The first concern raised is
 that the method we present is unjustifiably complex, involving a ``plethora of
 tuning parameters'', and that the much simpler method of Hansen, et al., would
-perform as well is trained on enough data.  In the data we presented, the simple
+perform as well if trained on enough data.  In the data we presented, the simple
 method developed by Hansen, et al., performed poorly regardless of the number of
@@
 being more effective, also offers a significant simplification.
+Additionally, we believe the structure of our models give some
+theoretical insight into these differences in efficacy.
 
 Secondly, it is asserted that we did not demonstrate that uniformity of read
-coverage is increased ofter correcting for bias using our method.  Yet, in
+coverage is increased after correcting for bias using our method.  Yet, in
 section 3.2 we had presented a rigorous statistical analysis over a large and
 diverse collection of data sets showing that this is indeed the case. This
-section of the paper was evidently overlooked by the reviewer, as it received
+section of the paper was perhaps overlooked by the reviewer, as it received
 no mention in this criticism.
@@
 Under these circumstances we respectively request that the manuscript, in this
-revised form, be reconsidered for publication.
+revised form, be reconsidered for publication.  Detailed responses to
+the reviewers concerns are presented below.
 
@@
 \begin{quote}
-(1) Page3, 2.2.1 mentions a correction for bias introduced by PCR amplification.
+(1) Page 3, 2.2.1 mentions a correction for bias introduced by PCR amplification.
 RNA sequencing methods are available where PCR amplification is not used.
@@
 
-Because model complexity is penalized, a very sparse, or even empty model will
-be trained on a data set without bias.  If PCR amplification is responsible for
-a majority of the bias, and an RNA-Seq method without PCR amplification is used
-producing largely bias-free data, then using the method to correct for bias
-will have little if any effect.
+First, we emphasize that PCR bias is only one of a number of technical steps
+involved in sequencing that potentially introduce bias.  Our method is agnostic
+to the source of bias, and because model complexity is penalized, a very sparse,
+or even empty model will be trained on a data set without bias.  Thus, if PCR
+amplification is responsible for a majority of the bias, and an RNA-Seq method
+without PCR amplification is used producing largely bias-free data, then using
+the method to correct for bias will have little if any effect.
 
-We have included new data examining the FRT-Seq protocol of Mamanova, et. al.,
-in which no PCR-amplification in performed during library preparation. We show
-that bias is reduced, but not eliminated in this method, and that our method
-accounts for this reduced bias by constructing a much sparser model, relative to
-the other RNA-Seq data sets we have examined.
+As a concrete example, we have included new data examining the FRT-Seq protocol
+of Mamanova, et al., in which no PCR-amplification is performed during library
+preparation. We observe that bias is still present in this data set, albeit at
+lower levels than other uncorrected RNAseq data examined.  Our method accounts
+for this reduced bias by constructing a much sparser model, relative to the
+other RNA-Seq data sets we have examined.
+
+To further explore the point raised, we repeatedly trained our model on sets of
+10,000 random simulated reads, drawn uniformly from within exons. We repeated
+this 100 times, and in each trial an empty model was trained.  Reducing the
+number of reads increases the risk of a spuriously non-empty model, but
+repeating this evaluating with only 100 reads led to only 3 out of 100 models
+being non-empty. This is indicative that the default BIC scoring, as predicted
+by statistical theory, is adequately protecting against false discovery of
+``bias'' when there is none.
 
@@
 
-The method has precisely the following parameters,
+The method has precisely the following six parameters,
 \begin{enumerate}
@@
 needed to train the algorithm. Each has the same very simple interpretation: a
-larger value leads to solution that is no worse and potentially better, at the
-cost of increased training time.  We mention these parameters in the manuscript
-only for the sake of transparency.  They generally require no intervention or
-tuning, being essentially an implementation detail.
+larger value leads to a potentially better solution, at the cost of increased
+training time.
+
+We mention these parameters in the manuscript only for the sake of transparency.
+They generally require no intervention or tuning, being essentially an
+implementation detail.
 
@@
 diverse collection of data sets. In this revision we have included new data
-showing explicitly that the quality of the solution is remarkably insensitive to
-values of these two parameters.
+(Supplementary Section 2) showing explicitly that the quality of the solution is
+remarkably insensitive to values of these two parameters over a range spanning 4
+orders of magnitude.
 
 It should also be noted that, while the brute-force method of Hansen, et. al.,
-has fewer parameters, the other methods evaluated in the manuscript (the GLM and
-MART methods of Li, et. al.), involve more parameters that the method we
-have described. (For example training the MART method involves parameters
+has fewer tuning parameters, the other methods evaluated in the manuscript (the
+GLM and MART methods of Li, et. al.), involve more parameters than the method we
+have described. (For example, training the MART method involves parameters
 controlling the complexity penalty, the number of trees, the lower bound
 threshold for read counts, in addition to others that are not described or
-easily changed from the values divined by the authors.) Our method therefore
-presents a simplification over competing methods.
+easily changed from the values divined by the authors.  Additionally, their
+method requires selection of a set of annotated genes for training.)  Our method
+therefore presents a simplification over competing methods.
 
@@
 
-We regret that we more not more explicit on this point. At least 600,000 reads
-on each data set, but as many as 23 million were used. (As noted in the paper,
-we did reimplement the Genominator method in order to do this.) The brute-force
-method performed very poorly on every data set, suggesting that simply throwing
-more data at a suboptimal solution will not adequately solve the problem. The
-results as they were presented are unambiguous: the other methods perform far
-better with far fewer reads.
+We regret that we were not more explicit on this point. At least
+600,000 reads on each data set, but as many as 23 million were used:
+see the revised table X. (As noted in the paper, we did reimplement
+the Genominator method in order to do this.)
 
-Moreover, even if the brute-force method were effective if trained with many
-hundreds of millions of reads, many (possibly most) existing data sets are not
-this large, rendering it generally impractical.
+We do not wish to ``pick on'' Hansen, et al.  Theirs was a pioneering
+paper and presents a simple, intuitive approach for tackling the
+problem they identified.  Nevertheless, their brute-force method
+performed poorly on every data set we tested, even the largest ones,
+suggesting that simply throwing more data at a suboptimal solution
+will not adequately solve the problem.  In our opinion, the results we
+presented were unambiguous: the other methods perform far better with
+far fewer reads.
+
+%Moreover, even if the brute-force method were effective if trained
+%with many hundreds of millions of reads, many (possibly most) existing
+%data sets are not this large, rendering it generally impractical.
+
+Understanding this failure of a simple, appealing method is important
+for making progress, and we feel that our paper provides a theoretical
+basis for understanding these differences in efficacy, and we have
+amplified these points in the revision.  Intuitively, one expects
+model accuracy to increase as the number $n$ of training reads
+increases, but to decrease as the number $r$ of model parameters
+grows, since the training data is ``spread more thinly.''  For the
+particular case of estimating a multinomial distribution (as in
+Hansen), Supplementary Section YY quantifies this intuition very
+precisely, showing that the expected KL divergence of the learned
+model with respect to the true distribution converges to $(r-1)/(2n)$,
+where $r=4^7=16384$ is the number of model parameters learned.  The
+same analysis does not directly apply to the MART, GLM or BN models,
+but we expect their behavior to be qualitatively similar.  Thus, to
+the extent that a more parsimonious model such as ours adequately
+captures the structure of the data, it is potentially orders of
+magnitude more accurate for a given quantity of training data, simply
+by virtue of having only a few hundred parameters learned from counts
+(Figure 3), as opposed to tens of thousands.  [KL divergence between
+two distributions is inversely proportional to the number of samples
+one needs to confidently test the hypothesis that the samples arise
+from one distribution versus the other; see Supplementary Section XX.]
+%% TODO: why the brackets here?
+
+Of course, training data is plentiful in NGS data sets, so,
+pragmatically, the above analysis probably identifies a contributor
+but not the root cause of Hansen et al.'s relatively poor performance
+in our experiments.  Rather, we believe the problem is that bias is
+not confined to the initial heptamer.  The model structures
+illustrated in Figure 3 show that nucleotides well outside that region
+are often biased.  (The randomization study sketched above shows that
+our model building algorithm is not prone to spurious discovery of
+overly-complex models.)  Furthermore, these examples show that the
+read sampling is often biased by positions \emph{in front of the
+  read}, which are not currently considered by Hansen.  Their method
+could be modified easily to consider $k$-mers spanning the read start,
+but the entropy analysis shows that extending it to $k \gg 7$ is
+infeasible, since the exponential growth in $r=4^k$ will destroy the
+model's accuracy even in large data sets.  E.g., $k \ge 20$ would be
+necessary to cover the biased regions in all data sets shown in Figure
+3, necessitating trillions of training reads to build an adequate
+model by Hansen's method.
 
@@
 - Does this unevenness go away with the proposed adjustment?
+\end{quote}
+
+During preparation we chose not to include a figure such as the one suggested
+by the reviewer as we believed it would not be legitimately informative. It
+would be far too easy to cherry-pick a miniscule genomic interval that makes
+even a very shoddy method appear favorable.
+
+Instead, we presented data (in Section 3.2) that answers the
+reviewer's question (i.e., ``does this unevenness go away with the
+proposed adjustment?'') in rigorous and comprehensive manner. We
+implemented a Poisson regression metric, which quantitatively measures
+``uniformity'' by calculating the likelihood that the reads sampled
+across a given exon were sampled from a uniform distribution.  Using
+this, we performed a large cross-validation study proving the
+effectiveness of our method at increasing uniformity across thousands
+of exons drawn from a diverse group of data sets, rather than
+qualitatively across one exon.  In broad terms, our method increases
+likelihood of the uniform model by XXX-fold across thousands of exons
+(Figure 6).
+
+In this revision, as suggested, we have revised Figure 2 to show an
+example of this increase in uniformity.  We believe the figure is a
+nice, graphic illustration of the benefits of our method and thank the
+reviewer for causing us to reconsider it.  We stress, however, that
+this figure represents a minute subset of the analysis that had
+previously been presented, and is included only as an illustration.
+The results in sections 3.1 and especially in 3.2 are far more
+comprehensive.
+
+\begin{quote}
 - Also, as is evident from that figure, for some of the experimental
@@
 
-During preparation we choose not to include a figure such as the one suggested
-by the reviewer as we believed it would not be legitimately informative. It
-would be far too easy to cherry-pick a minuscule genomic interval that makes
-even the very shoddy method appear favorable.
+%% TODO
 
-Instead, we presented data (in Section 3.2) that answers the reviewers question
-(i.e., ``does this unevenness go away with the proposed adjustment?'') in
-rigorous and comprehensive manner. We implemented a Poisson regression metric
-and performed a large cross-validation study proving the effectiveness of our
-method at increasing uniformity quantitatively, across thousands of exons drawn
-from a diverse group of data sets, rather than qualitatively across one exon.
-
-In this revision, we have included a figure, as suggested, showing an example of
-this increase in uniformity. Yet, we must stress, this figure represents a
-minute, unreliable subset of the analysis that had previously been presented,
-and is included only as an illustration.
+%The revised Figure 2 also graphically illustrates the range of
+%adjustment factors, which genome-wide are very cool and we hope we can say something soothing about.......
 
@@
 
-Our motivation for adopting a Bayesian network model is the very natural
-probabilistic interpretation that we have described in Section 2.1. We
-disagree with the reviewer's opinion that Bayesian networks are inherently more
-complicated than Lasso regression or support vector machines.  In this
-application, the latter two methods would generate output that is arguably less
-transparent and interpretable, and neither be significantly simpler to
-implement, nor involve fewer parameters.
+Our motivation for adopting a Bayesian network model is the very
+natural probabilistic interpretation that we have described in Section
+2.1. We disagree with the reviewer's opinion that Bayesian networks
+are inherently more complicated than Lasso regression or support
+vector machines.  In this application, the latter two methods would
+generate output that is arguably less transparent and interpretable,
+and would neither be significantly simpler to implement, nor involve
+fewer parameters.  Our hypotheses (above) concerning the reasons for
+the weakness of Hansen's method illustrate the benefit of this
+approach; we do not believe svm regression, say, would have yielded
+similar insights.
 
@@
 
+%\subsection*{Other Changes}
+
+%We have modified the title to more clearly acknowledge the existence of
+%the prior work upon which we build.
+
+%...more...?
+
+\subsection*{Conclusion}
+In summary, we believe the revised manuscript addresses all concerns
+raised by the reviewers, and provides a substantive improvement to the
+state of the art for an important problem.  We respectfully request
+that the manuscript, in this revised form, be reconsidered for
+publication.
+
 \end{document}

### 2011-09-29 seqbias-paper.tex
more revisions

@@
 
-The run time of the training procedure is further reduced in practice by imposing the
-following two restrictions on the structure of the network,
+The run time of the training procedure is further reduced in practice by
+imposing the following two restrictions on the structure of the network,
 \begin{enumerate}
@@
 positions surrounding the read start to consider. Increasing any of these
-parameters results in a solution that is no worse, and potentially better, at
+parameters results in a  solution that is potentially better, at
 the expense of increased training time.
@@
 deviation at which background sequences are sampled, and the weight applied to
-the penalty term of the BIC, yet it is not particularly sensitive to their values
-(see, Supplementary Section 2). In all our evaluation, no special tuning of the
+the penalty term of the BIC, yet it is not particularly sensitive to their
+values. (The median $R^2$ goodness-of-fit statistic used in section 3.2 varied
+by less that 25\% as these parameters were varied over a range of $10^4$. See
+Supplementary Section 2.) In all our evaluation, no special tuning of the
 parameters was performed, suggesting it can be used effectively across data sets
@@
 chosen in a manner identical to that which was used to select the test exons.
-Both the GLM and MART models were trained using the default parameters.
+Both the GLM and MART models were trained considering the initial read position
+and 20 nucleotides upstream and downsream, and otherwise using default
+parameters.
 
@@
 
-Our own method was trained on the 100,000 reads from chromosomes 1--8, with the
-highest number of duplicate reads.
+Our own method was trained on the 100,000 randomly selected reads from
+chromosomes 1--8, considering the initial read position and 20 nucleotides
+upstream and downstream.
 
-All data sets were mapped using Bowtie \citep{Langmead2009} against,
-respectively, the hg19, mm9, rheMac2, and panTro2 genome assemblies obtained
-from the UCSC Genome Browser \citep{Karolchik2008}.
+All data sets were mapped using Bowtie \citep{Langmead2009} using default
+parameters against, respectively, the hg19, mm9, rheMac2, and panTro2 genome
+assemblies obtained from the UCSC Genome Browser \citep{Karolchik2008}.
 
@@
 frequency, the KL divergence is computed as
-$$D_k( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
+$$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
 where the sum is over all $k$-mers.
@@
 The analysis provided demonstrates that our method shows significant improvement
-in two aspects: uniformity of read coverage, and consistency of nucleotide
-frequencies. These results strongly suggest an increase in accuracy, and we show
-in Supplementary Section 4, that this conclusion is further justified by
-comparison to qRT-PCR.
+in three aspects: uniformity of read coverage, and consistency of nucleotide
+frequencies, and agreement with qRT-PCR.
 
@@
 effective as the other models, even when data generated using random hexamer
-priming was used. Given the large number of parameters needed to estimate
-heptamer frequencies, a likely explanation is that the model is overfit to the
-training set.  We trained the 7mer model on a minimum of 600,000 reads (with the
-Bullard data set), and up to 23 million reads (with the Wetterbom data set), yet
-it seems more reads, or a greater diversity of reads are needed to produce an
-accurate or useful estimate.
+priming was used. A possible explaination is, given the large number of
+parameters needed to estimate heptamer frequencies ($4^{7} = 16,383$), these
+parameters are estimated with less accuracy than in models requiring fewer
+parameters. Yet, we trained the 7mer model on a minimum of 600,000 reads (with
+the Bullard data set), and up to 23 million reads (with the Wetterbom data set),
+numbers which our theoretical results, detailed in Supplementary Section 9,
+suggest should lead to accurate estimates. A perhaps more significant factor is
+that this method does not capture bias outside of the initial heptamer, though
+many data sets clearly are affected by bias in other positions.  However, if we
+were to incease the size of the $k$-mer being considered, exponentially more
+reads would be required for an accurate estimate.
 
 Our method generalizes this approach, attempting to overcome this problem by
-using an estimation of sequence probability that is more robust to overfitting
+using an estimation of sequence probability that requires fewer parameters
 and can account for bias outside of the initial heptamer. In all our tests, this

### 2011-09-29 seqbias-supplement.tex
more revisions

@@
 \usepackage{fontspec}
+\usepackage{float}
 
@@
 Observing the nonuniform distribution of nucleotide frequencies surrounding the
-5' end of reads, a natural step to take would be to trim the 5' end before mapping,
-in the hope that simply removing the portion of the read in which the bias
-occurs will also remove the bias. The figure below demonstrates that this is not
-the case. Trimming the initial heptamer in the Mortazavi data set does nothing to
-reduce the bias, and simply shifts the plot by seven positions. This indicates
-that the issue is \emph{sampling bias}, rather than a bias in base calling.
+5' end of reads, a natural step to take would be to trim the 5' end before
+mapping, in the hope that simply removing the portion of the read in which the
+bias occurs will also remove the bias. Figure \ref{fig:trimmedfreqs}
+demonstrates that this is not the case. Trimming the initial heptamer in the
+Mortazavi data set does nothing to reduce the bias, and simply shifts the plot
+by seven positions. This indicates that the issue is \emph{sampling bias},
+rather than a bias in base calling.
 
+\begin{figure}[H]
 \begin{center}
@@
 \end{center}
+\caption{Nucleotide frequencies observed before (plotted in red) and after
+(plotted in black) trimming the initial heptamer.}
+\label{fig:trimmedfreqs}
+\end{figure}
 
@@
 
+\begin{figure}[H]
 \begin{center}
@@
 \end{center}
+\caption{Sensitivity of the parameter control he standard deviation at which
+background sequences are sampled.}
+\label{fig:poissampstd}
+\end{figure}
 
-The above plot show the standard deviation used to draw background samples
-varied from 1 to 10,000. Apparent from this plot is that, while the optimal
-choice lies somewhere between 10 and 100, the model performs competitively for
-any reasonable value.
+Figure \ref{fig:poissampstd} show the standard deviation used to draw background
+samples varied from 1 to 10,000. Apparent from this plot is that, while the
+optimal choice lies somewhere between 10 and 100, the model performs
+competitively for any reasonable value.
 
+\begin{figure}[H]
 \begin{center}
@@
 \end{center}
+\caption{Sensitivity over adjustments to the complexity penalty.}
+\label{fig:poiscomppen}
+\end{figure}
 
-Here we multiply the complexity penalty term of the BIC by constants varying
+Next we multiplied the complexity penalty term of the BIC by a constants varying
 from $10^{-6}$ to $10^{3}$. (Recall that the BIC is $2\ell - m \log n$ where
@@
 
-It is clear from the above figure, that a very severe complexity penalty will
-result in a poor model. Here, for constants larger than 100, an empty model is
-trained, resulting in the goodness of fit of 0. Conversely, if this constant is
-very small, a overly-dense model will be trained.  In this data, for constants
-less than 0.01, the maximally dense model is trained (given the restraints on
-in-degree and edge distance). This result is a sub-optimal but entirely adequate
-solution.
+It is clear from Figure \ref{fig:poiscomppen}, that a very severe complexity
+penalty will result in a poor model. Here, for constants larger than 100, an
+empty model is trained, resulting in a median $R^2$ of 0. Conversely, if
+this constant is very small, a overly-dense model will be trained.  In this
+data, for constants less than 0.01, the maximally dense model is trained (given
+the restraints on in-degree and edge distance). This result is a sub-optimal but
+entirely adequate solution.
 
@@
 
-Below we plot the adjusted and unadjusted KL divergence plots, supplementing
-those in the results section of the main paper.
+Figure \ref{fig:klall} plots the adjusted and unadjusted KL divergence plots,
+supplementing those in the results section of the main paper.
 
+\begin{figure}[H]
+\centerline{
 \includegraphics[width=\textwidth]{fig/kl-all.pdf}
+}
+\caption{}
+\label{fig:klall}
+\end{figure}
 
@@
 
+Figure \ref{fig:aufreqs} plots sequence bias and KL divergence in the data
+published by Au, et al. \cite{Au2010} and examined by us in Section 3.3.
+
+\begin{figure}[H]
 \centerline{\includegraphics[width=0.6\textwidth]{fig/au-freqs.pdf}}
+\caption{}
+\label{fig:aufreqs}
+\end{figure}
 
@@
 but a longer training time. Here we investigate that trade-off by training our
-model on progressively more reads from the Mortazavi data set. We use the
-median pseudo-coefficient of determination ($R^2$) over exons selected for our
-test set. Each point is additionally labeled with the time used to train the
-model on one core of a 3Ghz Intel Xeon processor.
+model on progressively more reads from the Mortazavi data set. For each subset
+we train our method, evaluate the median pseudo-coefficient of determination
+($R^2$) over exons selected for our test set, and record the training time
+required on one core of a 3Ghz Intel Xeon processor. These results are plotted
+in Figure \ref{fig:poisn}.
 
+\begin{figure}[H]
 \includegraphics[width=\textwidth]{fig/pois-n.eps}
+\caption{Median $R^2$ is plotted against training set size. Each point is
+additionally labeled with the run time of the training procedure.}
+\label{fig:poisn}
+\end{figure}
+
+%% TODO: coptions/labels below this point
 
@@
 \end{center}
+% TODO: caption
+\label{fig:trimmedfreqs}
+\end{figure}
 
@@
 
+\begin{figure}[H]
 \begin{center}
-\includegraphics[width=0.5\textwidth]{fig/replicates.pdf}
+\includegraphics[width=0.7\textwidth]{fig/replicates.pdf}
 \end{center}
+% TODO: caption
+\label{fig:trimmedfreqs}
+\end{figure}
 
@@
 the MLE inferred distribution $Q$ with respect to the true distribution $P$ is
-$$E[H(Q||P)] = \frac{r - 1}{2n}$$
+\begin{equation}
+\label{eq:entropy}
+E[H(Q||P)] = \frac{r - 1}{2n}
+\end{equation}
 
@@
 formula is a very good fit when $n > r$.
-%% TODO: see figure X
+
+\begin{figure}[H]
+\centerline{\includegraphics[width=0.8\textwidth]{fig/entropy.pdf}}
+\caption{For each value of $n$ plotted, the circles indicate the value of
+$H(Q||P)$, averaged over 100 samples of size $n$, where $P$ is uniform and $Q$
+is the MLE estimator for $P$ basen on $n$ random samples drawn from $r$ bins.
+The asterisks and the straight line interpolating them are the theoretical
+approximation from Equation \ref{eq:entropy}.
+}
+\label{fig:entropy}
+\end{figure}
 

### 2011-09-29 seqbias-paper.tex
notes on non-empty probability

@@
 between genes or isoforms, and any test of differential expression hangs on the
-assumption that these biases are identical between replicates.  Additionally, in
-many tests of differential expression---similar to the effects of transcript
-length---a higher read count will result in higher statistical confidence. It
-follows that the sensitivity of such a test will also be biased by sequence,
-affecting downstream analysis such as gene ontology enrichment tests.
+assumption that these biases are identical between replicates, an undesirable
+assumption given that the causes of the bias are not well understood.
+Additionally, in many tests of differential expression---similar to the effects
+of transcript length---a higher read count will result in higher statistical
+confidence. It follows that the sensitivity of such a test will also be biased
+by sequence, affecting downstream analysis such as gene ontology enrichment
+tests.
 
@@
 
-\citet{Li2010} propose two models. The first is a Poisson linear
-model, in which read counts across a transcript follow an inhomogeneous Poisson
-process. The read count at position $i$ within the transcript is Poisson
-distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
+\citet{Li2010} propose two models. The first is a Poisson linear model, in which
+read counts across a transcript follow an inhomogeneous Poisson process. The
+read count at position $i$ within the transcript is Poisson distributed with
+parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of independent
 weights determined by the nucleotide at each position surrounding the read
@@
 (i.e. foreground and background distributions) requires \mbox{$2(4^7-1) =
-32,766$}
-parameters, so while no gene annotations are needed to train such a model, a
-significant number of reads are required.
+32,766$} parameters, so while no gene annotations are needed to train such a
+model, a significant number of reads are required, and a number that increases
+exponentially with $k$, if it were desirable to to model $k$-mers for $k > 7$.
 
@@
 this method applied to the $3'$ UTR of Apoa2, using data from
-\citet{Mortazavi2008}.}
+\citet{Mortazavi2008}. In bias coefficients predicted across 10 million positions
+of chromosome 1, the log10 bias of 95\% of the positions were between -1.14 and
+0.63, suggesting that most adjustments are not large.
+}
 \label{fig:overview}
@@
 specify each model is listed in parenthesis below. Applied to data with less
-bias, a sparser model is trained, as evinced by the Wetterbom data set.
+bias, a sparser model is trained, as evinced by the Wetterbom data set. Note
+that dependencies (i.e., arrows) tend to span a short distances, and nodes tend
+to have a small in-degree (i.e., have few inward arrows). In practice, we save
+time in training by prohibiting very distant dependencies (greater than 10, by
+default) or very high in-degrees (greater than 4, by default).
 }
@@
 \end{enumerate}
+
+Evident from Figure \ref{fig:overview}, the assumption of uniform read coverage
+often does not hold in typical RNA-Seq data sets. Although the bias corrected
+read counts across the exon pictured in this example are visibly more uniform,
+we sought a simple, objective tests that could be applied genome-wide. To this
+end, we used cross-validation tests (i.e., methods were trained and tested on
+disjoint subsets of the same RNA-Seq data sets) of a quantitative measure of the
+increase in uniformity of nucleotide frequencies (Kullback-Leibler divergence in
+Section 3.1) and increase in uniformity of read coverage (Poisson regression in
+Section 3.2). Additionally, we compare RNA-seq based estimate of gene expression
+to qRT-PCR based estimated for the same genes, showing increased correlation
+between the two methods after bias correction (Section 3.3).
 
@@
 $$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
-where the sum is over all $k$-mers.
+where the sum is over all $k$-mers. This can be thought of a measure
+dissimilarity between two probability distributions. If $f_x$ and $f'_x$ for a
+$k$-mer $x$ are approximately equal, their log-ratio will also be approximately
+zero, leading to a small KL divergence (zero, when the distributions are equal).
+Conversely, very different $k$-mer frequencies will result in a larger KL
+divergence.
 
@@
 This measure can be interpreted as the improvement in fit over the null model,
-with $R^2 = 1$ indicating a prefect fit, and smaller numbers, an increasingly
-worse fit. A more dubious, yet intuitive interpretation of $R^2$ is the
-proportion of variability explained by the model.
+with $R^2 = 1$ indicating a prefect fit, occurring model being evaluated
+achieves a likelihood of 1. Smaller number indicate an increasingly worse fit,
+with $R^2 = 0$ representing no improvement over the null model, and $R^2 = 0.5$,
+for example, indicating the model has a log-likelihood equal to half that of the
+null model (a large improvement, as log-likelihoods negative or zero). This
+measure can also take on values less than 0, indicating that the model has worse
+fit than the null model (i.e., when adjusting read counts according the bias
+coefficients leads to less uniform read coverage).
 

### 2011-09-29 seqbias-supplement.tex
notes on non-empty probability

@@
 
+\section{TODO: title}
+
+In the case that an experiment is unbiased, we would like an empty model to be
+trained, so that using the method to correct for bias will have no effect. Any
+non-empty model would otherwise decrease the accuracy of the quantification. To
+address this concern, we derive here an upper bound on the probability of a
+non-empty model being trained, when the data is unbiased.
+
+The training procedure begins by considering the the set of models formed by
+including each single nucleotide position from within the training sequences.
+For each single nucleotide model, the Bayesian infromation criterion is evaluated. If in any of these
+model the BIC score increases over the empty model, a non-empty model will be
+trained, otherwise the training procedure will halt with an empty model.
+
+We begin by considering one single nucleotide model. Suppose the background and
+foreground nucleotide distributions for this single position are equal (i.e.,
+the experiment is unbiased in this position) specified by the parameters $(p_{a}, p_{c},
+p_{g}, p_{t})$.
+
+The model is trained on a set of $n$ foreground and $n$ background sequences.
+Let $X = (x_{a}, x_{c}, x_{g}, x_{t})$ give the number of times each nucleotide
+was observed in the foreground sample, and $Y = (y_{a}, y_{c}, y_{g}, y_{t})$
+in the background sample.
+
+First, suppose the nucleotide distribution estimated from $X$ and from $Y$ both
+have KL divergence bounded by some number $\epsilon$:
+$$ D_{X} = \sum_{k \in \{a, c, g, t\}} (x_k / n) \log \left( \frac{x_k / n}{p_k}
+\right) < \epsilon $$
+and
+$$ D_{Y} = \sum_{k \in \{a, c, g, t\}} (y_k / n) \log \left( \frac{y_k / n}{p_k}
+\right) < \epsilon $$
+
+Now, objective function evaluated is the Bayesian information criterion (BIC)
+applied to the conditional log-likelihood (CLL).
+$$ \text{BIC} = 2 \cdot \text{CLL} - m \log n $$
+where CLL in this single nucleotide case can be written as,
+$$ \text{CLL} =
+\sum_{k \in \{a, c, g, t\}}
+\left(
+x_k \log \left( \frac{x_k/n}{(x_k + y_k)/n} \right) +
+y_k \log \left( \frac{y_k/n}{(x_k + y_k)/n} \right) \right)$$
+
+The KL divergence obtained by combining the two samples $X$ and $Y$ is,
+$$D_{XY} = \sum_{k \in \{a, c, g, t\}}
+\frac{(x_k + y_k)}{2n}
+\log \left( \frac{(x_k + y_k)/2n}{p_k} \right)$$
+
+The KL divergence is non-negative, and so adding $2n D_{XY}$ to the CLL, we get
+\begin{align*}
+CLL &\le CLL + 2n D_{XY} \\
+&=
+\textstyle{
+\sum_{k \in \{a, c, g, t\}} 
+x_k \log \left( \frac{x_k/n}{(x_k + y_k)/n} \right)
++ y_k \log \left( \frac{y_k/n}{(x_k + y_k)/n} \right)
++ (x_k + y_k) \log \left( \frac{(x_k + y_k)/2n}{p_i} \right)} \\
+&=
+\textstyle{
+\sum_{k \in \{a, c, g, t\}}
+\left(
+x_k \log \left( \frac{x_k/n}{2 p_i} \right) +
+y_k \log \left( \frac{y_k/n}{2 p_i} \right)
+\right)} \\
+&=
+\textstyle{
+n D_{X} + n D_{Y} + \sum_{k \ in \{a, c, g, t\}} (x_k + y_k) \log (1/2)
+} \\
+%\textstyle{
+%n \sum_{k \in \{a, c, g, t\}}
+%\left( (x_k / n) \log \left( \frac{x_k/n}{p_i} \right) \right) +
+%n \sum_{k \in \{a, c, g, t\}}
+%\left( (y_k / n) \log \left( \frac{y_k/n}{p_i} \right) \right) +
+%\sum_{k \in \{a, c, g, t\}} (x_k + y_k) \log (1/2)} \\
+&< 2n\epsilon + 2n \log (1/2)
+\end{align*}
+Thus, if the KL divergence of the two samples $X$ and $Y$ are bounded, the
+conditional log-likelihood is as well.
+
+The BIC score of an empty model is,
+$$\text{BIC}_0 = 2 \sum_{k \in \{a, c, g, t\}}
+\left(
+x_k \log(1/2) + y_k \log(1/2) \right) - m
+\log n = 4 n \log (1/2) $$
+where the number of parameters needed to specify the empty model is $m = 0$.
+
+The BIC score of the single nucleotide model, when the KL divergence of $X$ an
+$Y$ is bounded by $\epsilon$ is
+\begin{align*}
+\text{BIC} &= 2 \cdot CLL - m \log n \\
+&= 2 \cdot CLL - 6 \log n \\
+&< 4n \epsilon + 4 n \log (1/2) - 6 \log n
+\end{align*}
+
+It follows that if
+\begin{align*}
+4n \epsilon + 4 n \log 1 / 2 - 6 \log n  &\le   4 n \log 1/2 \\
+\epsilon &\le \frac{3}{2n} \log n
+\end{align*}
+then
+$$ \text{BIC} \le \text{BIC}_0 $$
+
+In summary, if $D_X, D_Y < \frac{3}{2n} \log n$, then $\text{BIC} \le
+\text{BIC}_0$ and the empty model is retained. That is, the probability of an
+empty model being trained is bounded below by the probability that the KL
+divergence of both sampled is bounded above $\frac{3}{2n} \log n$:
+$$ \Pr(D_{X} < \frac{3}{2n} \log n) \cdot 
+\Pr(D_{Y} < \frac{3}{2n} \log n) $$
+
+Since $X$ and $Y$ are drawn from the same distribution, $D_{X}$ and $D_{Y}$ are
+identically distributed as well, and so we can simplify this formula as,
+$$ (\Pr(D < \frac{3}{2n} \log n))^2 $$
+where $D$ is the divergence of any sample of size $n$. The probability of a
+non-empty model being trained is then less than
+$$ 1 - (\Pr(D < \frac{3}{2n} \log n))^2 $$
+
+In training we consider mulitple nucleotides. If $h$ nucleotides are considered,
+the probability of a non-empty model being trained is less than
+$$ 1 - (\Pr(D < \frac{3}{2n} \log n))^{2h} $$
+
+As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $chi^2$
+distribution with 3 degrees of freedom for large $n$, and so we can estimate
+this probability for various values of $n$. This is shown in Figure
+\ref{fig:nepr}. From this figure we can see, for example that training with more
+than 10,000 reads results in an incorrect non-empty model with probability less
+than 0.0004.
+
+In the low end, the upper bound on non-empty model probability is much higher.
+In our tests, this upper bound is not tight for a small number of reads. We
+trained the model on 100 sets of 100 simulated reads, drawn randomly at uniform
+from exonic sequences and in only 3 trials was a non-empty model trained, far
+less than the upper bound prediction of 23 out of 100. This test was repeated
+using sets of 50,000 reads, and in no trial was a non-empty model trained.
+
+From these results, and the results in Supplementary Section 5, we recommend
+training our method with at least 10,000 reads.
+
+\begin{figure}[H]
+\centerline{
+\includegraphics[width=0.8\textwidth]{fig/nepr.pdf}}
+\caption{Plotted is the upper bound on the probability of training a non-empty
+model when there is no real bias in the experiment when the model is trained
+with various number of reads $n$. We fixed the number of nucleotides being
+considered as $h = 41$, as this what was used for the evaluation in Section 3.}
+\ref{fig:nepr}
+\end{figure}
+
 \bibliographystyle{plain}

### 2011-09-29 response/response.tex
(almost) final draft

@@
 We appreciate the time and effort extended by the anonymous reviewers and the
-editorial staff to evaluate the original version of this manuscript, which was
-submitted on July 8. Upon review, the second of the two reviewers raised some
-concerns leading to its rejection. However, we believe that the bulk of these
-concerns are either incongruous with the content of the manuscript or
-addressable with very little effort, and therefore do not warrant its rejection.
-Here we have provided a detailed response to these comments, and have included
-additional material in an effort to clarify these points.
+editorial staff to evaluate the original version of this manuscript (titled
+``Correcting for bias in high-throughput sequencing data''), which was submitted
+on July 8. Upon review, the second of the two reviewers raised some concerns
+leading to its rejection. However, we believe that the bulk of these concerns
+are either incongruous with the content of the manuscript or addressable with
+very little effort, and therefore do not warrant its rejection.  Here we have
+provided a detailed response to these comments, and have included additional
+material in an effort to clarify these points.
 
@@
 
-To further explore the point raised, we repeatedly trained our model on sets of
-10,000 random simulated reads, drawn uniformly from within exons. We repeated
-this 100 times, and in each trial an empty model was trained.  Reducing the
-number of reads increases the risk of a spuriously non-empty model, but
-repeating this evaluating with only 100 reads led to only 3 out of 100 models
-being non-empty. This is indicative that the default BIC scoring, as predicted
-by statistical theory, is adequately protecting against false discovery of
-``bias'' when there is none.
+We have included additional material in Supplementary Section 10 exploring this
+point in detail. We first prove an upper bound on the probability of the model
+predicting the presence of bias when there is none, demonstrating that is less
+than 0.0004 when more than 10,000 reads are used. We then investigate this
+empirically by simulating random reads, drawn uniformly from exonic sequences
+within the human genome.  We trained our model 100 times with 10,000 simulated
+reads, and in no instance was bias predicted (i.e., a non-empty model trained).
+We repeated this with using only 100 reads, and in only 3 out of 100 trials was
+bias incorrectly predicted.
+
+This is indicative that the default BIC scoring, as predicted by statistical
+theory, is adequately protecting against false discovery of bias when there is
+none, and there is very little risk in applying the method to an unbiased
+experiment.
 
@@
 by virtue of having only a few hundred parameters learned from counts
-(Figure 3), as opposed to tens of thousands.  [KL divergence between
-two distributions is inversely proportional to the number of samples
-one needs to confidently test the hypothesis that the samples arise
-from one distribution versus the other; see Supplementary Section XX.]
+(Figure 3), as opposed to tens of thousands.
+
+%[KL divergence between
+%two distributions is inversely proportional to the number of samples
+%one needs to confidently test the hypothesis that the samples arise
+%from one distribution versus the other; see Supplementary Section XX.]
 %% TODO: why the brackets here?
@@
 
-%% TODO
-
-%The revised Figure 2 also graphically illustrates the range of
-%adjustment factors, which genome-wide are very cool and we hope we can say something soothing about.......
+In Figure 2, we now indicate that the adjustments made by the method to read
+counts tend to be small. In our test, 95\% of a sample of 10 million predicted
+bias coefficients were between 0.07 and 4.27. The lack of large adjustments is
+also reflected indirectly in the results in Section 3.  This is reflected
+indirectly in the results of Section 3: extremely large adjustments would result
+in large variance of the predictions, leading to poor performance in our
+evaluations, which we do not see.
 
@@
 
-%\subsection*{Other Changes}
+\subsection*{Other Changes}
 
-%We have modified the title to more clearly acknowledge the existence of
-%the prior work upon which we build.
+We have modified the title to more clearly acknowledge the existence of
+the prior work upon which we build.
 

### 2011-09-29 seqbias-supplement.tex
(almost) final draft

@@
 
-%% TODO: coptions/labels below this point
-
 \section{ChIP-Seq}
@@
 than that observed the RNA-Seq data sets we tested, but the data was by no means
-unbiased.
+unbiased, as shown in Figure \ref{fig:freqscao}.
 
@@
 \end{center}
-% TODO: caption
-\label{fig:trimmedfreqs}
+\caption{Nucleotide frequencies and KL divergence for the Cao data set.}
+\label{fig:freqscao}
 \end{figure}
@@
 
-Below, we plot directly the positional KL divergence, computed in tho same
-manner described in the Results section of the paper.
+Figure \ref{fig:caokl}, we plot directly the positional KL divergence, computed
+in tho same manner described in the Results section of the paper.  We see that
+our method is effective at reducing the bias in this ChIP-Seq data.
 
@@
 \end{center}
-% TODO: caption
-\label{fig:trimmedfreqs}
+\caption{KL divergence for the Cao data set, after adjusting read counts for
+bias.}
+\label{fig:caokl}
 \end{figure}
-
-We see that our method is effective at reducing the bias in this ChIP-Seq data.
 
@@
 \cite{Langmead2009}. The nucleotide frequencies and divergence are plotted
-below.
+in Figure \ref{fig:frtall}.
 
@@
 \end{center}
-% TODO: caption
-\label{fig:trimmedfreqs}
+\caption{Nucleotide frequencies and KL divergence of the Mamanova data set.}
+\label{fig:frtall}
 \end{figure}
@@
 only a small number of positions surrounding the read start. When our method is
-trained on this data we obtain the suitably sparse model pictured below.
+trained on this data we obtain the suitably sparse model pictured in Figure
+\ref{fig:frtgraph}.
 
@@
 \end{center}
-% TODO: caption
-\label{fig:trimmedfreqs}
+\caption{The graphical model learned by training our method on the Mamanova data
+set.}
+\label{fig:frtgraph}
 \end{figure}
@@
 question the accuracy of such tests. In our experiments, we have observed the
-bias to be mostly, but not entirely consistent between replicates. Below, we
+bias to be mostly, but not entirely consistent between replicates. In Figure
+\ref{fig:replicates} we
 plot the frequencies from the four runs in the Trapnell data set.
@@
 \end{center}
-% TODO: caption
-\label{fig:trimmedfreqs}
+\caption{Nucleotide frequencies of replicates in the Trapnell data set.}
+\label{fig:replicates}
 \end{figure}
@@
 
-\subsection{TODO: Title}
+\subsection{Overview}
 
@@
 
-\section{TODO: title}
+\section{False Discovery of Bias in Unbiased Experiments}
 
@@
 including each single nucleotide position from within the training sequences.
-For each single nucleotide model, the Bayesian infromation criterion is evaluated. If in any of these
-model the BIC score increases over the empty model, a non-empty model will be
-trained, otherwise the training procedure will halt with an empty model.
+For each single nucleotide model, the Bayesian infromation criterion is
+evaluated. If in any of these model the BIC score increases over the empty
+model, a non-empty model will be trained, otherwise the training procedure will
+halt with an empty model.
 
@@
 
-As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $chi^2$
+As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $\chi^2$
 distribution with 3 degrees of freedom for large $n$, and so we can estimate
@@
 less than the upper bound prediction of 23 out of 100. This test was repeated
-using sets of 50,000 reads, and in no trial was a non-empty model trained.
+using sets of 10,000 reads, and in no trial was a non-empty model trained.
 
@@
 considered as $h = 41$, as this what was used for the evaluation in Section 3.}
-\ref{fig:nepr}
+\label{fig:nepr}
 \end{figure}

### 2011-09-30 response/response.tex
final draft

@@
 method developed by Hansen, et al., performed poorly regardless of the number of
-reads used in training, which varied between hundreds of thousands to tens of
+reads used in training, which varied from hundreds of thousands to tens of
 millions.  The methods presented by Li, et al., are more effective, yet involve
@@
 
-Under these circumstances we respectively request that the manuscript, in this
-revised form, be reconsidered for publication.  Detailed responses to
-the reviewers concerns are presented below.
+Under these circumstances we respectfully request that the manuscript, in this
+revised form, be reconsidered for publication. Detailed responses to the
+reviewers concerns are presented below.
 
@@
 
-First, we emphasize that PCR bias is only one of a number of technical steps
+First, we emphasize that PCR amplification is only one of a number of technical steps
 involved in sequencing that potentially introduce bias.  Our method is agnostic
@@
 reads, and in no instance was bias predicted (i.e., a non-empty model trained).
-We repeated this with using only 100 reads, and in only 3 out of 100 trials was
+We repeated this using only 100 reads, and in only 3 out of 100 trials was
 bias incorrectly predicted.
@@
 
-We regret that we were not more explicit on this point. At least
-600,000 reads on each data set, but as many as 23 million were used:
-see the revised table X. (As noted in the paper, we did reimplement
-the Genominator method in order to do this.)
+We regret that we were not more accurate or explicit on this point. The 600,000
+figure somewhat underestimated the actual number used. Table 2 now given the
+correct numbers of reads used to train each method. The minimum number of reads
+used to train the Hansen method was in fact 1.9 million, and we used as many as
+23 million.  (As noted in the paper, we did reimplement the Genominator method
+in order to do this.)
 
@@
 
-Understanding this failure of a simple, appealing method is important
-for making progress, and we feel that our paper provides a theoretical
-basis for understanding these differences in efficacy, and we have
-amplified these points in the revision.  Intuitively, one expects
-model accuracy to increase as the number $n$ of training reads
-increases, but to decrease as the number $r$ of model parameters
-grows, since the training data is ``spread more thinly.''  For the
-particular case of estimating a multinomial distribution (as in
-Hansen), Supplementary Section YY quantifies this intuition very
-precisely, showing that the expected KL divergence of the learned
-model with respect to the true distribution converges to $(r-1)/(2n)$,
-where $r=4^7=16384$ is the number of model parameters learned.  The
-same analysis does not directly apply to the MART, GLM or BN models,
-but we expect their behavior to be qualitatively similar.  Thus, to
-the extent that a more parsimonious model such as ours adequately
-captures the structure of the data, it is potentially orders of
-magnitude more accurate for a given quantity of training data, simply
-by virtue of having only a few hundred parameters learned from counts
-(Figure 3), as opposed to tens of thousands.
+Understanding this failure of a simple, appealing method is important for making
+progress, and we feel that our paper provides a theoretical basis for
+understanding these differences in efficacy, and we have amplified these points
+in the revision.  Intuitively, one expects model accuracy to increase as the
+number $n$ of training reads increases, but to decrease as the number $r$ of
+model parameters grows, since the training data is ``spread more thinly.''  For
+the particular case of estimating a multinomial distribution (as in Hansen),
+Supplementary Section 9 quantifies this intuition very precisely, showing that
+the expected KL divergence of the learned model with respect to the true
+distribution very closely approximates $(r-1)/(2n)$, where $r=4^7=16384$ is the
+number of model parameters learned.  The same analysis does not directly apply
+to the MART, GLM or BN models, but we expect their behavior to be qualitatively
+similar.  Thus, to the extent that a more parsimonious model such as ours
+adequately captures the structure of the data, it is potentially orders of
+magnitude more accurate for a given quantity of training data, simply by virtue
+of having only a few hundred parameters learned from counts (Figure 3), as
+opposed to tens of thousands.
 
-%[KL divergence between
-%two distributions is inversely proportional to the number of samples
-%one needs to confidently test the hypothesis that the samples arise
-%from one distribution versus the other; see Supplementary Section XX.]
-%% TODO: why the brackets here?
-
-Of course, training data is plentiful in NGS data sets, so,
-pragmatically, the above analysis probably identifies a contributor
-but not the root cause of Hansen et al.'s relatively poor performance
-in our experiments.  Rather, we believe the problem is that bias is
-not confined to the initial heptamer.  The model structures
-illustrated in Figure 3 show that nucleotides well outside that region
-are often biased.  (The randomization study sketched above shows that
-our model building algorithm is not prone to spurious discovery of
-overly-complex models.)  Furthermore, these examples show that the
-read sampling is often biased by positions \emph{in front of the
-  read}, which are not currently considered by Hansen.  Their method
-could be modified easily to consider $k$-mers spanning the read start,
-but the entropy analysis shows that extending it to $k \gg 7$ is
-infeasible, since the exponential growth in $r=4^k$ will destroy the
-model's accuracy even in large data sets.  E.g., $k \ge 20$ would be
-necessary to cover the biased regions in all data sets shown in Figure
-3, necessitating trillions of training reads to build an adequate
-model by Hansen's method.
+Of course, training data is plentiful in NGS data sets, so, pragmatically, the
+above analysis probably identifies a contributor but not the root cause of
+Hansen et al.'s relatively poor performance in our experiments. Rather, we
+believe the problem is that \emph{bias is not confined to the initial
+heptamer}.The model structures illustrated in Figure 3 show that nucleotides
+well outside that region are often biased.  (The the theoretical results and
+randomization studies sketched above, and presented in Supplementary Section 10
+show that our model building algorithm is not prone to spurious discovery of
+overly-complex models.)  Furthermore, these examples show that the read sampling
+is often biased by positions \emph{in front of the read}, which are not
+currently considered by Hansen.  Their method could be modified easily to
+consider $k$-mers spanning the read start, but the entropy analysis in
+Supplementary Section 9 shows that extending it to $k \gg 7$ is infeasible,
+since the exponential growth in $r=4^k$ will destroy the model's accuracy even
+in large data sets. E.g., $k \ge 20$ would be necessary to cover the biased
+regions in all data sets shown in Figure 3, necessitating trillions of training
+reads to build an adequate model by Hansen's method.
 
@@
 
-Instead, we presented data (in Section 3.2) that answers the
-reviewer's question (i.e., ``does this unevenness go away with the
-proposed adjustment?'') in rigorous and comprehensive manner. We
-implemented a Poisson regression metric, which quantitatively measures
-``uniformity'' by calculating the likelihood that the reads sampled
-across a given exon were sampled from a uniform distribution.  Using
-this, we performed a large cross-validation study proving the
-effectiveness of our method at increasing uniformity across thousands
-of exons drawn from a diverse group of data sets, rather than
-qualitatively across one exon.  In broad terms, our method increases
-likelihood of the uniform model by XXX-fold across thousands of exons
-(Figure 6).
+Instead, we presented data (in Section 3.2) that answers the reviewer's question
+(i.e., ``does this unevenness go away with the proposed adjustment?'') in
+rigorous and comprehensive manner. We implemented a Poisson regression metric,
+which quantitatively measures ``uniformity'' by calculating the likelihood that
+the reads sampled across a given exon were sampled from a uniform distribution.
+Using this, we performed a large cross-validation study proving the
+effectiveness of our method at increasing uniformity across thousands of exons
+drawn from a diverse group of data sets (Figure 5), rather than qualitatively
+across one exon.
 
@@
 The results in sections 3.1 and especially in 3.2 are far more
-comprehensive.
+quantitative and comprehensive.
 
@@
 counts tend to be small. In our test, 95\% of a sample of 10 million predicted
-bias coefficients were between 0.07 and 4.27. The lack of large adjustments is
-also reflected indirectly in the results in Section 3.  This is reflected
-indirectly in the results of Section 3: extremely large adjustments would result
-in large variance of the predictions, leading to poor performance in our
-evaluations, which we do not see.
+bias coefficients were between 0.07 and 4.27. This is also reflected indirectly
+in the results of Section 3: extremely large adjustments would result in large
+variance of the predictions, leading to poor performance in our evaluations,
+which we do not see.
 
@@
 
-In this revision, we have included more data, showing that in addition to the
-improvements in uniformity and nucleotide frequencies, increased correlation
-between RNA-Seq and qRT-PCR, after adjusting for bias in RNA-Seq. These results
-are consistent with, and reinforce our previous results.
+In this revision, we have included more data (Section 3.3), showing, in addition
+to the improvements in uniformity and nucleotide frequencies, significantly
+increased correlation between RNA-Seq and qRT-PCR, after adjusting for bias in
+RNA-Seq.  These results are consistent with, and reinforce our previous results.
 
@@
 
-Our motivation for adopting a Bayesian network model is the very
-natural probabilistic interpretation that we have described in Section
-2.1. We disagree with the reviewer's opinion that Bayesian networks
-are inherently more complicated than Lasso regression or support
-vector machines.  In this application, the latter two methods would
-generate output that is arguably less transparent and interpretable,
-and would neither be significantly simpler to implement, nor involve
-fewer parameters.  Our hypotheses (above) concerning the reasons for
-the weakness of Hansen's method illustrate the benefit of this
-approach; we do not believe svm regression, say, would have yielded
-similar insights.
+We agree completely that a spectrum that a spectrum of methodhs are available.
+Our motivation for adopting a Bayesian network model is the very natural
+probabilistic interpretation that we have described in Section 2.1. We disagree
+however with the reviewer's opinion that Bayesian networks are inherently more
+complicated than Lasso regression or support vector machines.  In this
+application, the latter two methods would generate output that is arguably less
+transparent and interpretable, and would neither be significantly simpler to
+implement, nor involve fewer parameters.  Our hypotheses (above) concerning the
+reasons for the weakness of Hansen's method illustrate the benefit of this
+approach; we do not believe svm regression, say, would have yielded similar
+insights.
 

### 2011-09-30 seqbias-paper.tex
(almost) final draft
final draft

@@
 
-\title{Correcting for bias in high-throughput sequencing data}
+%\title{Correcting for bias in high-throughput sequencing data}
+\title{A new approach to bias correction in high-throughput sequencing data}
 \author[Jones \textit{et~al}]
 {Daniel C. Jones\,$^{1,}$
-\footnote{to whom correspondence should be addressed}\hspace{0.5em}
-Walter L. Ruzzo\,$^{1,2,3}$
-Xinxia Peng\,$^{4}$
+\hspace{-0.5em}\footnote{to whom correspondence should be addressed}\hspace{0.5em},
+Walter L. Ruzzo\,$^{1,2,3}$,
+Xinxia Peng$^{4}$
+and
 Michael G. Katze$^{4}$
@@
 Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
-often conflated by protocol-specific sequence bias. The sources of the bias are
-unknown, but may influenced by PCR amplification, or differing primer affinities
-and mixtures, for example.  The result is decreased accuracy in many
-applications, such as de novo gene annotation and isoform quantification in
+often conflated by protocol-specific sequence bias. The exact sources of the
+bias are unknown, but may be influenced by PCR amplification, or differing
+primer affinities and mixtures, for example.  The result is decreased accuracy
+in many applications, such as de novo gene annotation, isoform quantification in
 RNA-Seq, and peak calling in ChIP-Seq.
@@
 \section{Results:}
-We present a method to measure and correct for these influences using a simple
-graphical model. Our model does not rely on existing gene annotations, making it
-applicable to any high throughput sequencing data. We evaluate our method on
-several data sets, and by multiple criteria, demonstrating that it effectively
-decreases bias and increases uniformity.
+We present a new method to measure and correct for these influences using a
+simple graphical model. Our model does not rely on existing gene annotations,
+making it applicable to any quantitative high throughput sequencing data. We
+evaluate our method on several data sets, and by multiple criteria,
+demonstrating that it effectively decreases bias and increases uniformity.
 
@@
 normalization to accounting for cross-hybridization, RNA-Seq presents a new set
-of problems.
-
-A significant example is the effect of transcript length on quantification.  As
-noted by \citet{Oshlack2009} transcript length biases discovery of
-differential expression, since more reads are aligned to longer transcripts,
-granting higher statistical significance in tests of differential expression.
-Recently, some methods have been developed attempting to account for this effect
-in gene set enrichment tests \citep{Young2010,Gao2011}, leading to more accurate
-downstream analysis.
+of challenges. A significant example is the effect that transcript length has in
+biasing quantification, as first explored by \citet{Oshlack2009}, and
+subsequently addressed by \cite{Young2010, Gao2011}.
 
@@
 illustration, Figure \ref{fig:freqs} plots this nonuniformity in nucleotide
-frequencies on five data sets (see, Table \ref{tab:datasets}), each using a
+frequencies on five data sets (see Table \ref{tab:datasets}), each using a
 different protocol.
@@
 background heptamer probabilities. Specifying two distributions over heptamers
-(i.e. foreground and background distributions) requires \mbox{$2(4^7-1) =
+(i.e., foreground and background distributions) requires \mbox{$2(4^7-1) =
 32,766$} parameters, so while no gene annotations are needed to train such a
 model, a significant number of reads are required, and a number that increases
-exponentially with $k$, if it were desirable to to model $k$-mers for $k > 7$.
+exponentially with $k$, if it were desirable to model $k$-mers for $k > 7$.
 
@@
 That is, if $N$ reads are generated, and $m_i$ is the event that a generated
-read maps to position $i$.
+read maps to position $i$, then
 $$ E[x_i] = N \Pr[m_i] $$
@@
 of chromosome 1, the log10 bias of 95\% of the positions were between -1.14 and
-0.63, suggesting that most adjustments are not large.
+0.63, suggesting that most adjustments are not large. The $R^2$ measure,
+detailed in Section 3.2, gives the relative increase in log-likeliood under a
+uniform sampling model, after correcting for bias, with 1.0 indicating a perfect
+fit, and the score of 0.38 here indicating a significant increase.
 }
@@
 parameters needed to specify the model, we maximize, 
-$$ \ell' = 2 \ell - m \log n $$
+$ \ell' = 2 \ell - m \log n $.
 
@@
 \begin{enumerate}
-\item The in-degree (i.e. number of parents) of any node must be less than some
+\item The in-degree (i.e., number of parents) of any node must be less than some
 number $p_{\text{max}}$.
@@
 the expense of increased training time.
-
-The quality of the solution depends also on two other parameters: the standard
-deviation at which background sequences are sampled, and the weight applied to
-the penalty term of the BIC, yet it is not particularly sensitive to their
-values. (The median $R^2$ goodness-of-fit statistic used in section 3.2 varied
-by less that 25\% as these parameters were varied over a range of $10^4$. See
-Supplementary Section 2.) In all our evaluation, no special tuning of the
-parameters was performed, suggesting it can be used effectively across data sets
-without any intervention.
 
@@
 process.
-\item Adjusting for bias should increase the agreement between RNA-Seq and
-qRT-PCR.
+\item Adjusting for bias in RNA-Seq should increase the agreement between RNA-Seq and
+an other methods of quantification.
 \end{enumerate}
@@
 Section 3.1) and increase in uniformity of read coverage (Poisson regression in
-Section 3.2). Additionally, we compare RNA-seq based estimate of gene expression
-to qRT-PCR based estimated for the same genes, showing increased correlation
+Section 3.2). Additionally, we compare RNA-Seq based estimate of gene expression
+to qRT-PCR based estimates for the same genes, showing increased correlation
 between the two methods after bias correction (Section 3.3).
@@
 \begin{tabular}{lllcc}\toprule
- & & & Read & Num. of \\
-Experiment & Species & Platform & length & reads \\\midrule
-\citet{Wetterbom2010} & Chimpanzee & ABI & 33 & $3.9 \times 10^7$ \\
-Katze (unpublished) & Macaque & ABI & 50  & $4.8 \times 10^7$ \\
-\citet{Bullard2010} & Human & Illumina & 35 & $1.2 \times 10^7$ \\
-\citet{Mortazavi2008} & Mouse & Illumina & 33 & $3.9 \times 10^6$ \\
-\citet{Trapnell2010} & Mouse & Illumina & 75 & $6.6 \times 10^7$ \\\botrule
+ & & & Read  \\
+Experiment & Species & Platform & length \\\midrule
+\citet{Wetterbom2010} & Chimpanzee & ABI & 33  \\
+Katze (unpublished) & Macaque & ABI & 50  & \\
+\citet{Bullard2010} & Human & Illumina & 35 \\
+\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
+\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
 \end{tabular}
 }{}
+\end{table}
+
+\begin{table}
+\processtable{Numbers of reads used to train each method.\label{tab:counts}}
+{
+\begin{tabular}{lcccc}\toprule
+Experiment & Total & 7mer & GLM/MART & BN \\\midrule
+Wetterbom & $3.9 \times 10^7$ & $2.3 \times 10^7$ & $5.9 \times
+10^6$ & $1 \times 10^5$ \\
+Katze & $4.8 \times 10^7$ & $2.0 \times 10^7$ & $3.1 \times 10^6$ & $1 \times
+10^5$ \\
+Bullard & $1.2 \times 10^7$ & $3.1 \times 10^6$ & $1.8 \times 10^5$ & $1 \times
+10^5$\\
+Mortazavi & $3.9 \times 10^6$ & $1.9 \times 10^6$ & $3.8 \times 10^5$ & $1
+\times 10^5$ \\
+Trapnell & $3.4 \times 10^7$ & $1.6 \times 10^7$ & $5.9 \times 10^6$ & $1 \times
+10^5$
+\end{tabular}
+}{Note that in the Trapnell data set (the only paired-end set considered) we
+count only the first mate of each read.}
 \end{table}
@@
 The differences in the methods being tested necessitated training procedures
-unique to each.
+unique to each. The total number of reads used to train each method is listed in
+Table \ref{tab:counts}, and below we describe the procedure used for
+each.
 
@@
 $$D_k( f, f' ) = \sum_{x} \left( f_x \log_{2}( f_x / f'_x ) + f'_x \log_{2}( f'_x / f_x) \right)$$
-where the sum is over all $k$-mers. This can be thought of a measure
+where the sum is over all $k$-mers. This can be thought of as a measure
 dissimilarity between two probability distributions. If $f_x$ and $f'_x$ for a
-$k$-mer $x$ are approximately equal, their log-ratio will also be approximately
-zero, leading to a small KL divergence (zero, when the distributions are equal).
+$k$-mer $x$ are approximately equal, their log-ratio will be approximately
+zero, leading to a small KL divergence (exactly zero, when the distributions are equal).
 Conversely, very different $k$-mer frequencies will result in a larger KL
@@
 
-In this comparison, we measure how well the counts conform to a Poisson process.
-The assumption of positional read counts following a Poisson distribution is
-known to be a poor fit \citep{Srivastava2010}, but measuring the improvement in
-the fit derived from correcting for bias remains a principled and easily
-interpreted criterion. This increase in uniformity is illustrated in Figure
-\ref{fig:overview}.
+In this comparison, we measure the uniformity of the data, or more precisely
+how well the counts conform to a Poisson process.  The assumption of positional
+read counts following a Poisson distribution is known to be a poor fit
+\citep{Srivastava2010}, but measuring the improvement in the fit derived from
+correcting for bias remains a principled and easily interpreted criterion. This
+increase in uniformity is illustrated in Figure \ref{fig:overview}.
 
@@
 This measure can be interpreted as the improvement in fit over the null model,
-with $R^2 = 1$ indicating a prefect fit, occurring model being evaluated
-achieves a likelihood of 1. Smaller number indicate an increasingly worse fit,
-with $R^2 = 0$ representing no improvement over the null model, and $R^2 = 0.5$,
-for example, indicating the model has a log-likelihood equal to half that of the
-null model (a large improvement, as log-likelihoods negative or zero). This
-measure can also take on values less than 0, indicating that the model has worse
-fit than the null model (i.e., when adjusting read counts according the bias
+with $R^2 = 1$ indicating a perfect fit, occurring when the model being
+evaluated achieves a likelihood of 1. Smaller number indicate an increasingly
+worse fit, with $R^2 = 0$ representing no improvement over the null model, and
+$R^2 = 0.5$, for example, indicating the model has a log-likelihood equal to
+half that of the null model (a large improvement, corresponding to, for example,
+the likelihood increasing over 100-fold if the initial log-likelihood was -9.6,
+which is the mean per-position log-likehood under the null model). This measure
+can also take on values less than 0, indicating that the model has worse fit
+than the null model (i.e., when adjusting read counts according the bias
 coefficients leads to less uniform read coverage).
@@
 Figure \ref{fig:pois}.  To summarize each model with a single number, we can
-examine the median $R^2$ value, as listed in Table \ref{tab:pois}.
+examine the median $R^2$ value, as listed in Table \ref{tab:pois}. Our method
+shows a highly statistically significant improvement in performance over other
+methods in all but one comparison, in which the MART method performs equally.
 
@@
 \centerline{\includegraphics[width=0.30\textwidth]{pois-boxplot.eps}}
-\caption{For each of the 1000 test exons, we compute a pseudo-coefficients of
+\caption{For each of the 1000 test exons, we compute McFadden's pseudo-coefficient of
 determination $R^2$, equivalent to the improvement in log-likelihood under the
@@
 }
-
     \label{fig:pois}
@@
     correcting for bias.  Here the median $R^2$ across the test exons is listed
-    for each method and sample. A higher $R^2$ indicates a better fit.
+    for each method and sample. A higher $R^2$ indicates a better fit. The
+    highest value in each row is highlighted in bold.
 }
@@
 RNA-Seq data shows a pattern of bias similar to that seen in the other samples
-sequenced on an Illumina platform (see, Supplementary Section 5).
+sequenced on an Illumina platform (see Supplementary Section 5). This evaluation
+does not rely on an assumption that qRT-PCR necissarily more accurate than
+RNA-Seq based quantification, only that qRT-PCR is not biased in the same way as
+the RNA-Seq data.
 
@@
 dividing by the length of these genes. We then removed any genes with a read
-count less than 10, or that did not correspond to
+count less than 10, or that did not correspond to a unique TaqMan probe.
 
@@
 log TaqMan expression values, which are averaged across three replicates. These
-correlations are listed in Table \ref{tab:pcr}.
+correlations are listed in Table \ref{tab:pcr}. Our method shows a statistically
+significant increase in correlation compared to the other methods.
 
@@
 using the adjusted count from each method. We repeated this procedure one
-million times, and counted the number of times each of the competing method achieves a
-higher correlation than the BN method.  Those marked with a sigle asterisk
-achieved a higher correlation fewer than 1000 times, resulting in a p-value
-less than $10^{-3}$.  Those marked with two asterisks achieved a higher
+million times, and counted the number of times each of the competing methods
+achieved a higher correlation than the BN method.  Those marked with a sigle
+asterisk achieved a higher correlation fewer than 1000 times, resulting in a
+p-value less than $10^{-3}$.  Those marked with two asterisks achieved a higher
 correlation in none of the bootstrap samples, indicating a p-value less than
@@
 \end{table}
+
+\subsection{Robustness}
+
+Training our model on more reads leads to more accurate estimation of bias, but
+an increasingly long training time. For example, in our tests, fitting our model to
+100,000 reads from the Mortazavi data, training time was approximately 45
+minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
+the training to 25,000 reads leads to a model that is nearly as accurate while
+requiring less than 4 minutes to train. A full discussion of this trade-off is in
+Supplementary Section 5.
+
+The quality of the solution depends also on two other parameters: the standard
+deviation at which background sequences are sampled, and the weight applied to
+the penalty term of the BIC, yet it is not particularly sensitive to their
+values. (The median $R^2$ goodness-of-fit statistic used in section 3.2 varied
+by less that 25\% as these parameters were varied over a range of $10^4$. See
+Supplementary Section 2.) In all our evaluation, no special tuning of the
+parameters was performed, suggesting it can be used effectively across data sets
+without any intervention.
+
+Additionally, experimental and theoritical analysis suggest that the procedure
+is very resistant to inclusion of extranious parameters. In Supplementary
+Section 10 we prove an upper bound on the probability of our model predicting
+any bias, if the experiment is in fact unbiased, showing that there very little
+risk in the applying the method to an unbiased data set. In particular, if more
+than 10,000 reads are used in the training procedure, the probability that any
+adjustment at all will is made is less than 0.0004.
 
@@
 parameters are estimated with less accuracy than in models requiring fewer
-parameters. Yet, we trained the 7mer model on a minimum of 600,000 reads (with
-the Bullard data set), and up to 23 million reads (with the Wetterbom data set),
-numbers which our theoretical results, detailed in Supplementary Section 9,
-suggest should lead to accurate estimates. A perhaps more significant factor is
-that this method does not capture bias outside of the initial heptamer, though
-many data sets clearly are affected by bias in other positions.  However, if we
-were to incease the size of the $k$-mer being considered, exponentially more
-reads would be required for an accurate estimate.
+parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
+(with the Mortazavi data set), and up to 23 million reads (with the Wetterbom
+data set), numbers which our theoretical results, detailed in Supplementary
+Section 9, suggest should lead to accurate estimates. A perhaps more significant
+factor is that this method does not capture bias outside of the initial
+heptamer, though many data sets clearly are affected by bias in other positions.
+Thus to improve the performance, it seems necessary to incease the size of the
+$k$-mers being considered. However, exponentially more reads would be required
+for an accurate estimate since the accuracy of the model, as quantified by its
+KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is the
+number of parameters that must be estimated (See Supplementary Section 9).
 
@@
 
-Training our model on more reads leads to more accurate estimation of bias, but
-an increasingly long training time. For example, in our tests, fitting our model to
-100,000 reads from the Mortazavi data, training time was approximately 45
-minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
-the training to 25,000 reads leads to a model that is nearly as accurate while
-requiring less than 4 minutes to train. A full discussion of this trade-off is in
-Supplementary Section 5.
-
 Because we do not require annotations, ChIP-Seq, and other high-throughput
@@
 evaluated; however, our method is still able to effectively correct for the bias
-that was observed (see, Supplementary Section 6).  Protocol differences, as we
+that was observed (see Supplementary Section 6).  Protocol differences, as we
 have seen, can result in significant differences in observed nucleotide
@@
 generated on the Illumina platform, it appears to be effected only by relatively
-few positions adjacent to the read start (see, Supplementary Section 7).
+few positions adjacent to the read start (see Supplementary Section 7).
 
@@
 concern is the consistency of the bias between samples. In the data we examined,
-the bias appears to be largely, but not entirely consistent (see, Supplementary
+the bias appears to be largely, but not entirely consistent (see Supplementary
 Section 8).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced

### 2011-09-30 seqbias-supplement.tex
final draft

@@
 
-\title{SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data}
+\title{SUPPLEMENTARY:\\A new approach to bias correction in high-throughput sequencing data}
 \author{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}
@@
 
-\dcjtitle{\sc{(SUPPLEMENTARY)}}{\sc{Correcting for Bias in High-Throughput
-Sequencing Data}}{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}
+\dcjtitle{\sc{(SUPPLEMENTARY)}}
+{\sc{A new approach to bias correction in high-throughput sequencing data}}
+{Daniel C. Jones, Walter L. Ruzzo, Xinxia Peng, Michael G. Katze}
 
@@
 \end{center}
-\caption{Sensitivity of the parameter control he standard deviation at which
-background sequences are sampled.}
+\caption{Sensitivity of the parameter controling the standard deviation at which
+background sequences are sampled, as evaluated with McFadden's $R^2$
+goodness-of-fit statistic.}
 \label{fig:poissampstd}
@@
 
-Figure \ref{fig:poissampstd} show the standard deviation used to draw background
-samples varied from 1 to 10,000. Apparent from this plot is that, while the
-optimal choice lies somewhere between 10 and 100, the model performs
-competitively for any reasonable value.
+Figure \ref{fig:poissampstd} shows the median McFaddens $R^2$ goodness-of-fit
+statastic across 1000 test exons (as described in Section 3.2 of the main paper)
+versus the standard deviation used to draw background samples varied from 1 to
+10,000. Apparent from this plot is that, while the optimal choice lies somewhere
+between 10 and 100, the model performs competitively for any reasonable value.
 
@@
 \end{center}
-\caption{Sensitivity over adjustments to the complexity penalty.}
+\caption{Sensitivity over adjustments to the complexity penalty. The $Y$ axis is
+as in Figure \ref{fig:poissampstd}.}
 \label{fig:poiscomppen}
@@
 Figure \ref{fig:caokl}, we plot directly the positional KL divergence, computed
-in tho same manner described in the Results section of the paper.  We see that
+in the same manner described in the Results section of the paper.  We see that
 our method is effective at reducing the bias in this ChIP-Seq data.
@@
 generate this data, is not without bias, yet there is significant divergence at
-only a small number of positions surrounding the read start. When our method is
-trained on this data we obtain the suitably sparse model pictured in Figure
+only a small number of positions near the read start, including the position
+before it. When our method is trained on this data we obtain the suitably sparse
+model pictured in Figure
 \ref{fig:frtgraph}.
@@
 differential expression tests. This would bias the discovery of differentially
-expressed genes, but not the test itself.
+expressed genes, but not the test itself.  Additionally, as sequencing bias
+appears to be at least partly related to sequence context, alternative splicing
+potentially will change the bias at positions somewhat distant from the splice
+site itself, complicating any analysis of differential isoform usage.
 
-\section{KL Divergence}
+\section{KL Divergence and Parameter Estimation}
 
 \subsection{Overview}
+
+In this section and the next, we present some theoretical analysis quantifying
+the accuracy of parameter estimation and the likelihood of falsely learning a
+biased model from unbiased data.
 
@@
 a measure of the dissimilarity between the distributions: if $p_{i} \approx
-q_{i}$ everywhere, their log rations will be near zero and $H$ will be small; as
-$q_{i}$ and $p_{i}$ diverge, their log rotions will deviate from zero and $H$
+q_{i}$ everywhere, their log ratios will be near zero and $H$ will be small; as
+$q_{i}$ and $p_{i}$ diverge, their log ratios will deviate from zero and $H$
 will increase.
@@
 $P$ (arbitrarily called the ``null'') or from $Q$ (the ``alternative''), how
-large should $m$ be to cenfidently choose between these cases? A natural
+large should $m$ be to confidently choose between these cases? A natural
 approach is to use a likelihood ratio test: letting $Y_i$ be the number of times
@@
 sequence of observations under model $P$ is $$\prod_{i} p_{i}^{Y_{i}}$$
-and similarily for $Q$. So, the logarithm of the likelihood ratios is
+and similarly for $Q$. So, the logarithm of the likelihood ratios is
 $$
@@
 If the sample is drawn from the alternative distribution $Q$, then the
-expectation of $frac{Y_{i}}{m}$ is $q_i$, so the expected value of the LLR is
+expectation of $\frac{Y_{i}}{m}$ is $q_i$, so the expected value of the LLR is
 $$ m \sum_{i} q_{i} \ln \frac{q_{i}}{p_{i}} = m H(Q || P) $$
@@
 So, assuming the null hypothesis is false, in order for it to be rejected with
-$1000:1$ odds, one should choose $m$ to be inversely proportional to $H(Q ||
+say, $1000:1$ odds, one should choose $m$ to be inversely proportional to $H(Q ||
 P)$:
@@
 As a concrete example, all of the RNA-Seq data sets examined in this paper show
-a KL divergence between the 1$^{\text{st}}$ positon of reads versus the
+a KL divergence between the 1$^{\text{st}}$ position of reads versus the
 transcriptomic background nucleotide distribution of $0.05$ or greater. Thus,
@@
 sampled across the transcriptome by examining only a few hundred randomly
-selected reads (i.e., $\ln 1000 / 0.05 \approx 140$).
+selected reads (i.e., $\ln 1000 / 0.05 \approx 140$), a surprisingly small
+number, given the supposedly ``random sampling'' accomplished by RNA-Seq.
 
@@
 in the sample space (e.g. $r = 4^{k}$ in the case of $k$-mers). Given a random
-ample $X_1, X_2, \dots, X_r$ of size $n = \sum_{i} X_i$ from $P$, it is well
+sample $X_1, X_2, \dots, X_r$ of size $n = \sum_{i} X_i$ from $P$, it is well
 known that the maximum likelihood estimators for the parameters are $q_i =
 \frac{X_i}{n} \approx p_i$. How good an estimate for $P$ is this distribution
-$Q$?
-
+$Q$? The estimators are unbiased:
 $$E[q_i] = E\left[\frac{X_i}{n}\right] = \frac{E[X_i]}{n} = \frac{np_i}{n} =
 p_i$$
-And the standard deviation of each estimate is proportional to $1/\sqrt{n}$. So
+and the standard deviation of each estimate is proportional to $1/\sqrt{n}$, so
 these estimates are increasingly accurate as the sample size increases. A more
@@
 = \sum_{i = 1}^{r} q_i \ln \left(1 + \frac{q_i - p_i}{p_i} \right) $$
-Using the first two terms of he Taylor series for $\ln (1 + x)$, this is
+Using the first two terms of the Taylor series for $\ln (1 + x)$, this is
 \begin{align*}
@@
 \end{align*}
-Since $\sum_{i = 1}^{r} q_i = \sum_{i = 1}^{r} pi = 1$,
+Since $\sum_{i = 1}^{r} q_i = \sum_{i = 1}^{r} p_i = 1$,
 $\sum_{i = 1}^{r} p_i \frac{q_i - p_i}{p_i} = 0$, so
@@
 \begin{align*}
-H(Q||P) &\approx \frac{1}{2n} \sum_{i=1}^{r} \frac{n q_i - n p_i)^2}{n p_i} \\
+H(Q||P) &\approx \frac{1}{2n} \sum_{i=1}^{r} \frac{(n q_i - n p_i)^2}{n p_i} \\
         &= \frac{1}{2n} \sum_{i=1}^{r} \frac{(X_i - E[X_i])^2}{E[X_i]}
@@
 
-Stochastic simulations of this proves when $P$ is uniform show that the above
-formula is a very good fit when $n > r$.
+Stochastic simulations of this, shown in Figure \ref{fig:entropy}, proves when
+$P$ is uniform show that the above formula is a very good fit when $n > r$.
 
@@
 $H(Q||P)$, averaged over 100 samples of size $n$, where $P$ is uniform and $Q$
-is the MLE estimator for $P$ basen on $n$ random samples drawn from $r$ bins.
+is the MLE estimator for $P$ based on $n$ random samples drawn from $r$ bins.
 The asterisks and the straight line interpolating them are the theoretical
@@
 
-The training procedure begins by considering the the set of models formed by
+The training procedure begins by considering the set of models formed by
 including each single nucleotide position from within the training sequences.
-For each single nucleotide model, the Bayesian infromation criterion is
-evaluated. If in any of these model the BIC score increases over the empty
+For each single nucleotide model, the Bayesian information criterion is
+evaluated. If in any of these models the BIC score increases over the empty
 model, a non-empty model will be trained, otherwise the training procedure will
@@
 
-Now, objective function evaluated is the Bayesian information criterion (BIC)
+Now, the objective function evaluated is the Bayesian information criterion (BIC)
 applied to the conditional log-likelihood (CLL).
@@
 empty model being trained is bounded below by the probability that the KL
-divergence of both sampled is bounded above $\frac{3}{2n} \log n$:
+divergence of both samples is bounded above by $\frac{3}{2n} \log n$:
 $$ \Pr(D_{X} < \frac{3}{2n} \log n) \cdot 
@@
 
-In training we consider mulitple nucleotides. If $h$ nucleotides are considered,
+In training we consider multiple nucleotides. If $h$ nucleotides are considered,
 the probability of a non-empty model being trained is less than
@@
 this probability for various values of $n$. This is shown in Figure
-\ref{fig:nepr}. From this figure we can see, for example that training with more
-than 10,000 reads results in an incorrect non-empty model with probability less
-than 0.0004.
+\ref{fig:nepr}. From this figure we can see, for example, that training with
+more than 10,000 reads results in an incorrect non-empty model with probability
+less than 0.0004.
 
-In the low end, the upper bound on non-empty model probability is much higher.
+For small $n$, the upper bound on non-empty model probability is much higher.
 In our tests, this upper bound is not tight for a small number of reads. We
-trained the model on 100 sets of 100 simulated reads, drawn randomly at uniform
+trained the model on 100 sets of 100 simulated reads, drawn uniformly at random
 from exonic sequences and in only 3 trials was a non-empty model trained, far
-less than the upper bound prediction of 23 out of 100. This test was repeated
-using sets of 10,000 reads, and in no trial was a non-empty model trained.
+less than the upper bound expectation of 23 out of 100. This test was repeated
+using sets of 10,000 reads, and in no trial was a non-empty model trained.  From
+these results, and the results in Supplementary Section 5, we recommend training
+our method with at least 10,000 reads.
 
-From these results, and the results in Supplementary Section 5, we recommend
-training our method with at least 10,000 reads.
+Here we considered only the case of an empty model versus a non-empty model. A
+similar analysis can be conducted to show that if an extraneous position were
+added to to the initial empty model, it would be very improbable that more
+extraneous positions are edged were added. So that if, despite the very low
+probability, a non-empty model were trained from unbiased data, it would be
+extremely sparse and have little effect.
 
@@
 model when there is no real bias in the experiment when the model is trained
-with various number of reads $n$. We fixed the number of nucleotides being
-considered as $h = 41$, as this what was used for the evaluation in Section 3.}
+with various numbers of reads. We fixed the number of nucleotides being
+considered as $h = 41$, as was used for the evaluation in Section 3.}
 \label{fig:nepr}

### 2011-09-30 seqbias-paper.tex
tweaks to reduce page number to 8

@@
 This bias, though observed primarily in the $5'$ end of a read, is not resolved by
-trimming the reads prior to mapping \citep{Hansen2010} (see also, Supplementary
+trimming the reads prior to mapping \citep{Hansen2010} (see also Supplementary
 Section 1), suggesting it is not a result of erroneous base calling, and that a
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{overview.eps}}
+\centerline{\includegraphics[width=0.37\textwidth]{overview.eps}}
 \caption{An overview of the approach taken: (a) foreground sequences are sampled
@@
 found, in which no single alteration to the network will increase the score.
-
 Given the network structure, the parameters are estimated directly from the
@@
 The run time of the training procedure is further reduced in practice by
-imposing the following two restrictions on the structure of the network,
-\begin{enumerate}
-\item The in-degree (i.e., number of parents) of any node must be less than some
-number $p_{\text{max}}$.
-\item $|j - i| \le d_{\text{max}}$ for all edges $(i,j)$ and some number $d_{\text{max}}$.
-\end{enumerate}
-The second of these encodes the assumption that distant nucleotides are
-effectively independent. We choose $p_{\text{max}} = 4$ and $d_{\text{max}} =
-10$, as reasonable default values.
+imposing the following two restrictions on the structure of the network, First,
+the in-degree (i.e., number of parents) of any node must be less than some
+number $p_{\text{max}}$. Secondly, for all edges $(i,j)$, $|j - i| \le
+d_{\text{max}}$ for some number $d_{\text{max}}$.  This encodes the assumption
+that distant nucleotides are effectively independent. We choose $p_{\text{max}}
+= 4$ and $d_{\text{max}} = 10$, as reasonable default values.
 
@@
 parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
-(with the Mortazavi data set), and up to 23 million reads (with the Wetterbom
-data set), numbers which our theoretical results, detailed in Supplementary
-Section 9, suggest should lead to accurate estimates. A perhaps more significant
-factor is that this method does not capture bias outside of the initial
-heptamer, though many data sets clearly are affected by bias in other positions.
-Thus to improve the performance, it seems necessary to incease the size of the
-$k$-mers being considered. However, exponentially more reads would be required
-for an accurate estimate since the accuracy of the model, as quantified by its
-KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is the
-number of parameters that must be estimated (See Supplementary Section 9).
+(see Table \ref{tab:counts}), a number which our theoretical results, detailed
+in Supplementary Section 9, suggest should lead to accurate estimates. A perhaps
+more significant factor is that this method does not capture bias outside of the
+initial heptamer, though many data sets clearly are affected by bias in other
+positions.  Thus to improve the performance, it seems necessary to incease the
+size of the $k$-mers being considered. Yet, exponentially more reads would be
+required for an accurate estimate since the accuracy of the model, as quantified
+by its KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is
+the number of parameters that must be estimated (See Supplementary Section 9).
 
@@
 have seen, can result in significant differences in observed nucleotide
-frequencies, so we may not safely assert that bias in ChIP-Seq data is always
+frequencies, so we can not safely assert that bias in ChIP-Seq data is always
 low.  Given the weak assumptions made by our model, our estimation of bias could
@@
 
+%\vspace{-8em}
+
 \bibliographystyle{natbib}

### 2011-11-22 seqbias-paper.tex
Added p_max/d_max plots, fixed ensemble 60/62 mixup, removed the use of the word 'classification'.

@@
 
-Towards that end, we propose training a Bayesian network classifier trained on
-examples of foreground and background sequences. Though our goal is not
-classification, by using the machinery of classification we can avoid a model
-that is over-parametrized. Parameters that are not informative in
-discriminating between foreground and background are not included in the model.
-The Bayesian network can then be used to evaluate sequence probability, and thus
-bias, at any genomic position. Figure \ref{fig:overview} gives a schematic overview
-of the proposed model.
+Towards that end, we propose training a Bayesian network using examples of
+foreground and background sequences. By training the model discriminatively and
+penalizing model complexity, we can avoid a model that is over-parametrized,
+excluding parameters that insufficiently informitive in discriminating between
+foreground and background. The Bayesian network can then be used to evaluate
+sequence probability, and thus bias, at any genomic position. Figure
+\ref{fig:overview} gives a schematic overview of the proposed model.
 
@@
 counted reads overlapping each gene, defining the gene by the union of every
-transcript in release 62 of the Ensembl gene annotations, excluding UTRs, and
-100 additional bases from both ends of the gene. Counts were then normalized by
-dividing by the length of these genes. We then removed any genes with a read
-count less than 10, or that did not correspond to a unique TaqMan probe.
+transcript in release 60 of the Ensembl gene annotations. Counts were then
+normalized by dividing by the length of these genes. We then removed any genes
+with a read count less than 10, or that did not correspond to a unique TaqMan
+probe.
 
@@
 Method & Correlation & \\ \midrule
-Unadjusted &  0.6665 & $\ast\ast$ \\
-7mer &  0.6677 & $\ast\ast$ \\
-GLM & 0.6854 & $\ast\ast$ \\
-MART & 0.6921 & $\ast$ \\
-BN & \textbf{0.7016}
+Unadjusted &  0.6650 & $\ast\ast$ \\
+7mer &  0.6680 & $\ast\ast$ \\
+GLM & 0.6874 & $\ast\ast$ \\
+MART & 0.6998 & $\ast$ \\
+BN & \textbf{0.7086}
 \end{tabular*}}
@@
 using the BN method over the other methods using a simple boostrap procedure. A
-bootstrap sample is formed by sampling, with replacement, 592 genes from the
+bootstrap sample is formed by sampling, with replacement, 648 genes from the
 original set of the same size. The correlation is then computed for this set,

### 2011-11-22 seqbias-supplement.tex
Added p_max/d_max plots, fixed ensemble 60/62 mixup, removed the use of the word 'classification'.

@@
 entirely adequate solution.
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{fig/pois-p_max.pdf}
+\end{center}
+\caption{Sensitivity over values of the $p_{\text{max}}$ parameter,
+controlling the maximum in-degree of any node in the directed graph. The $Y$ axis is
+as in Figure \ref{fig:poissampstd}.}
+\label{fig:pois_p_max}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{fig/pois-d_max.pdf}
+\end{center}
+\caption{Sensitivity over values of the $d_{\text{max}}$ parameter, controlling
+the maximum distance any edge in the directed graph may traverse.  The $Y$ axis
+is as in Figure \ref{fig:poissampstd}.}
+\label{fig:pois_d_max}
+\end{figure}
+
+Figures \ref{fig:pois_p_max} and \ref{fig:pois_d_max} suggest that most
+dependencies are between adjacent nucleotides, as increasing the maximum
+in-degree ($p_{\text{max}}$), or maximum edge distance ($d_{\text{max}}$) has
+little effect for values of at least 1.
 

### 2011-11-22 results/plots/plot-d_max.R

@@ -1,0 +1,50 @@
+
+
+d <- '../d_max'
+
+ns <- c('0', '1', '2', '3', '4', '5', '6', '7', '8', '9',
+        '10', '11', '12', '13', '14', '15', '16', '17', '18', '19', '20')
+
+
+xs <- NULL
+for (n in ns) {
+    fn <- sprintf('%s/pois.%s.tab', d, n)
+    ys <- read.table(fn, header = T)
+    ys$n <- n
+    xs <- rbind(xs, ys)
+}
+
+
+ys <- read.table(paste('../20110610/', '/pois/mortazavi.tab', sep = ''), header = T)
+ys <- subset(ys, method == 'Unadjusted')[,c('sample', 'gene', 'L')]
+xs <- merge(xs, ys, by = c('sample', 'gene'))
+names(xs) <- c('sample', 'gene', 'method', 'mu', 'L', 'n', 'L.base')
+
+
+r2 <- function(L, L.base) { 1 - L / L.base }
+xs$r2 <- r2(xs$L, xs$L.base)
+med <- tapply(xs$r2, INDEX = list(xs$sample, xs$method, xs$n), FUN = median)
+
+library(reshape)
+
+med <- melt(med)
+names(med) <- c('sample', 'method', 'n', 'med')
+
+
+
+library(Cairo)
+library(ggplot2)
+source('theme_dcjstd.R')
+
+CairoSVG('pois-d_max.svg', width = 12, height = 6)
+p <- qplot(data = med, x = n, y = med, geom = 'line')
+p <- p + geom_point()
+p <- p + ylim(c(0, max(med$med)))
+p <- p + scale_x_continuous(name = 'd_max')
+p <- p + scale_y_continuous(name = 'Median Goodness of Fit', limits = c(0, max(med$med)))
+p <- p + theme_dcjstd()
+print(p)
+
+dev.off()
+
+

### 2011-11-22 results/plots/plot-p_max.R

@@ -1,0 +1,53 @@
+
+d <- '../p_max'
+
+ns <- c('1', '2', '3', '4', '5', '6', '7', '8', '9',
+        '10', '11', '12', '13')
+
+
+xs <- NULL
+for (n in ns) {
+    fn <- sprintf('%s/pois.%s.tab', d, n)
+    ys <- read.table(fn, header = T)
+    ys$n <- n
+    xs <- rbind(xs, ys)
+}
+
+
+ys <- read.table(paste('../20110610/', '/pois/mortazavi.tab', sep = ''), header = T)
+ys <- subset(ys, method == 'Unadjusted')[,c('sample', 'gene', 'L')]
+xs <- merge(xs, ys, by = c('sample', 'gene'))
+names(xs) <- c('sample', 'gene', 'method', 'mu', 'L', 'n', 'L.base')
+
+
+r2 <- function(L, L.base) { 1 - L / L.base }
+xs$r2 <- r2(xs$L, xs$L.base)
+med <- tapply(xs$r2, INDEX = list(xs$sample, xs$method, xs$n), FUN = median)
+
+library(reshape)
+
+med <- melt(med)
+names(med) <- c('sample', 'method', 'n', 'med')
+
+
+
+library(Cairo)
+library(ggplot2)
+source('theme_dcjstd.R')
+
+CairoSVG('pois-p_max.svg', width = 12, height = 6)
+p <- qplot(data = med, x = n - 1, y = med, geom = 'line')
+p <- p + geom_point()
+#p <- p + ylim(c(0, max(med$med)))
+p <- p + scale_x_continuous(name = 'p_max')
+p <- p + scale_y_continuous(name = 'Median Goodness of Fit', limits = c(0, max(med$med)))
+p <- p + theme_dcjstd()
+print(p)
+
+dev.off()
+
+
+
+
+
+

### 2011-11-23 seqbias-paper.tex
Rewordings.

@@
 %\title{Correcting for bias in high-throughput sequencing data}
-\title{A new approach to bias correction in high-throughput sequencing data}
+\title{A new approach to bias correction in RNA-Seq}
 \author[Jones \textit{et~al}]
@@
 \section{Motivation:}
-Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
-often conflated by protocol-specific sequence bias. The exact sources of the
-bias are unknown, but may be influenced by PCR amplification, or differing
-primer affinities and mixtures, for example.  The result is decreased accuracy
-in many applications, such as de novo gene annotation, isoform quantification in
-RNA-Seq, and peak calling in ChIP-Seq.
+Quantification of sequence abundance in RNA-Seq experiments is often conflated
+by protocol-specific sequence bias. The exact sources of the bias are unknown,
+but may be influenced by PCR amplification, or differing primer affinities and
+mixtures, for example. The result is decreased accuracy in many applications,
+such as de novo gene annotation and gene or isoform quantification.
 
@@
 
-\begin{table}
-\processtable{Numbers of reads used to train each method.\label{tab:counts}}
-{
-\begin{tabular}{lcccc}\toprule
-Experiment & Total & 7mer & GLM/MART & BN \\\midrule
-Wetterbom & $3.9 \times 10^7$ & $2.3 \times 10^7$ & $5.9 \times
-10^6$ & $1 \times 10^5$ \\
-Katze & $4.8 \times 10^7$ & $2.0 \times 10^7$ & $3.1 \times 10^6$ & $1 \times
-10^5$ \\
-Bullard & $1.2 \times 10^7$ & $3.1 \times 10^6$ & $1.8 \times 10^5$ & $1 \times
-10^5$\\
-Mortazavi & $3.9 \times 10^6$ & $1.9 \times 10^6$ & $3.8 \times 10^5$ & $1
-\times 10^5$ \\
-Trapnell & $3.4 \times 10^7$ & $1.6 \times 10^7$ & $5.9 \times 10^6$ & $1 \times
-10^5$
-\end{tabular}
-}{Note that in the Trapnell data set (the only paired-end set considered) we
-count only the first mate of each read.}
-\end{table}
-
 Testing was performed by cross-validation. Each method was trained on data taken
@@
 unique to each. The total number of reads used to train each method is listed in
-Table \ref{tab:counts}, and below we describe the procedure used for
+Supplementary Section 3, and below we describe the procedure used for
 each.
@@
 
-To avoid the risk of a small number of reads with many duplicates dominating the
-measure, we do not count duplicate reads when computing the KL divergence. Since we
-explicitly chose exons with high coverage, not counting duplicates may not
-fully capture the effects of the bias correction methods. Thus, for each exon,
-we compute the number of duplicates of each read and count only reads in the
-upper half. Under the assumption of uniform sampling, the set of reads falling
-in the upper half should not depend on sequence, and we should expect the KL
-divergence to be low.
+When computing the KL divergence, there is a risk of the measure being dominated
+by a small number of reads with many duplicates. Yet, given the high coverage of
+the exons being tested, if duplicate reads are excluded, it may not capture the
+full effect of bias correction. To account for these opposing concerns we adopt
+the following method: all reads contained within the exon being tested are
+ranked by the number of duplicates. We then exclude reads that are ranked in the
+lower half, and count each read ranked in the upper half only once, ignoring
+duplicates.
 
-We compute the divergence by reweighting the read counts using the predicted
-bias coefficient, choosing the upper half in each exon, discarding duplicate
-reads, and then tallying frequencies of overlapping $k$-mers. The $k$-mer
-distribution obtained is then compared to a background distribution obtained by
-redistributing reads uniformly at random within their exons.
+Under the assumption of uniform sampling, the set of reads ranked in the upper
+half should not depend on sequence, and we should expect the KL divergence to be
+low.  We compute the divergence by reweighting the read counts using the
+predicted bias coefficient before ranking the reads, choosing those reads
+ranked in the upper half of each exon, ignoring duplicate reads, and then
+tallying frequencies of overlapping $k$-mers. The $k$-mer distribution obtained
+is then compared to a background distribution obtained by redistributing reads
+uniformly at random within their exons.
 
@@
 analysis are plotted in Figure \ref{fig:kl}, for $k = 1$ and $k = 4$. The
-remaining cases are plotted in Supplementary Section 3.
+remaining cases are plotted in Supplementary Section 4.
 
@@
 RNA-Seq data shows a pattern of bias similar to that seen in the other samples
-sequenced on an Illumina platform (see Supplementary Section 5). This evaluation
+sequenced on an Illumina platform (see Supplementary Section 6). This evaluation
 does not rely on an assumption that qRT-PCR necissarily more accurate than
@@
 requiring less than 4 minutes to train. A full discussion of this trade-off is in
-Supplementary Section 5.
+Supplementary Section 6.
 
@@
 is very resistant to inclusion of extranious parameters. In Supplementary
-Section 10 we prove an upper bound on the probability of our model predicting
+Section 11 we prove an upper bound on the probability of our model predicting
 any bias, if the experiment is in fact unbiased, showing that there very little
@@
 parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
-(see Table \ref{tab:counts}), a number which our theoretical results, detailed
-in Supplementary Section 9, suggest should lead to accurate estimates. A perhaps
+(see Supplementary Section 3), a number which our theoretical results, detailed
+in Supplementary Section 10, suggest should lead to accurate estimates. A perhaps
 more significant factor is that this method does not capture bias outside of the
@@
 by its KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is
-the number of parameters that must be estimated (See Supplementary Section 9).
+the number of parameters that must be estimated (See Supplementary Section 10).
 
@@
 evaluated; however, our method is still able to effectively correct for the bias
-that was observed (see Supplementary Section 6).  Protocol differences, as we
+that was observed (see Supplementary Section 7).  Protocol differences, as we
 have seen, can result in significant differences in observed nucleotide
@@
 generated on the Illumina platform, it appears to be effected only by relatively
-few positions adjacent to the read start (see Supplementary Section 7).
+few positions adjacent to the read start (see Supplementary Section 8).
 
@@
 the bias appears to be largely, but not entirely consistent (see Supplementary
-Section 8).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
+Section 9).  Similarly, in Figure \ref{fig:freqs}, the three data sets sequenced
 on the Illumina platform display similar patterns of nonuniformity, yet differ

### 2011-11-23 seqbias-supplement.tex
Rewordings.

@@
 little effect for values of at least 1.
+
+\section{Total numbers of reads used to train each method}
+
+\begin{table}[H]
+\begin{center}
+\begin{tabular}{lcccc}
+Experiment & Total & 7mer & GLM/MART & BN \\\hline
+Wetterbom & $3.9 \times 10^7$ & $2.3 \times 10^7$ & $5.9 \times
+10^6$ & $1 \times 10^5$ \\
+Katze & $4.8 \times 10^7$ & $2.0 \times 10^7$ & $3.1 \times 10^6$ & $1 \times
+10^5$ \\
+Bullard & $1.2 \times 10^7$ & $3.1 \times 10^6$ & $1.8 \times 10^5$ & $1 \times
+10^5$\\
+Mortazavi & $3.9 \times 10^6$ & $1.9 \times 10^6$ & $3.8 \times 10^5$ & $1
+\times 10^5$ \\
+Trapnell & $3.4 \times 10^7$ & $1.6 \times 10^7$ & $5.9 \times 10^6$ & $1 \times
+10^5$
+\end{tabular}
+\caption{Note that in the Trapnell data set (the only paired-end set considered) we
+count only the first mate of each read.}
+\label{tab:counts}
+\end{center}
+\end{table}
 

### 2011-11-23 response2/response.tex
Some minor rewarding, added the start of a response letter and the round 2 reviews.

\documentclass{article}
\usepackage{amsmath}
\usepackage{dcj}

\begin{document}

\subsection*{Response to Reviewer 1}

\begin{quote}
In the supplement the authors show an upper bound on the probability that their
method will learn a non-empty model when in fact there is no bias in the data. I
think this is an important feature of the method and should be highlighted
better in the paper.
\end{quote}

We appreciate this suggestion, and made note of the theoretical results in the
revised abstract of the paper.

\begin{quote}
My only concern is with the sampling described on pg3, column 2, lines 32-47.
In RNA-seq data the distribution of read count can span 5-6 orders of
magnitudes, the top expressed gene accounting for up to 10\% of the data
(especially tissue specific). If you sample at random you are selecting more
frequently from the top, because of this heavy skew in the distribution. You
discuss this issue when measuring KL-divergence in section 3.1, but it does not
appear to affect the training. My main worry is that this will bias the
adjustment to mostly highly expressed genes. How do you take care of this?

One approach would be to simply discard duplicates. In that case how do you
count the sequences, i.e. if two reads map to the same position with one with
sequencing errors, one without do you count them as two sequences or do you
treat them as one and use the sequence of the genome?
\end{quote}

In fact, we do ignore duplicate reads when training the method. This is stated
in the first paragraph of Section 2.2.1: ``we take sequences surrounding
(extending 20 nt to either side, by default) the start positions of a randomly
sampled set of $n/2$ aligned reads, ignoring duplicate reads.'' The reviewer is
correct that this is an important detail, as the method might otherwise be
overfit to a few highly expressed genes, and in this revision we have expanded
this paragraph to clarify it.

\begin{quote}
pg 3, column 1, lines 42-50 \\
Although BN's are mostly used for classification purposes they are just a
compact representation of a discrete probability distribution, the discussion
of classification is somewhat confusing.
\end{quote}

It was our intention to point out that the training algorithm used is typical of
classification applications, but this point was perhaps needlessly confusing.
We have rewritten the paragraph in question (the fifth paragraph in Section 2.1,
beginning with ``Towards that end...'') to avoid this.

\begin{quote}
supplement pg 7, figure 11. \\
Since you have the 4 replicates from the Trapnell et al. data and you show the
nucleotide frequencies for each replicate, why not run your method independently
on each replicate and compare the modified results.
\end{quote}

We agree that a looking for an increase in correlation of gene abundance
estimates between replicates after bias correction would be an interesting
analysis. Unfortunately, the four RNA-Seq data sets from Trapnell, et al, were
taken from four different biological states and compared without technical or
biological replicates, so comparing gene abundance estimates would be conflated
with genuine changes in gene expression.

While we wish to draw attention to the issue of batch effects in RNA-Seq, a larger
survey of variability of technical bias is outside the intended scope of the
paper.

\subsection*{Response to Reviewer 2}

\begin{quote}
counts are biased and mainly affected by surrounding nucleotide frequencies.
For RNA-Seq data, this assumption has been justified by previous work (Hansen et
al. 2010 and Li et al. 2010). But it is not readily extendable to other NGS
data, such as ChIP-Seq. ChIP-Seq has an enrichment step for DNA fragments with
very specific characteristics (e.g. transcription factor binding sites) , it is
imaginable that the surrounding nucleotide frequencies around a ChIP-Seq peak
may be quite different from other genomic regions, and most sequencing reads
from real binding sites may have very similar surrounding nucleotide
frequencies. To accurately measure read counts for a ChIP-Seq peak, it matters
more to distinguish reads from background and from real binding sites and
correct for protocol-related biases such as PCR over-amplification. The authors
may need more effort to establish the fact that biases related to surrounding
nucleotide frequencies exist in multiple ChIP-Seq data sets. Moreover, even for
RNA-Seq data, only considering biases in base-level read counts does not seem to
be enough for accurate quantitation of transcript level as shown in (Zheng et
al. 2011 BMC Bioinformatics). Features other than the 20 upstream and downstream
nucleotides surrounding read starts may also cause biases in gene/transcript
quantitation. The authors may discuss a little more on their assumptions.
\end{quote}

TODO

\begin{quote}
Page 2, Figure 1. Wetterbom data does not seem to have much bias in surrounding
nucleotide frequencies. What is the difference between this and Katze data set
that makes such difference? Why do the different Illumina data sets have similar
bias trends whereas ABI data sets do not?
\end{quote}

TODO

\begin{quote}
Page 4, line 31, left column. How are the default values of pmax and dmax
determined? Could you provide a table with model complexity, model performance
and running time with different pmax and dmax choices?
\end{quote}

As suggested by Figure 3 in the paper, most edges added to the model are between
adjacent nucleotides, and the in-degree of most positions is small.  We have now
included additional data (Figures 4 and 5 in the Supplement) showing that the
performance of the model on the Mortazavi data set is not greatly effected by
these parameters as long as they are both non-zero.

\begin{quote}
Page 5, line28, right column. This sentence is confusing. Are the duplicated
reads counted or not? Is “upper half” referring to the position in the exon or
reads with number of duplicates in the upper 50 percentile?
\end{quote}

We have rewritten the 3rd paragraph in section 3.1 to provide a more lucid
description of the procedure.

\begin{quote}
Page 6, figure 4. Is the unadjusted k=1 case equivalent to the KL panel in
Figure 1? If so, why is Bullard data not consistent KL=0.6 in figure 4 and 0.3
in Figure 1. Other data sets seem to be consistent.
\end{quote}

Regrettably, Figure 1 figure used the natural logarithm to compute KL
divergence, while Figure 4 used the base 2 logarithm, causing the scale to
somewhat shifted between the two. This has been corrected in the revision.

\begin{quote}
Page 7, table 4. Is the presented correlation $r$ or $r^2$?
\end{quote}

The correlation measurement used was the Pearson's $r$, which he have clarified.

\begin{quote}
Different versions of annotation were referred. Page 4 line 49: Ensembl 60 is
used in training models, page 7 line 32, Ensembl 62 is used in comparing to
RT-PCR data. Should make these consistent.
\end{quote}

We have repeated tho RT-PCR analysis with Ensembl 60, to remain consistent with
the other results.

\subsection*{Response to Reviewer 3}

\begin{quote}
The result sections, however, did not include any comparison directly
on bias reduction. Rather, results are focused on improvement of model
fitting and statistical significance of correlation -- which are both
important but not the ultimate goal.

In terms of bias reduction: first, it is necessary to see the amount of bias
reduced using the proposed approach versus the other methods in
comparison.  The increase in uniformaity is a promising sign, but does
not guarantee that the gene expression measure is improved.  With the
bias correction implemented, what will be the recommended measure for
a gene or exon level expression? Sum of $\Pr[m_i|s_i]$? Whether the bias of
this value is reduced is a much more convincing argument than the
increased uniformity itself. There are  biological information that
would allow this type of comparison. We have seen only one example in
Fig 2, but no direct quantification in bias reduction.
\end{quote}

TODO

\begin{quote}
Second, as the authors point out in introduction, another problem of
these biases is that they cannot be assumed to be ``identical between
replicates". How much of the non-identical bias is removed with the
proposed approach?
\end{quote}

TODO

\begin{quote}
Another issue as important as bias reduction is variance reduction. The
bias-variance trade off is unavoidable. By estimating the bias and trying to
adjust for it, more variance is introduced to the system. Is the amount of bias
reduced worth the increased variance, for downstream analysis?
\end{quote}

We agree that it possible to reduce bias and yet introduce so much variance that
that experiments are actually rendered less reliable. In the paper, we attempted
to capture this tradeoff most directly with Figure 5, which demonstrates that,
while uniformity of read coverage is increased ($R^2 > 0$) in the vast majority
of exons after applying our method, inevitably it is decreased ($R^2 < 0$) in
some.

TODO: expand upon this thought.

\begin{quote}
A minor point: If all examples are from RNAseq experiment, how should
one modify the method when it is applied to other sequencing data?
Would one do the same for ChIP-Seq? IF the current method is meant for
RNAseq only, the title should reflect this instead of general ``high
throughput sequencing data".
\end{quote}

No modification at all is needed to apply the method to ChIP-Seq data. This is
true also of the Hansen method, but not the case with methods of Li or Roberts.
While we believe application to ChIP-Seq is a promising research direction and
one that our method is particularly suited for, besides some preliminary
analysis the results presented here are directed at RNA-Seq, and we have
reworded the title and abstract to reflect this.

\section*{Conclusion}

TODO

\end{document}

### 2011-11-23 seqbias-paper.tex
Some minor rewarding, added the start of a response letter and the round 2 reviews.

@@
 simple graphical model. Our model does not rely on existing gene annotations,
-making it applicable to any quantitative high throughput sequencing data. We
-evaluate our method on several data sets, and by multiple criteria,
-demonstrating that it effectively decreases bias and increases uniformity.
+and model selection is performed automatically making it applicable with few
+assumptions. We evaluate our method on several data sets, and by multiple
+criteria, demonstrating that it effectively decreases bias and increases
+uniformity. Additionally, we provide theoretical and empirical results showing
+that the method is unlikely to have any effect on unbiased data, suggesting it
+can be applied with little risk of spurious adjustment.
 
@@
 penalizing model complexity, we can avoid a model that is over-parametrized,
-excluding parameters that insufficiently informitive in discriminating between
-foreground and background. The Bayesian network can then be used to evaluate
-sequence probability, and thus bias, at any genomic position. Figure
+excluding parameters that are insufficiently informitive in discriminating
+between foreground and background. The Bayesian network can then be used to
+evaluate sequence probability, and thus bias, at any genomic position. Figure
 \ref{fig:overview} gives a schematic overview of the proposed model.
@@
 sequences, we take sequences surrounding (extending 20 nt to either side, by
-default) the start positions of a randomly sampled set of $n/2$ aligned reads,
-ignoring duplicate reads.
+default) the start positions of a randomly sampled set of $n/2$ aligned reads.
+To avoid the risk of the method being overfit to reads deriving from a few
+highly expressed genes, we ignore duplicate reads, which we define as tow reads
+mapping to the same location in the genome. The nucleotide sequence is taken
+from the genome, rather then the reads themselves, allowing us to include
+positions outside of the read.
 
@@
 eight chromosomes.  After adjusting read counts according to the predicted
-sequence bias, we computed the Pearson correlation between log read counts and
-log TaqMan expression values, which are averaged across three replicates. These
-correlations are listed in Table \ref{tab:pcr}. Our method shows a statistically
-significant increase in correlation compared to the other methods.
+sequence bias, we computed the Pearson correlation coefficient $r$ between log
+read counts and log TaqMan expression values, which are averaged across three
+replicates. These correlations are listed in Table \ref{tab:pcr}. Our method
+shows a statistically significant increase in correlation compared to the other
+methods.
 
@@
 \processtable{
-    Pearson correlation between log-adjusted read counts and log-adjusted TaqMan
-    values.\label{tab:pcr}}{
+    The Pearson correlation coefficient $r$ between log-adjusted read counts and
+    log-adjusted TaqMan values.\label{tab:pcr}}{
     \centerline{

### 2011-11-27 response2/response.tex
Most of the response letter.

@@
 
-\subsection*{Response to Reviewer 1}
+TODO: Introduction
+
+\section*{Summary}
+
+TODO
+
+\section*{Response to Reviewer 1}
 
@@
 overfit to a few highly expressed genes, and in this revision we have expanded
-this paragraph to clarify it.
+this paragraph to clarify this decision.
 
@@
 taken from four different biological states and compared without technical or
-biological replicates, so comparing gene abundance estimates would be conflated
-with genuine changes in gene expression.
+biological replicates, so a comparison of gene abundance estimates would be
+conflated with genuine changes in gene expression.
 
@@
 
-\subsection*{Response to Reviewer 2}
+\section*{Response to Reviewer 2}
 
 \begin{quote}
+General concerns. This work has an underlying assumption that base-level read
 counts are biased and mainly affected by surrounding nucleotide frequencies.
@@
 data, such as ChIP-Seq. ChIP-Seq has an enrichment step for DNA fragments with
-very specific characteristics (e.g. transcription factor binding sites) , it is
+very specific characteristics (e.g. transcription factor binding sites), it is
 imaginable that the surrounding nucleotide frequencies around a ChIP-Seq peak
@@
 may need more effort to establish the fact that biases related to surrounding
-nucleotide frequencies exist in multiple ChIP-Seq data sets. Moreover, even for
-RNA-Seq data, only considering biases in base-level read counts does not seem to
-be enough for accurate quantitation of transcript level as shown in (Zheng et
-al. 2011 BMC Bioinformatics). Features other than the 20 upstream and downstream
-nucleotides surrounding read starts may also cause biases in gene/transcript
-quantitation. The authors may discuss a little more on their assumptions.
+nucleotide frequencies exist in multiple ChIP-Seq data sets.
 \end{quote}
 
-TODO
+The reviewer is correct that we have by no means established that applying our
+method to ChIP-Seq leads to more accurate results, and have instead concentrated
+on making this case for RNA-Seq. We have included some very preliminary ChIP-Seq
+results, only to suggest that investigating the effects of sequence bias in
+ChIP-Seq is a promising future research direction. To avoid the impression that
+we are making an overly-broad claim, we have reworded the title and abstract to
+reflect that our results relate most readily to RNA-Seq.
+
+\begin{quote}
+Moreover, even for RNA-Seq data, only considering biases in base-level read
+counts does not seem to be enough for accurate quantitation of transcript level
+as shown in (Zheng et al. 2011 BMC Bioinformatics). Features other than the 20
+upstream and downstream nucleotides surrounding read starts may also cause
+biases in gene/transcript quantitation. The authors may discuss a little more on
+their assumptions.
+\end{quote}
+
+The Zheng paper make the assertion that ``the bias patterns across genes were
+not adequately addressed by the base-level bias correction methods, and GAM
+[Generalized Additive Model] correction is necessary to further remove these
+biases.'' This follows from their observation that Hansen's heptamer reweighting
+method did not improve correlation bewteen RNA-Seq and RT-PCR. Indeed, our
+results are consistent, showing only a slight improvment.
+
+However, the authors apparently did not evaluate this correlation after applying
+a more sophisticated model such as the GLM or MART model propsed by Li. As we
+demonstrate in Section 3.3, applying a more flexible model results in a large
+increase in correlation, suggesting such an approach should not be hastily
+dismissed.
 
@@
 
-TODO
+While both these data sets were generated using the ABI platform, there are
+significant differences in the protocol used. Most significantly, the Wetterbom
+data was generated using poly-A primers to enrich for mature, polyadenylated
+mRNA. In the Katze data, ribosomal RNA was depleted, but polyadenylated
+transcripts were not selected for, in an effort to sequence more of the
+non-polyadenylated transcriptome. The Kate data set is unique in this aspect:
+all three of the Illumina data sets were generated with mRNA-Seq protocols as
+well.  We have revised Table 1 to make note of this fact.
 
@@
 performance of the model on the Mortazavi data set is not greatly effected by
-these parameters as long as they are both non-zero.
+these parameters as long as they are both non-zero. It should not bed assumed
+that this necessarily the case with every data set, and so these parameters
+default to somewhat arbitrary larger numbers that balance training time with
+flexibility.
 
@@
 Page 5, line28, right column. This sentence is confusing. Are the duplicated
-reads counted or not? Is “upper half” referring to the position in the exon or
+reads counted or not? Is ``upper half'' referring to the position in the exon or
 reads with number of duplicates in the upper 50 percentile?
@@
 
-Regrettably, Figure 1 figure used the natural logarithm to compute KL
-divergence, while Figure 4 used the base 2 logarithm, causing the scale to
-somewhat shifted between the two. This has been corrected in the revision.
+Regrettably, in Figure 1, te natural logarithm was used to compute KL
+divergence, while in Figure 4, the base 2 logarithm, causing the scale to
+somewhat shifted between the two. This error has been corrected in the revision.
 
@@
 We have repeated tho RT-PCR analysis with Ensembl 60, to remain consistent with
-the other results.
+the Poisson regression and KL divergence results.
 
-\subsection*{Response to Reviewer 3}
+\section*{Response to Reviewer 3}
 
@@
 
-TODO
+We use bias to refer to the fact that the estimator in question (RNA-Seq
+based measurement of gene abundance) does not, in expectation, agree with the
+value we wish to estimate (the true gene abundance). To measure the bias
+directly, we would then need to know the true number of copies of each
+transcript we wish to quantify and compare RNA-Seq based estimates before and
+after correcting for sequence bias. Because exact quantification of transcript
+copy number is not feasible at a large scale with any existing technology, we
+instead used the largest publically available RT-PCR data set that we are aware
+of and demonstrated a substantial increase in the agreement between RNA-Seq and
+RT-PCR after correcting for bias in the former (see Section 3.3).
+
+In the example shown in Figure 2, we provided a qualitative illustration of the
+large scale quantitative analysis of Section 3.1 and Section 3.2, demonstrating
+a reduction in the symptoms of sequence bias across thousands of exons in five
+data sets, and in comparison to three previously proposed methods. In sum, we
+believe this represents the most complete and rigorous characterization of
+sequence bias in RNA-Seq yet presented and is a compelling convergence of
+evidence that the accuracy RNA-Seq is improved by correcting for sequence
+bias, and that the method we propose is particularily suited to do so.
 
@@
 
-TODO
+We have seen that sequence bias is not always consistent, even in experiments
+using similar protocols (see Figure 1, for example), and believe that batch
+effects in RNA-Seq are a serious concern. So far, we have concentrated only on
+rigorously demonstrating a absolute reduction in sequence bias, and have not
+undertaken any systematic comparison of the consistency of this bias between
+replicates. Such comparison would be meaningful, but outside of the intended
+scope of the paper.
 
@@
 to capture this tradeoff most directly with Figure 5, which demonstrates that,
-while uniformity of read coverage is increased ($R^2 > 0$) in the vast majority
+while uniformity of read coverage is increased ($R^2 > 0$) in the large majority
 of exons after applying our method, inevitably it is decreased ($R^2 < 0$) in
-some.
-
-TODO: expand upon this thought.
+some. This suggests that the complexity penalty we use in model selection works
+as desired, and as as its ansymptotic properties predict, controlling variance by
+preventing very dense models from being trained.
 

### 2011-11-27 seqbias-paper.tex
Most of the response letter.

@@
 \begin{tabular}{lllcc}\toprule
- & & & Read  \\
-Experiment & Species & Platform & length \\\midrule
-\citet{Wetterbom2010} & Chimpanzee & ABI & 33  \\
-Katze (unpublished) & Macaque & ABI & 50  & \\
-\citet{Bullard2010} & Human & Illumina & 35 \\
-\citet{Mortazavi2008} & Mouse & Illumina & 33 \\
-\citet{Trapnell2010} & Mouse & Illumina & 75 \\\botrule
+                      &            &          &          & Read  \\
+Experiment            & Species    & Platform & Protocol & Length \\\midrule
+\citet{Wetterbom2010} & Chimp.     & ABI      & mRNA     & 33  \\
+Katze (unpublished)   & Macaque    & ABI      & WT       & 50 \\
+\citet{Bullard2010}   & Human      & Illumina & mRNA     & 35 \\
+\citet{Mortazavi2008} & Mouse      & Illumina & mRNA     & 33 \\
+\citet{Trapnell2010}  & Mouse      & Illumina & mRNA     & 75 \\\botrule
 \end{tabular}
-}{}
+}{
+The protocol column lists whether a poly-A priming step to select for
+polyadenylated transcripts was used (``mRNA''), or depletion of ribosomal RNA
+with no step to select for polyadenylated transcripts (``WT'').
+}
 \end{table}
@@
 by less that 25\% as these parameters were varied over a range of $10^4$. See
-Supplementary Section 2.) In all our evaluation, no special tuning of the
+Supplementary Section 2.) The same is true of the $p_{\text{max}}$ and
+$d_{\text{max}}$ parameters, used restrict the in-degree and edge distance of
+the model, respectively, in order to control training time. Our tests show that
+these parameters need only be greater than zero for an adequate model to trained
+for the Mortazavi data.  In all our evaluation, no special tuning of the
 parameters was performed, suggesting it can be used effectively across data sets
@@
 
-Additionally, experimental and theoritical analysis suggest that the procedure
-is very resistant to inclusion of extranious parameters. In Supplementary
+Additionally, experimental and theoretical analysis suggest that the procedure
+is very resistant to inclusion of extraneous parameters. In Supplementary
 Section 11 we prove an upper bound on the probability of our model predicting

### 2011-11-27 response2/response.tex
More work on the response.

@@
 
-TODO: Introduction
+\section*{Summary of Changes}
 
-\section*{Summary}
+\begin{itemize}
+\item In response concerns raised by reviewers 2 and 3, we we have reworded the
+title and abstract to make clear that our results are most directly applicable
+to RNA-Seq, and we are not yet able to make strong assertions regarding
+ChIP-Seq.
 
-TODO
+\item We have taken the first reviewer's suggestion and have highlighted the
+strong theoretical results our method provides in the abstract.
+
+\item In response to the first reviewers primary concern, we have clarified the
+first paragraph of Section 2.1.1 to make clear that our method does ignore
+duplicate reads when training.
+
+\item We have removed our use of the term ``classification'' in the fifth
+paragraph of Section 2.1, as it seemed to cause some confusion.
+
+\item We have clarified protocol differences between the Wetterbom and Katze
+data sets in Table 1, in response to the second reviewer's question.
+
+\item We have included new data in Supplementary Section 2 showing that the
+results presented are not typically sensitive to the values of $p_{\text{max}}$
+and $d_{\text{max}}$
+
+\item The description of the procedure we used to compute KL divergence in the
+3rd paragraph of Section 3.1 has been rewritten to be more clear.
+
+\item We have corrected an inconsistency in the scale of figures 1 and 4.
+
+\item We have clarified that the values in Table 3 are Pearson correlation
+coefficient $r$, and not $r^2$.
+
+\item We have repeated the RT-PCR analysis of 3.3 using version 60 of the
+Ensembl gene annotations, rather than version 62, to remain consistent with the
+Poisson regression and KL divergence analysis.
+
+\item We have added the proper citation of what we believe to be first published
+derivation of the results of Supplementary Section 10.
+
+\time We have cited and made note of in Section 4 a related paper by Zheng, et
+al, brought to our attention by reviewer 2.
+
+\item Table 2 was moved to Supplementary Section 3 to preserve space.
+
+\end{itemize}
 
@@
 
-We have rewritten the 3rd paragraph in section 3.1 to provide a more lucid
-description of the procedure.
+We have rewritten the 3rd paragraph in Section 3.1 to provide a more lucid
+description of the procedureve.
 
@@
 
-\section*{Conclusion}
-
-TODO
-
 \end{document}

### 2012-01-03 response2/response.tex
Final paper, let's hope.

@@
 \usepackage{amsmath}
-\usepackage{dcj}
+%\usepackage{dcj}
 
 \begin{document}
+
+TODO: brief introduction
+
+We have included below a summary of the changes made in this revision in
+addition to specific responses to the points raised by the reviewers.
 
@@
 
-\item In response to the first reviewers primary concern, we have clarified the
-first paragraph of Section 2.1.1 to make clear that our method does ignore
-duplicate reads when training.
+\item The first paragraph of Section 2.1.1 has been reworded to make clear that
+our method does ignore duplicate reads when training.
 
@@
 
-\item We have added the proper citation of what we believe to be first published
-derivation of the results of Supplementary Section 10.
+\item In the second paragraph of Section 4, we have added the proper citation of
+what we believe to be first published derivation of the results of Supplementary
+Section 10.
 
-\time We have cited and made note of in Section 4 a related paper by Zheng, et
-al, brought to our attention by reviewer 2.
+%\time We have cited and made note of in Section 4 a related paper by Zheng, et
+%al, brought to our attention by reviewer 2.
 
@@
 
-In fact, we do ignore duplicate reads when training the method. This is stated
-in the first paragraph of Section 2.2.1: ``we take sequences surrounding
+In fact, we do ignore duplicate reads when training the method. This was stated
+in the first paragraph of Section 2.2.1 (``we take sequences surrounding
 (extending 20 nt to either side, by default) the start positions of a randomly
-sampled set of $n/2$ aligned reads, ignoring duplicate reads.'' The reviewer is
+sampled set of $n/2$ aligned reads, ignoring duplicate reads''), but the reviewer is
 correct that this is an important detail, as the method might otherwise be
-overfit to a few highly expressed genes, and in this revision we have expanded
+overfit to a few highly expressed genes. In this revision we have expanded
 this paragraph to clarify this decision.
@@
 biological replicates, so a comparison of gene abundance estimates would be
-conflated with genuine changes in gene expression.
-
-While we wish to draw attention to the issue of batch effects in RNA-Seq, a larger
-survey of variability of technical bias is outside the intended scope of the
-paper.
+conflated with genuine changes in gene expression.  While we wish to draw
+attention to the potential of batch effects in RNA-Seq, a larger survey of
+variability of technical bias is outside the intended scope of the paper.
 
@@
 
-The Zheng paper make the assertion that ``the bias patterns across genes were
+The Zheng paper makes the assertion that ``the bias patterns across genes were
 not adequately addressed by the base-level bias correction methods, and GAM
@@
 However, the authors apparently did not evaluate this correlation after applying
-a more sophisticated model such as the GLM or MART model propsed by Li. As we
-demonstrate in Section 3.3, applying a more flexible model results in a large
+a more sophisticated model such as the GLM or MART models as propsed by Li. As
+we demonstrate in Section 3.3, applying a more flexible model results in a large
 increase in correlation, suggesting such an approach should not be hastily
@@
 performance of the model on the Mortazavi data set is not greatly effected by
-these parameters as long as they are both non-zero. It should not bed assumed
+these parameters as long as they are both non-zero. It should not be assumed
 that this necessarily the case with every data set, and so these parameters
 default to somewhat arbitrary larger numbers that balance training time with
-flexibility.
+model flexibility.
 
@@
 We have rewritten the 3rd paragraph in Section 3.1 to provide a more lucid
-description of the procedureve.
+description of the procedure.
 
@@
 effects in RNA-Seq are a serious concern. So far, we have concentrated only on
-rigorously demonstrating a absolute reduction in sequence bias, and have not
+rigorously demonstrating an absolute reduction in sequence bias, and have not
 undertaken any systematic comparison of the consistency of this bias between
@@
 of exons after applying our method, inevitably it is decreased ($R^2 < 0$) in
-some. This suggests that the complexity penalty we use in model selection works
-as desired, and as as its ansymptotic properties predict, controlling variance by
-preventing very dense models from being trained.
+some. This suggests that the (BIC) complexity penalty we use during model
+selection works as desired, and as as its ansymptotic properties predict,
+controlling variance by preventing very dense models from being trained.
 

### 2012-01-03 seqbias-paper.tex
Final paper, let's hope.

@@
 \pubyear{2011}
-
-\usepackage{comment}
 
@@
 mixtures, for example. The result is decreased accuracy in many applications,
-such as de novo gene annotation and gene or isoform quantification.
+such as de novo gene annotation and transcript quantification.
 
@@
 normalization to accounting for cross-hybridization, RNA-Seq presents a new set
-of challenges. A significant example is the effect that transcript length has in
-biasing quantification, as first explored by \citet{Oshlack2009}, and
-subsequently addressed by \cite{Young2010, Gao2011}.
+of challenges. As first noted by \citet{Dohm2008}, a particular challenge is the
+often complex and protocol-specific influence of nucleotide sequence on
+quantification.
 
-The underlying nucleotide sequence also influences quantification in a manner
-that is complex and protocol-specific. In an ideal experiment, the number of
-reads mapping to a particular position in the genome is a function of RNA
-abundance and should not be additionally dependent on the sequence at that
-position. As noted first by \citet{Dohm2008}, this is not the case.  As
-illustration, Figure \ref{fig:freqs} plots this nonuniformity in nucleotide
-frequencies on five data sets (see Table \ref{tab:datasets}), each using a
-different protocol.
+In an ideal experiment, the number of RNA-Seq reads mapping to a particular
+position in the genome is a function of RNA abundance and should not be
+additionally dependent on the sequence at that position. Yet, this is not the
+case. As illustration, Figure \ref{fig:freqs} plots this nonuniformity in
+nucleotide frequencies on five data sets (see Table \ref{tab:datasets}), each
+using a different protocol.
 
@@
 assumption given that the causes of the bias are not well understood.
-Additionally, in many tests of differential expression---similar to the effects
-of transcript length---a higher read count will result in higher statistical
-confidence. It follows that the sensitivity of such a test will also be biased
-by sequence, affecting downstream analysis such as gene ontology enrichment
-tests.
+Additionally, in many tests of differential expression higher read count will
+result in higher statistical confidence. It follows that the sensitivity of such
+a test will also be biased by sequence, affecting downstream analysis such as
+gene ontology enrichment tests.
 
@@
 That is, if $N$ reads are generated, and $m_i$ is the event that a generated
-read maps to position $i$, then
-$$ E[x_i] = N \Pr[m_i] $$
+read maps to position $i$, then $ E[x_i] = N \Pr[m_i] $.
 
@@
 number of reads sampled from position $i$ is independent of $s_i$, i.e., if
-
 $$ E[ x_i | s_i ] = N \Pr[ m_i | s_i ] = N \Pr[ m_i ] = E[ x_i ] $$
@@
 penalizing model complexity, we can avoid a model that is over-parametrized,
-excluding parameters that are insufficiently informitive in discriminating
+excluding parameters that are insufficiently informative in discriminating
 between foreground and background. The Bayesian network can then be used to
@@
 uniform sampling model, after correcting for bias, with 1.0 indicating a perfect
-fit, and the score of 0.38 here indicating a significant increase.
-}
+fit, and the score of 0.38 here indicating a significant increase.}
 \label{fig:overview}
@@
 
-However, we have so far ignored one complication: the RNA abundance that we wish
+We have so far ignored one complication: the RNA abundance that we wish
 to estimate is not itself independent of the nucleotide sequence. Notably,
@@
 within the same transcript, but we prefer to avoid dependence on existing
-gene annotations as such a model would be limited to RNA-Seq analysis in which
-trustworthy and reasonably complete gene annotations are available.
+gene annotations.
 
@@
 \subsubsection{Sampling}
-
 The model is trained on $n$ sequences, one half labeled as foreground, the other
@@
 To avoid the risk of the method being overfit to reads deriving from a few
-highly expressed genes, we ignore duplicate reads, which we define as tow reads
+highly expressed genes, we ignore duplicate reads, which we define as two reads
 mapping to the same location in the genome. The nucleotide sequence is taken
-from the genome, rather then the reads themselves, allowing us to include
+from the genome, rather than the reads themselves, allowing us to include
 positions outside of the read.
@@
 \subsubsection{Training}
-
 To determine the structure and parameters of the Bayesian network, we use a
@@
 \begin{align*}
-\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) 
+\ell &= \sum_{i=1}^{n} \log \Pr[ x_i | s_i ]
 =
-\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
-\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
+\sum_{i=1}^{n} \log \frac{ \Pr[s_i | x_i] \Pr[ x_i ] }{
+\sum_{x \in \{0,1\}} \Pr[ s_i | x ] \Pr[x] } \\
 \end{align*}
-where $\Pr(x)$ is flat (i.e.  $\Pr( x = 0 ) = \Pr( x = 1 ) =
+where $\Pr[x]$ is flat (i.e.  $\Pr[ x = 0 ] = \Pr[ x = 1 ] =
 0.5$) since we sample foreground and background positions equally.
@@
 At each step of the optimization procedure, every possible edge or position
-addition, removal, or edge reversal that produces a valid network (i.e., does not
-introduce a cycle) is evaluated, and the alteration that increases the score
+addition, removal, or edge reversal that produces a valid, acyclic network
+is evaluated, and the alteration that increases the score
 $\ell'$ the most is kept.  This process is repeated until a local maximum is
@@
 number $p_{\text{max}}$. Secondly, for all edges $(i,j)$, $|j - i| \le
-d_{\text{max}}$ for some number $d_{\text{max}}$.  This encodes the assumption
-that distant nucleotides are effectively independent. We choose $p_{\text{max}}
-= 4$ and $d_{\text{max}} = 10$, as reasonable default values.
-
-Along with $p_{\text{max}}$ and $d_{\text{max}}$, the time required to train the
-model is controlled by the number of reads used (i.e., $n$) and the number of
-positions surrounding the read start to consider. Increasing any of these
-parameters results in a  solution that is potentially better, at
-the expense of increased training time.
+d_{\text{max}}$ for some number $d_{\text{max}}$. This latter rule encodes the
+assumption that distant nucleotides are effectively independent. We choose
+$p_{\text{max}} = 4$ and $d_{\text{max}} = 10$, as reasonable default values
+(see Supplementary Section 2).
 
@@
 Because we cannot observe directly the underlying RNA-abundance, our evaluation
-strategy relies on testing three assumptions we make of an ideal, unbiased experiment.
+strategy relies on testing three assumptions we make of an ideal, unbiased
+RNA-Seq experiment.
 \begin{enumerate}
@@
 
-Testing was performed by cross-validation. Each method was trained on data taken
+Each method was trained on data taken
 from chromosomes 1--8 of the genome from which the reads were mapped (including
@@
 which is the mean per-position log-likehood under the null model). This measure
-can also take on values less than 0, indicating that the model has worse fit
-than the null model (i.e., when adjusting read counts according the bias
-coefficients leads to less uniform read coverage).
+has the added advantage that it can take on values less than 0, indicating that
+the model has worse fit than the null model (i.e., when adjusting read counts
+according the bias coefficients leads to less uniform read coverage).
 
@@
 sequenced on an Illumina platform (see Supplementary Section 6). This evaluation
-does not rely on an assumption that qRT-PCR necissarily more accurate than
+does not rely on an assumption that qRT-PCR is necessarily more accurate than
 RNA-Seq based quantification, only that qRT-PCR is not biased in the same way as
@@
 effective as the other models, even when data generated using random hexamer
-priming was used. A possible explaination is, given the large number of
+priming was used. A possible explanation is, given the large number of
 parameters needed to estimate heptamer frequencies ($4^{7} = 16,383$), these
@@
 parameters. Yet, we trained the 7mer model on a minimum of 1.9 million reads
-(see Supplementary Section 3), a number which our theoretical results, detailed
-in Supplementary Section 10, suggest should lead to accurate estimates. A perhaps
-more significant factor is that this method does not capture bias outside of the
-initial heptamer, though many data sets clearly are affected by bias in other
-positions.  Thus to improve the performance, it seems necessary to incease the
-size of the $k$-mers being considered. Yet, exponentially more reads would be
-required for an accurate estimate since the accuracy of the model, as quantified
-by its KL divergence from the true distribution, is $(r-1)/2n$, where $r=4^k$ is
-the number of parameters that must be estimated (See Supplementary Section 10).
+(see Supplementary Section 3), a number which theoretical results,
+following from work by \citet{Birch1964} and included in Supplementary Section
+10 for completeness, suggest should lead to accurate estimates
+
+A perhaps more significant factor is that this method does not capture bias
+outside of the initial heptamer, though many data sets clearly are affected by
+bias in other positions.  Thus to improve the performance, it seems necessary to
+increase the size of the $k$-mers being considered. However, exponentially more reads
+would be required for an accurate estimate since the accuracy of the model, as
+quantified by its KL divergence from the true distribution, is $(r-1)/2n$, where
+$r=4^k$ is the number of parameters that must be estimated (See Supplementary
+Section 10).
 
@@
 
+During the review of this manuscript, another method addressing sequence bias in
+RNA-Seq was published by \citet{Zheng2011}. Rather than fitting a model of the
+specific base-level sequence bias surrounding read start, they propose making
+adjustments according to summary statistics at the gene level. Such an approach
+is disadvantaged in its inability to model the very specific pattern of sequence
+bias we have observed. Yet, such an approach is efficient, and though we have
+not yet evaluated it, claimed to be effective.
+
 Because we do not require annotations, ChIP-Seq, and other high-throughput
@@
 generated on the Illumina platform, it appears to be effected only by relatively
-few positions adjacent to the read start (see Supplementary Section 8).
+few positions adjacent to the read start (see Supplementary Section 8). Other
+protocol improvements might further reduce sequence bias. Notably, promising work
+by \citet{Jayaprakash2011} proposes a pooled adapter strategy to deal with this
+issue in small RNA sequencing experiments.
 

### 2012-01-03 seqbias-supplement.tex
Final paper, let's hope.

@@
 
-\section{Total numbers of reads used to train each method}
+\section{Total Numbers of Reads Used to Train Each Method}
 
@@
 the accuracy of parameter estimation and the likelihood of falsely learning a
-biased model from unbiased data.
+biased model from unbiased data. These results follow mostly from work
+originally done by Birch \cite{Birch1964}, but are presented here for
+completeness, and in a manner more directly addressing the the question at hand.
 
@@
 
-As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $\chi^2$
+As discussed in Supplementary Section 10.2, $2n D$ is approximated by a $\chi^2$
 distribution with 3 degrees of freedom for large $n$, and so we can estimate
@@
 using sets of 10,000 reads, and in no trial was a non-empty model trained.  From
-these results, and the results in Supplementary Section 5, we recommend training
+these results, and the results in Supplementary Section 6, we recommend training
 our method with at least 10,000 reads.

### 2012-01-03 response2/response.tex
Revise the response letter.

@@
 
-TODO: brief introduction
+The paper submitted here is a revision of one originally titled ``A new approach
+to bias correction in high-throughput sequencing data''.  We found the reviewers
+comments to be very constructive overall, and are grateful for their time and
+attention. Some small errors were brought to our attention, as well as several
+points of confusion which we have taken care to address in this revision.
 
-We have included below a summary of the changes made in this revision in
-addition to specific responses to the points raised by the reviewers.
+Belew is a summary of the changes made, in addition to specific responses to the
+points reaised by the reviewers.
 
@@
 \item We have removed our use of the term ``classification'' in the fifth
-paragraph of Section 2.1, as it seemed to cause some confusion.
+paragraph of Section 2.1, as it seemed only to cause confusion.
 
@@
 
-%\time We have cited and made note of in Section 4 a related paper by Zheng, et
-%al, brought to our attention by reviewer 2.
+\item In Section 4, we have made note of recent paper by Jayaprakash, et al, as
+well as the work by Zheng, et al, that was brought to our attention by the
+second reviewer.
 
@@
 
-It was our intention to point out that the training algorithm used is typical of
-classification applications, but this point was perhaps needlessly confusing.
-We have rewritten the paragraph in question (the fifth paragraph in Section 2.1,
+It was our intention to point out that the training algorithm used (i.e.,
+maximizing the conditional log-likelihood) is typical of classification
+applications, but this point was perhaps needlessly confusing. We have
+rewritten the paragraph in question (the fifth paragraph in Section 2.1,
 beginning with ``Towards that end...'') to avoid this.
@@
 method did not improve correlation bewteen RNA-Seq and RT-PCR. Indeed, our
-results are consistent, showing only a slight improvment.
+results are mostly consistent, showing only a slight improvment.
 
 However, the authors apparently did not evaluate this correlation after applying
-a more sophisticated model such as the GLM or MART models as propsed by Li. As
+a more sophisticated model such as the GLM or MART models as proposed by Li. As
 we demonstrate in Section 3.3, applying a more flexible model results in a large
@@
 transcripts were not selected for, in an effort to sequence more of the
-non-polyadenylated transcriptome. The Kate data set is unique in this aspect:
+non-polyadenylated transcriptome. The Katze data set is unique in this aspect:
 all three of the Illumina data sets were generated with mRNA-Seq protocols as
@@
 these parameters as long as they are both non-zero. It should not be assumed
-that this necessarily the case with every data set, and so these parameters
+that this is necessarily the case with every data set, and so these parameters
 default to somewhat arbitrary larger numbers that balance training time with
@@
 believe this represents the most complete and rigorous characterization of
-sequence bias in RNA-Seq yet presented and is a compelling convergence of
+sequence bias in RNA-Seq yet presented. It is a compelling convergence of
 evidence that the accuracy RNA-Seq is improved by correcting for sequence
-bias, and that the method we propose is particularily suited to do so.
+bias, and that the method we propose is particularly suited to do so.
 
@@
 some. This suggests that the (BIC) complexity penalty we use during model
-selection works as desired, and as as its ansymptotic properties predict,
+selection works as desired, and as its ansymptotic properties predict,
 controlling variance by preventing very dense models from being trained.
@@
 one that our method is particularly suited for, besides some preliminary
-analysis the results presented here are directed at RNA-Seq, and we have
+analysis the results presented here are directed at RNA-Seq. We have
 reworded the title and abstract to reflect this.

### 2012-01-05 response2/response.tex
Fix some typos discovered by Xinxia.

@@
 
-The paper submitted here is a revision of one originally titled ``A new approach
-to bias correction in high-throughput sequencing data''.  We found the reviewers
-comments to be very constructive overall, and are grateful for their time and
-attention. Some small errors were brought to our attention, as well as several
-points of confusion which we have taken care to address in this revision.
+\vspace{8em}
+
+The paper presented here is revised version of the one submitted on Sept. 30
+2011. In this last round of review, we found the reviewers comments to be
+very constructive overall, and are grateful for their time and attention. Some
+small errors were discovered, as well as several points of confusion which we
+have taken care to address. While no major changes to the method or results were
+required, we have, in a number of ways, improved the clarity of our
+presentation, and believe that is now fit for publication in Bioinformatics.
 
 Belew is a summary of the changes made, in addition to specific responses to the
-points reaised by the reviewers.
+points raised by the reviewers.
 
@@
 
-\item We have repeated the RT-PCR analysis of 3.3 using version 60 of the
+\item We have repeated the RT-PCR analysis of Section 3.3 using version 60 of the
 Ensembl gene annotations, rather than version 62, to remain consistent with the
@@
 
-Regrettably, in Figure 1, te natural logarithm was used to compute KL
+Regrettably, in Figure 1, the natural logarithm was used to compute KL
 divergence, while in Figure 4, the base 2 logarithm, causing the scale to
@@
 sequence bias in RNA-Seq yet presented. It is a compelling convergence of
-evidence that the accuracy RNA-Seq is improved by correcting for sequence
+evidence that the accuracy of RNA-Seq is improved by correcting for sequence
 bias, and that the method we propose is particularly suited to do so.

### 2018-04-12 response2/response.tex
Add readme. (and commit some weirdly uncommited changes from yore.)

@@
 \begin{document}
-
-\vspace{8em}
 
@@
 correct that this is an important detail, as the method might otherwise be
-overfit to a few highly expressed genes. In this revision we have expanded
-this paragraph to clarify this decision.
+overfit to a few highly expressed genes.
+
+In this revision we have expanded this paragraph to clarify this decision, and
+it now reads as follows:
+``The model is trained on $n$ sequences, one half labeled as foreground, the other
+background, sampled from the reference genome. To obtain the foreground
+sequences, we take sequences surrounding (extending 20 nt to either side, by
+default) the start positions of a randomly sampled set of $n/2$ aligned reads.
+To avoid the risk of the method being overfit to reads deriving from a few
+highly expressed genes, we ignore duplicate reads, which we define as two reads
+mapping to the same location in the genome. The nucleotide sequence is taken
+from the genome, rather than the reads themselves, allowing us to include
+positions outside of the read.''
 
@@
 maximizing the conditional log-likelihood) is typical of classification
-applications, but this point was perhaps needlessly confusing. We have
-rewritten the paragraph in question (the fifth paragraph in Section 2.1,
-beginning with ``Towards that end...'') to avoid this.
+applications, but this point was perhaps needlessly confusing.
+
+We have rewritten the paragraph in question to avoid this (the fifth paragraph
+in Section 2.1), and it now reads as: ``Towards that end, we propose training a
+Bayesian network using examples of foreground and background sequences. By
+training the model discriminatively and penalizing model complexity, we can
+avoid a model that is over-parametrized, excluding parameters that are
+insufficiently informative in discriminating between foreground and background.
+The Bayesian network can then be used to evaluate sequence probability, and thus
+bias, at any genomic position. Figure 2 gives a schematic
+overview of the proposed model.''
 
@@
 description of the procedure.
+
+It reads now as:
+``When computing the KL divergence, there is a risk of the measure being dominated
+by a small number of reads with many duplicates. Yet, given the high coverage of
+the exons being tested, if duplicate reads are excluded, it may not capture the
+full effect of bias correction. To account for these opposing concerns we adopt
+the following method: all reads contained within the exon being tested are
+ranked by the number of duplicates. We then exclude reads that are ranked in the
+lower half, and count each read ranked in the upper half only once, ignoring
+duplicates.''
 
